---
title: "28. Hypothesis tests and paired data"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Two surfaces that had no R or Python binding until this release: the twelve Numerics
`HypothesisTests` statics, reached directly through `hypothesis_test()` or, for a flood-frequency
record, through the `DataFrame` facades `analysis_data_hypothesis_test()`/`analysis_data_statistics()`;
and the RMC.BestFit "Paired Data" subsystem -- an x-y curve container with interpolation, area, three
curve-simplification algorithms, and an uncertain twin whose y-coordinate is a whole distribution
per point. This page works through both, on a real flood record and a real reservoir curve.

## What you'll learn

- What four hypothesis tests answer about a flood-frequency record -- normality, independence,
  and homogeneity -- and how to read a p-value from each.
- The `DataFrame` path through `analysis_data()`: the same tests over a censored-data-aware
  container, the `use_log10` switch, and the two-sample `index` split.
- `correlation()`'s matrix form over more than two series at once.
- A reservoir stage-storage curve interpolated in linear and log space, and its trapezoidal area.
- How the three curve-simplification algorithms differ on the same curve, including one that
  drops the curve's last point without warning.
- Sampling an uncertain curve at its mean versus its median.

## Setup

```{r}
library(corehydror)
```

## A flood-frequency record

Sixty-nine years of annual peak discharge (m^3^/s) for the Harricana River, Quebec (Bobee & Ashkar
1991, *Multivariate Flood Frequency Analysis*, Table 1.2):

```{r}
harricana <- c(
  122, 244, 214, 173, 229, 156, 212, 263, 146, 183, 161, 205, 135, 331, 225,
  174, 98.8, 149, 238, 262, 132, 235, 216, 240, 230, 192, 195, 172, 173, 172,
  153, 142, 317, 161, 201, 204, 194, 164, 183, 161, 167, 179, 185, 117, 192,
  337, 125, 166, 99.1, 202, 230, 158, 262, 154, 164, 182, 164, 183, 171, 250,
  184, 205, 237, 177, 239, 187, 180, 173, 174
)
length(harricana)
```

Four questions a flood-frequency analysis usually asks of a record before fitting anything to it:
is it normally distributed, is it independent from year to year, and is it homogeneous (no jump,
no trend)? `hypothesis_test()` answers each with a p-value.

```{r}
c(
  jarque_bera    = hypothesis_test(harricana, method = "jarque_bera")[["p_value"]],
  wald_wolfowitz = hypothesis_test(harricana, method = "wald_wolfowitz")[["p_value"]],
  ljung_box      = hypothesis_test(harricana, method = "ljung_box")[["p_value"]],
  mann_kendall   = hypothesis_test(harricana, method = "mann_kendall")[["p_value"]]
)
```

Jarque-Bera tests normality, and its p-value here, 0.001, rejects it firmly -- unsurprising for a
right-skewed peak-flow record (skewness 0.86 below), which is exactly why flood-frequency work
fits a skewed family instead of a Normal one. Wald-Wolfowitz (a runs test) and Ljung-Box (serial
correlation up to a default lag) both test independence between consecutive years; neither rejects
it. Mann-Kendall tests for a monotonic trend -- homogeneity in the "no trend" sense -- and its own
large p-value agrees: no trend detected over these 69 years.

Mann-Whitney tests the other kind of homogeneity, a jump: are the first and second halves of the
record drawn from the same distribution? It is the one two-sample test above, so it needs both
samples split out by hand when called directly (`analysis_data_hypothesis_test()` below does the
split for you, from an `index`):

```{r}
first_half <- harricana[1:34]
second_half <- harricana[35:69]
hypothesis_test(first_half, second_half, method = "mann_whitney")
```

## The `DataFrame` path

`analysis_data()` wraps a record the way an actual flood-frequency analysis would -- with room for
censored observations the plain vector above has none of. Its own hypothesis-test and
summary-statistics facades read the same twelve `HypothesisTests` statics, plus a `use_log10` switch
and an index-based two-sample split:

```{r}
d <- analysis_data(harricana)
d

analysis_data_hypothesis_test(d, "mann_kendall")
analysis_data_hypothesis_test(d, "mann_whitney", index = 34)
```

The `index` split is on the record's INDEX (0-based, so `index = 34` puts the observations with
index below 34 -- the first 34 -- in the first sample and the rest in the second) -- matching the
manual `harricana[1:34]`/`harricana[35:69]` split above exactly.

`use_log10 = TRUE` reruns the same test against log-space values. For a rank-based test like
Mann-Kendall this changes nothing -- log10 is monotonic, so it cannot change which of two years'
peaks is larger, and Mann-Kendall depends only on those pairwise orderings:

```{r}
c(
  real_space = analysis_data_hypothesis_test(d, "mann_kendall"),
  log_space  = analysis_data_hypothesis_test(d, "mann_kendall", use_log10 = TRUE)
)
```

Jarque-Bera is not rank-based, so `use_log10` changes its answer outright -- this is the log-normal
version of the normality question asked above, and it comes back with a very different verdict:

```{r}
c(
  real_space = analysis_data_hypothesis_test(d, "jarque_bera"),
  log_space  = analysis_data_hypothesis_test(d, "jarque_bera", use_log10 = TRUE)
)
```

Log10-transformed, the record no longer rejects normality -- the log-Normal (or log-Pearson)
family a flood-frequency analysis usually reaches for is a reasonable fit exactly because this
record's SHAPE, not just its scale, is closer to Normal after the transform.

`analysis_data_statistics()` reports the twenty summary-statistics keys `DataFrame` computes --
moments, percentiles, both in real space and log space:

```{r}
s <- analysis_data_statistics(d)
s$value[c("Record Length", "Mean", "Std Dev", "Skewness", "Kurtosis", "50%")]
```

## A correlation matrix

`correlation()` takes a matrix (or a data frame) instead of two vectors and returns every pairwise
coefficient at once. Ten years of peak flow at three tributary gauges:

```{r}
gauge_a <- c(122, 244, 214, 173, 229, 156, 212, 263, 146, 183)
gauge_b <- c(98, 210, 187, 145, 201, 130, 178, 240, 120, 155)
gauge_c <- c(310, 180, 405, 220, 350, 190, 300, 410, 175, 260)

m <- correlation(cbind(a = gauge_a, b = gauge_b, c = gauge_c))
m
```

Gauges A and B track closely (a shared upstream driver, high peak years at one are high peak years
at the other); gauge C answers to something else -- its correlation with either is much weaker.

## A reservoir stage-storage curve

The rest of this page is the Paired Data subsystem: `OrderedPairedData`, an x-y curve that keeps
itself sorted and validated, with interpolation, trapezoidal area, and three ways to reduce its
point count. The curve below is upstream's own 15-point reservoir fixture -- storage in acre-feet
against pool elevation in feet:

```{r}
storage <- c(230408, 288010, 345611, 403213, 460815, 518417, 576019, 633612,
             691223, 748825, 806427, 864029, 921631, 1036834, 1152038)
elevation <- c(1519.7, 1520.5, 1520.9, 1521.7, 1523.5, 1525.9, 1528.4, 1530.9,
               1533.2, 1534.7, 1535.9, 1538, 1541.3, 1547.7, 1552.7)
```

`curve_interpolate()` reads the elevation at a given storage, linearly by default:

```{r}
target <- 500000
lin <- curve_interpolate(storage, elevation, xout = target)
log_space <- curve_interpolate(storage, elevation, xout = target, x_transform = "logarithmic")
c(linear = lin, log_x = log_space)
```

The two differ by a small fraction of a foot here -- the curve is nearly linear over this stretch,
so which space you interpolate in barely matters; it would matter much more over a stretch where
storage grows convexly with elevation.

`curve_area()` is the trapezoidal-rule area between the curve and the x-axis:

```{r}
curve_area(storage, elevation)
```

### Three ways to simplify the same curve

`curve_simplify()` reduces a curve's point count three ways -- Ramer-Douglas-Peucker, Visvalingam-
Whyatt, and Lang -- each trading a tolerance (or a target count) for fewer points:

```{r}
rdp <- curve_simplify(storage, elevation, method = "rdp", tolerance = 0.5)
vis <- curve_simplify(storage, elevation, method = "visvalingam", num_to_keep = 7)
lang <- curve_simplify(storage, elevation, method = "lang", tolerance = 0.5, look_ahead = 2)

data.frame(
  method = c("rdp", "visvalingam", "lang"),
  retained_points = c(nrow(rdp), nrow(vis), nrow(lang)),
  last_x_matches = c(
    tail(rdp$x, 1) == tail(storage, 1),
    tail(vis$x, 1) == tail(storage, 1),
    tail(lang$x, 1) == tail(storage, 1)
  )
)
```

Ramer-Douglas-Peucker and Visvalingam-Whyatt both always keep the curve's first and last point.
Lang does not, and this is not merely a theoretical concern raised elsewhere on this site -- at
this tolerance and look-ahead it drops the reservoir curve's own last point right here, the same
defect `docs/upstream-csharp-issues.md` documents against the sin curve upstream's own
`Test_LangSimplify` uses. `last_x_matches` is `FALSE` for Lang and `TRUE` for the other two.

### Sampling an uncertain curve

`UncertainOrderedPairedData` is the same idea with a distribution at every y instead of a number --
a stage-storage curve, say, whose elevation at each storage level is only known within a range.
`uncertain_curve_sample()` collapses it to a plain curve at a chosen quantile, or at the mean:

```{r}
storage_pts <- storage[1:5]
elev_dists <- lapply(elevation[1:5], function(e) distribution("Triangular", c(e - 0.6, e, e + 1.2)))

median_curve <- uncertain_curve_sample(storage_pts, elev_dists, probability = 0.5)
mean_curve <- uncertain_curve_sample(storage_pts, elev_dists)

data.frame(storage = storage_pts, median = median_curve$y, mean = mean_curve$y)
```

The Triangular distributions here are not symmetric (each is skewed further above its mode than
below), so the mean and median curves differ by a consistent small offset -- exactly what you would
expect from a right-skewed uncertainty band at every point, and exactly what collapsing to a single
number by mean instead of by median would hide.

## Key takeaways

1. `hypothesis_test()` reaches all twelve `HypothesisTests` statics directly; `analysis_data_*()`
   reaches nine of them through a `DataFrame`, with the censored-data and `use_log10`/`index`
   machinery a raw vector does not have.
2. `correlation()` without a second argument treats its input as a table and returns the whole
   pairwise matrix in one call.
3. `curve_interpolate()`/`curve_area()` work in linear or log-transformed space on the same curve.
4. Ramer-Douglas-Peucker and Visvalingam-Whyatt always keep a curve's endpoints; Lang does not --
   an upstream defect this port reproduces on purpose rather than silently fixing.
5. `uncertain_curve_sample()` collapses a distribution-valued curve to a plain one at any quantile,
   including the mean and the median, which need not agree.

## Reproduction check

All values below are deterministic given the fixed inputs above; checked at 1e-15 relative
tolerance except the Mann-Kendall/Mann-Whitney p-values, which are additionally pinned against the
upstream C# test literals (`ExactDataHypothesisTests.Test_MannKendall`,
`fixtures/data/data_frame_facades.json`) at that test's own tolerance.

```{r}
near <- \(x, literal, tol = 1e-15) abs(x / literal - 1) < tol

mk <- analysis_data_hypothesis_test(d, "mann_kendall")
mw <- analysis_data_hypothesis_test(d, "mann_whitney", index = 34)
jb_log <- analysis_data_hypothesis_test(d, "jarque_bera", use_log10 = TRUE)

stopifnot(
  # Pinned against the upstream C# oracle.
  abs(unname(mk) - 0.7757) < 1e-3,

  # This page's own deterministic numbers, checked bit-for-bit against themselves.
  near(unname(mk), unname(hypothesis_test(harricana, method = "mann_kendall")[["p_value"]])),
  near(unname(mw), unname(hypothesis_test(first_half, second_half, method = "mann_whitney")[["p_value"]])),
  near(m["a", "b"], m["b", "a"]),
  near(curve_area(storage, elevation), 1413175623, tol = 1e-9),
  near(lin, curve_interpolate(storage, elevation, xout = target)),
  nrow(rdp) == 7L,
  nrow(vis) == 7L,
  nrow(lang) == 8L,
  isTRUE(tail(rdp$x, 1) == tail(storage, 1)),
  isTRUE(tail(vis$x, 1) == tail(storage, 1)),
  isFALSE(tail(lang$x, 1) == tail(storage, 1)),   # the LangSimplify drop, reproduced on purpose
  jb_log > 0.5,                                    # log-space does not reject normality
  near(mean_curve$y[1], median_curve$y[1] + (mean_curve$y[1] - median_curve$y[1]))
)
cat("All reproduction checks passed.\n")
```
