---
title: "20. Bayesian regression"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Ported from
[`bayesian_regression.py`](https://github.com/USACE-RMC/Numerics-Python-Examples/blob/main/examples/bayesian_regression.py)
in the USACE-RMC Numerics-Python-Examples repository (0BSD licensed). The upstream
script drives the C# `Numerics.dll` through pythonnet, handing the `DEMCzs` sampler a
list of hand-picked `Uniform` priors and a Python log-likelihood callback. `corehydror`
takes that pattern too, through `mcmc_posterior()`
([example 14](../14-custom-posterior/r.qmd)), but a linear regression has a model of
its own here, so this version recasts it through `arimax_analysis` (details below) and
the reproduction is statistical rather than exact. The synthetic data, however, are
bit-identical to the upstream script's, and the Python version of this example uses
the same core and prints the same numbers.

## What you'll learn

- How to regenerate the upstream script's synthetic data exactly (same seeded
  Mersenne Twister).
- How to express a Bayesian linear regression as an ARIMAX(0,0,0) model with a
  linear time trend.
- What the returned curves are: the MAP fitted line and a posterior predictive band.

## Synthetic data

The upstream script simulates a straight line with Normal noise:
`x = linspace(0, 10, 80)` and `y = a + b*x + Normal(0, sigma)` with
`a = 2.0`, `b = 1.4`, `sigma = 1.2`, drawing the noise with seed 123. The port
keeps the C# Mersenne Twister bit-exact, so `dist_random` with the same seed
reproduces the upstream noise vector to the last bit. We write `x` as `10*t/79`
for the time index `t = 0..79`, which is the same vector as the upstream
`linspace` and makes the mapping used below explicit.

One quirk worth knowing about, since the upstream run sees the identical draw: this
particular seeded noise vector runs light. Its sample standard deviation is 0.904,
well under the nominal 1.2, so any faithful posterior for `sigma` concentrates near
0.9.

```{r}
library(corehydror)

n <- 80
true_a <- 2.0
true_b <- 1.4
true_sigma <- 1.2

t_idx <- 0:(n - 1)        # time index 0..79
x <- 10 * t_idx / 79      # identical to the upstream linspace(0, 10, 80)

noise <- dist_random(distribution("Normal", c(0, true_sigma)), n, seed = 123)
y <- true_a + true_b * x + noise

cat("first observations:", round(y[1:4], 8), "\n")
cat(sprintf("realized noise sd: %.4f (nominal %.1f)\n", sd(noise), true_sigma))
```

## Recasting the regression

The upstream script builds a C# `List` of three priors (`Uniform(-10, 10)` for `a`,
`Uniform(0, 5)` for `b`, `Uniform(0.1, 5)` for `sigma`) and a Python callback that
sums `Normal(0, sigma).LogPDF(residual)` over the data. Neither custom priors nor
likelihood callbacks are part of the public API here. But regression on the time
index is a built-in model: `arimax_analysis` with orders `p = d = q = 0`, an
intercept, and `trend = "Linear"` fits

$$y_t = \mu + \gamma t + \varepsilon_t, \qquad \varepsilon_t \sim \text{Normal}(0, \sigma),$$

which is Bayesian simple linear regression, sampled with the same DEMCzs sampler
the upstream uses. Because `x = 10t/79` is an affine map of the time index, the
upstream parameters translate directly:

- intercept: $a = \mu$, so the true value is 2.0;
- slope: $\gamma = b \cdot 10/79$, so the true trend per time step is
  $1.4 \cdot 10/79 = 0.17722$, and the fitted slope maps back as
  $b = \gamma \cdot 79/10$;
- `sigma` is the residual standard deviation in both parameterizations.

Two honest differences from the upstream run. First, the priors are the model's own
data-scaled uniforms (order-of-magnitude bounds built around a least-squares
initialization), not the upstream's hand-picked ones, so posterior numbers can only
be compared statistically. Second, the runtime: the upstream script takes about five
minutes because every likelihood evaluation crosses back into Python; here the
likelihood is compiled, and the modest settings below finish in about a second.

`parameters` in the returned list is the posterior mode (MAP) vector, in model
order: intercept $\mu$, trend $\gamma$, then $\sigma$ last.

```{r}
fit <- arimax_analysis(
  y,
  order_p = 0L, order_d = 0L, order_q = 0L, # no AR, differencing, or MA terms
  trend = "Linear",                         # deterministic trend gamma * t
  include_intercept = TRUE,                 # intercept mu
  training_time_steps = n,                  # fit on all 80 points
  forecasting_time_steps = 0L,
  sampler = "DEMCzs",                       # the sampler the upstream script uses
  iterations = 2000L,
  output_length = 5000L,
  credible_level = 0.90,
  seed = 45678L
)

mu_map <- fit$parameters[1]
gamma_map <- fit$parameters[2]
sigma_map <- fit$parameters[3]
b_map <- gamma_map * 79 / 10 # slope per time step -> slope per unit x

summary_df <- data.frame(
  parameter = c("intercept a", "slope b", "slope per step gamma", "sigma"),
  truth = c(true_a, true_b, true_b * 10 / 79, true_sigma),
  MAP = c(mu_map, b_map, gamma_map, sigma_map)
)
print(summary_df, row.names = FALSE)
cat(sprintf("\nrmse: %.4f\n", fit$rmse))
```

All three parameters land close to the generating values, with `sigma` tracking
the realized noise level (0.904) rather than the nominal 1.2, exactly as the
upstream run does with the same data.

## Fitted line and band

What the returned curves are, since they differ from the upstream plot:

- `mode_curve` is the deterministic fitted line $\mu + \gamma t$ evaluated at the
  MAP parameters.
- `mean_curve`, `lower_ci`, and `upper_ci` come from a posterior predictive
  ensemble: each posterior draw simulates a full series including observation
  noise. The 90% band is therefore a predictive band for observations, wider than
  the mean-function credible band the upstream script plots.

The upstream 2x2 figure also shows posterior histograms of `a`, `b`, and `sigma`;
the analysis returns curves and MAP values rather than raw chains, so those two
panels cannot be reproduced here.

```{r}
true_line <- true_a + true_b * x

op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))

plot(x, y,
  pch = 16, cex = 0.6, col = "#8c8c7a",
  ylim = range(fit$lower_ci, fit$upper_ci, y),
  xlab = "x", ylab = "y", main = "Bayesian regression fit"
)
polygon(c(x, rev(x)), c(fit$lower_ci, rev(fit$upper_ci)),
  col = adjustcolor("#5b7a8c", alpha.f = 0.25), border = NA
)
points(x, y, pch = 16, cex = 0.6, col = "#8c8c7a")
lines(x, true_line, col = "#b06a3b", lty = 2, lwd = 1.5)
lines(x, fit$mode_curve, col = "#6b7f3f", lwd = 2)
legend("topleft",
  legend = c("Observed", "True line", "MAP fit", "90% predictive band"),
  col = c("#8c8c7a", "#b06a3b", "#6b7f3f", adjustcolor("#5b7a8c", alpha.f = 0.25)),
  pch = c(16, NA, NA, 15), lty = c(NA, 2, 1, NA), lwd = c(NA, 1.5, 2, NA),
  bty = "n", cex = 0.7
)

residuals <- y - fit$mode_curve
plot(fit$mode_curve, residuals,
  pch = 16, cex = 0.6, col = "#6b7f3f",
  xlab = "Fitted", ylab = "Residual", main = "Residuals vs fitted"
)
abline(h = 0, col = "#b06a3b", lty = 2, lwd = 1.5)

par(op)
```

```{r}
inside <- (y >= fit$lower_ci) & (y <= fit$upper_ci)
covers_true <- (true_line >= fit$lower_ci) & (true_line <= fit$upper_ci)
cat(sprintf("observations inside the 90%% predictive band: %.4f\n", mean(inside)))
cat(sprintf("ordinates where the band covers the true line: %d of %d\n", sum(covers_true), n))
```

## Reproduction check

The upstream file is a script with no committed outputs, so there are no upstream
posterior literals to compare against; the reference values are the true generating
parameters. The priors and the parameterization deliberately differ (data-scaled
uniform priors, time-index trend), so bit-comparison with the upstream posterior is
impossible and the parameter checks are statistical. The exact tier covers what is
exactly checkable: the seeded synthetic data (the same Mersenne Twister stream the
upstream draws) and cross-language identity of the whole analysis (the Python twin
asserts the same MAP literals from the same seed).

| Quantity | Upstream C# | This port | Status |
|----------|------------:|----------:|--------|
| Seeded data `y[1]` (seed 123) | same MT stream | 2.61712675977869 | exact |
| Seeded data `y[80]` | same MT stream | 15.89237584375904 | exact |
| Intercept `a` | 2.0 (true) | MAP 2.0437 | statistical |
| Slope `b` | 1.4 (true) | MAP 1.3916 | statistical |
| `sigma` | 1.2 nominal, 0.9037 realized | MAP 0.9185 | statistical |
| 90% band covers the true line | - | 80 of 80 ordinates | statistical |

The chunk below fails the render if any value drifts.

```{r}
# Upstream: examples/bayesian_regression.py (a script with no committed outputs;
# the reference values are the true generating parameters, not upstream posteriors).

# Exact tier: seeded data and the fit itself, identical across R, Python, and the
# C# library. The values are bit-exact (the fixture suite enforces that); the
# comparisons carry a 1e-15 relative tolerance only because R's decimal parser can
# land one ulp away from the written literal.
stopifnot(
  abs(y[1] / 2.6171267597786905 - 1) < 1e-15,
  abs(y[n] / 15.892375843759043 - 1) < 1e-15,
  abs(mu_map / 2.043676501381344 - 1) < 1e-15,
  abs(gamma_map / 0.1761574955229039 - 1) < 1e-15
)

# Statistical tier: recovered parameters near the truth. Priors and parameterization
# deliberately differ from the upstream script, so only closeness can be asserted.
stopifnot(
  abs(mu_map - true_a) < 0.5,
  abs(b_map - true_b) < 0.1,
  sigma_map > 0.7, sigma_map < 1.2 # tracks the realized noise sd, 0.9037
)

# Statistical tier: the 90% predictive band covers the true line everywhere and
# most observations.
stopifnot(
  all(covers_true),
  mean(inside) > 0.8
)
cat("All reproduction checks passed.\n")
```
