---
title: "21. Flood frequency analysis"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Ported from
[`flood_frequency_analysis.py`](https://github.com/USACE-RMC/Numerics-Python-Examples/blob/main/examples/flood_frequency_analysis.py)
in the USACE-RMC Numerics-Python-Examples repository (0BSD licensed). The upstream
file is a standalone script, not a notebook: it drives the C# `Numerics.dll` through
pythonnet, prints two tables to the terminal, and pops up a figure. This version uses
`corehydror`, whose compiled core is a validated C++ port of the same library, so the
deterministic fits and tables below are the values that script prints. The Python
version of this example uses the same core and prints the same numbers.

## What you'll learn

- Fit several candidate distributions to annual peak flows, each with the estimation
  method the upstream script uses.
- Rank the fits and pick a model.
- Build a design-flow table at standard return periods and plot frequency curves.

## Setup

The upstream script opens with roughly fifty lines of .NET plumbing: loading the
CoreCLR runtime, resolving `Numerics.dll` from the NuGet cache, and converting arrays
to `System.Array[Double]`. None of that exists here.

```{r}
library(corehydror)
```

## The data

Forty-eight years of annual peak flows in cubic feet per second, hardcoded in the
upstream script.

```{r}
peak_flows <- c(
  6290, 2700, 13100, 16900, 14600, 9600, 7740, 8490, 8130, 12000,
  17200, 15000, 12400, 6960, 6500, 5840, 10400, 18800, 21400, 22600,
  14200, 11000, 12800, 15700, 4740, 6950, 11800, 12100, 20600, 14600,
  14600, 8900, 10600, 14200, 14100, 14100, 12500, 7530, 13400, 17600,
  13400, 19200, 16900, 15500, 14500, 21900, 10400, 7460
)
cat(sprintf(
  "n = %d, min = %s, mean = %s, max = %s cfs\n",
  length(peak_flows),
  format(min(peak_flows), big.mark = ","),
  format(round(mean(peak_flows)), big.mark = ","),
  format(max(peak_flows), big.mark = ",")
))
```

## Fit three candidate models

The upstream script calls `dist.Estimate(data, ParameterEstimationMethod.X)` on three
freshly constructed C# objects. Here `dist_fit(family, data, method =)` does the same
in one call, with the same three methods: LogNormal by maximum likelihood, GEV by
L-moments, Weibull by maximum likelihood. (GEV also has a dedicated `gev_fit`
helper whose bespoke path carries quantile standard errors.)

As in the C# library, `LogNormal` is the base-10 family, so its two parameters are
the mean and standard deviation of `log10(flow)`.

```{r}
ln <- dist_fit("LogNormal", peak_flows, method = "mle")
gev <- dist_fit("GeneralizedExtremeValue", peak_flows, method = "lmom")
wb <- dist_fit("Weibull", peak_flows, method = "mle")

models <- list(LogNormal = ln, GEV = gev, Weibull = wb)
for (name in names(models)) {
  cat(sprintf(
    "%9s: %s\n", name,
    paste(sprintf("%.6g", dist_params(models[[name]])), collapse = ", ")
  ))
}
```

## Goodness of fit

The upstream script ranks the fits with the Kolmogorov-Smirnov statistic from
`Numerics.Data.Statistics.GoodnessOfFit`. That helper is not part of the ported API,
so this version ranks by log-likelihood and AIC instead. The log-likelihoods are the
exact values the script also prints, and the conclusion is the same either way:
Weibull edges out GEV, and LogNormal trails.

```{r}
log_liks <- sapply(models, \(d) dist_log_likelihood(d, peak_flows))
n_params <- sapply(models, \(d) length(dist_params(d)))

fit_df <- data.frame(
  Model = names(models),
  LogLikelihood = log_liks,
  AIC = 2 * n_params - 2 * log_liks,
  row.names = NULL
)
fit_df <- fit_df[order(fit_df$AIC), ]
fit_df$dAIC <- fit_df$AIC - min(fit_df$AIC)

cat("Model fit summary (lower AIC is better):\n")
print(fit_df, row.names = FALSE, digits = 8)
```

The packages also ship this ranking as a one-liner: `fit_distributions` fits all 15
supported candidates by maximum likelihood and returns AIC, BIC, and RMSE for each.
(It refits GEV by MLE, so its AIC differs slightly from the L-moment fit above.)

```{r}
ranking <- as.data.frame(fit_distributions(peak_flows))
ranking <- ranking[order(ranking$aic), ]
print(head(ranking, 5), row.names = FALSE, digits = 6)
```

## Design flow table

Design flows at the standard return periods. A T-year event has annual exceedance
probability 1/T, so the design flow is the quantile at `1 - 1/T`; the upstream
script's `InverseCDF(p)` is `dist_quantile(d, p)` here.

```{r}
return_periods <- c(2, 5, 10, 25, 50, 100)
probs <- 1 - 1 / return_periods

design <- data.frame(
  ReturnPeriod = return_periods,
  lapply(models, \(d) dist_quantile(d, probs))
)
cat("Design flow table (cfs):\n")
print(round(design, 1), row.names = FALSE)
```

The three models agree closely through the 10-year event and then diverge: the
unbounded LogNormal keeps climbing while the L-moment GEV (positive shape, so upper
bounded) and Weibull flatten. This tail disagreement is exactly why flood frequency
practice cares so much about distribution choice.

## Visualizing the fits

The upstream script ends with a 2x2 figure: fitted PDFs over a histogram, fitted
CDFs against the empirical CDF, the frequency curves, and a goodness-of-fit
comparison. Two substitutions here: the empirical CDF uses Weibull plotting
positions `i / (n + 1)` from `plotting_positions` instead of the script's naive
`i / n` (which forces the largest observation to probability 1), and the
goodness-of-fit panel shows dAIC instead of the unported KS statistic.

```{r}
colors <- c(LogNormal = "#6b7f3f", GEV = "#b06a3b", Weibull = "#5b7a8c")
sorted_flows <- sort(peak_flows)
pp <- plotting_positions(length(peak_flows))
x_grid <- seq(min(peak_flows) * 0.8, max(peak_flows) * 1.1, length.out = 500)

op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))

# 1) Histogram + fitted PDFs
hist(peak_flows, breaks = 16, freq = FALSE, col = adjustcolor("#6b7f3f", 0.4),
     border = "white", main = "Fitted PDFs", xlab = "Peak flow (cfs)")
for (name in names(models)) {
  lines(x_grid, dist_pdf(models[[name]], x_grid), col = colors[[name]], lwd = 2)
}
legend("topleft", legend = names(models), col = colors, lwd = 2, bty = "n", cex = 0.8)

# 2) Plotting positions + fitted CDFs
plot(sorted_flows, pp, pch = 16, cex = 0.6, col = "#8c8c7a",
     main = "Empirical vs fitted CDFs", xlab = "Peak flow (cfs)", ylab = "CDF")
for (name in names(models)) {
  lines(x_grid, dist_cdf(models[[name]], x_grid), col = colors[[name]], lwd = 2)
}
legend("topleft", legend = c("Plotting positions", names(models)),
       col = c("#8c8c7a", colors), pch = c(16, NA, NA, NA),
       lwd = c(NA, 2, 2, 2), bty = "n", cex = 0.8)

# 3) Frequency curves
matplot(return_periods, design[names(models)], type = "b", pch = 16, lty = 1,
        lwd = 2, col = colors, log = "x", main = "Flood frequency curves",
        xlab = "Return period (years)", ylab = "Design flow (cfs)")
legend("topleft", legend = names(models), col = colors, lwd = 2, bty = "n", cex = 0.8)

# 4) Goodness-of-fit comparison
barplot(fit_df$dAIC, names.arg = fit_df$Model, col = colors[fit_df$Model],
        border = NA, main = "Goodness of fit (dAIC, lower is better)",
        ylab = "dAIC")

par(op)
```

## What the packages add

The upstream script assembles this analysis by hand from Numerics building blocks.
The packages also ship complete flood-frequency workflows: `bulletin17c_analysis`
runs the Bulletin 17C log-Pearson Type III analysis (the US federal standard,
including confidence intervals), and `univariate_analysis` fits any family by
Bayesian MCMC with a full credible band. The 100-year estimate below lands between
the bounded GEV/Weibull tails and the LogNormal tail, and comes with uncertainty
attached:

```{r}
b17c <- bulletin17c_analysis(peak_flows)
i <- which(b17c$exceedance_probabilities == 0.01)
q100 <- 10^b17c$point_estimates[i] # the B17C curve is tabulated in log10 space
cat(sprintf(
  "Bulletin 17C 100-yr flow: %s cfs (90%% CI %s to %s)\n",
  format(round(q100), big.mark = ","),
  format(round(b17c$lower_ci[i]), big.mark = ","),
  format(round(b17c$upper_ci[i]), big.mark = ",")
))
```

## Reproduction check

The upstream file is a script with no committed outputs, so there are no embedded
numbers to quote; it prints its fit summary and design-flow table when run against
the C# library. Every quantity in this example is deterministic (no sampling), so
the ported pipeline reproduces those printed tables exactly. The assertions below
pin the same literals the Python notebook asserts (the cross-language identity) and
check the quantile/CDF round trip for internal consistency.

| Quantity | Upstream C# | This port | Status |
|----------|-------------|----------:|--------|
| LogNormal MLE parameters | printed at run time, not embedded | 4.067428, 0.186794 | exact |
| GEV L-moment parameters | printed at run time, not embedded | 10916.18, 4683.918, 0.251455 | exact |
| Weibull MLE parameters | printed at run time, not embedded | 14197.26, 2.982835 | exact |
| Log-likelihoods (LN, GEV, WB) | printed at run time, not embedded | -477.1525, -473.2227, -473.0458 | exact |
| 100-yr design flows (LN, GEV, WB) | printed at run time, not embedded | 31767.1, 23684.9, 23689.6 | exact |
| `dist_quantile(dist_cdf(x))` round trip | n/a (internal consistency) | rel. error < 1e-12 | exact |

The chunk below fails the render if any value drifts.

```{r}
# Upstream: examples/flood_frequency_analysis.py (standalone script; its tables are
# printed at run time, not embedded in the repo). The pipeline is deterministic, so
# these pinned values are what that script computes; the Python notebook asserts the
# same literals, which proves the cross-language identity. The values are bit-exact
# (the fixture suite enforces that); the comparisons carry a 1e-15 relative
# tolerance only because R's decimal parser can land one ulp away from the written
# literal.
near <- \(x, literal) abs(x / literal - 1) < 1e-15
stopifnot(
  near(dist_params(ln)[1], 4.067428305746908),
  near(dist_params(ln)[2], 0.18679432530177564),
  near(dist_params(gev)[[1]], 10916.176130989406),
  near(dist_params(gev)[[2]], 4683.917553015235),
  near(dist_params(gev)[[3]], 0.25145450827789206),
  near(dist_params(wb)[1], 14197.259246092635),
  near(dist_params(wb)[2], 2.98283522441679),

  near(dist_log_likelihood(ln, peak_flows), -477.15248319801253),
  near(dist_log_likelihood(gev, peak_flows), -473.2226655186914),
  near(dist_log_likelihood(wb, peak_flows), -473.045841282679),

  near(dist_quantile(ln, 1 - 1 / 100), 31767.050051870196),
  near(dist_quantile(gev, 1 - 1 / 100), 23684.9359641315),
  near(dist_quantile(wb, 1 - 1 / 100), 23689.649055911825)
)

# Internal consistency: quantile inverts the CDF at every observation, and each
# frequency curve is strictly increasing in return period.
for (d in models) {
  stopifnot(all(abs(dist_quantile(d, dist_cdf(d, peak_flows)) / peak_flows - 1) < 1e-12))
}
stopifnot(sapply(design[names(models)], \(q) all(diff(q) > 0)))

# Ranking: Weibull wins by AIC, both among the three fits here and among the 15
# MLE candidates in fit_distributions.
stopifnot(
  fit_df$Model[1] == "Weibull",
  ranking$distribution[1] == "Weibull"
)
cat("All reproduction checks passed.\n")
```
