---
title: "14. A custom posterior"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

`mcmc_sample()` samples the posterior of a built-in distribution family under uniform priors
spanning that family's parameter constraints. `mcmc_posterior()` drops both restrictions: you
write the log-likelihood, you choose the priors, and all eight ported samplers become
reachable, including Gibbs, which neither package could run before this surface existed because
Gibbs needs a proposal function and there was no way to pass one in.

## What you'll learn

- Write a log-likelihood by hand and sample its posterior with `mcmc_posterior()`.
- Read the trace plot, R-hat and posterior summary the same way you would for a built-in model.
- Where the cross-language guarantee stops, and why a callback posterior is the place it stops
  hardest.
- Run the Gibbs sampler with a proposal function, drawing from the handle the sampler gives you
  rather than from R's own generator.

## Setup

```{r}
library(corehydror)
```

## A model the package does not have

Runoff against basin precipitation at a gauge, twelve water years, in inches. The model is a
straight line with Gaussian errors and three parameters: intercept $a$, slope $b$, and error
scale $\sigma$. No `model_*()` constructor builds this, so `fit_bayesian()` cannot reach it.

$$
y_i = a + b\,x_i + \varepsilon_i, \qquad \varepsilon_i \sim \mathcal{N}(0, \sigma^2)
$$

```{r}
precip <- c(18.2, 22.5, 25.1, 19.8, 30.4, 27.6, 21.3, 33.0, 24.7, 28.9, 20.6, 31.5)
runoff <- c(4.1, 6.8, 8.0, 5.2, 11.9, 9.6, 6.0, 13.4, 7.7, 10.5, 5.5, 12.6)
n <- length(runoff)

plot(precip, runoff, pch = 19, col = "#5b7a8c",
     xlab = "Basin precipitation (in)", ylab = "Runoff (in)",
     main = "Twelve water years")
```

Dropping the constant $-\tfrac{n}{2}\log(2\pi)$, the log-likelihood is

$$
\ell(a, b, \sigma) = -n \log \sigma
  - \frac{1}{2\sigma^2} \sum_{i=1}^{n} \bigl(y_i - a - b\,x_i\bigr)^2 .
$$

Written as a plain loop, which is deliberate and is explained in the next section.

```{r}
log_likelihood <- function(p) {
  a <- p[1]
  b <- p[2]
  sigma <- p[3]
  acc <- 0
  for (i in seq_len(n)) {
    resid <- runoff[i] - (a + b * precip[i])
    acc <- acc + resid * resid
  }
  -n * log(sigma) - acc / (2 * sigma * sigma)
}
```

One prior per parameter, in the order `log_likelihood` reads them. These are flat, so the
log-likelihood carries no prior term: a parameter outside its prior's support is rejected before
your function is called, and the priors are used for the feasible bounds and for chain
initialization. A prior that is not flat would need its log density added to the returned value
yourself, which is the upstream contract rather than a corehydro choice.

```{r}
priors <- list(
  distribution("Uniform", c(-20, 10)),   # intercept
  distribution("Uniform", c(0, 2)),      # slope
  distribution("Uniform", c(0.01, 5))    # error scale
)

fit <- mcmc_posterior(log_likelihood, priors, sampler = "RWMH",
                      iterations = 1000, warmup = 500, chains = 3, thinning = 1,
                      seed = 12345)
```

Keep the run small. Every evaluation crosses back into R, and the count is
`(iterations + output_length / chains) * thinning * chains`, so a thinning interval or an
unset `iterations` costs far more than it looks like it should. This fit takes about a third of
a second.

### Trace plot

```{r}
#| fig-height: 7
labels <- c("intercept a", "slope b", "sigma")
colors <- c("#5b7a8c", "#b06a3b", "#6b7f3f")
par(mfrow = c(3, 1), mar = c(4, 4, 2, 1))
for (j in 1:3) {
  draws <- sapply(fit$chains, function(ch) ch[, j])
  matplot(draws, type = "l", lty = 1, col = colors, xlab = "Draw", ylab = labels[j],
          main = labels[j])
}
```

### R-hat and the posterior summary

R-hat compares between-chain and within-chain variance. Values near 1 mean the three chains have
settled on the same distribution.

```{r}
summary_table <- data.frame(
  parameter = labels,
  mean = round(fit$posterior_mean, 4),
  sd = round(fit$posterior_sd, 4),
  median = round(fit$posterior_median, 4),
  lower = round(fit$posterior_lower_ci, 4),
  upper = round(fit$posterior_upper_ci, 4),
  rhat = round(fit$rhat, 4),
  ess = round(fit$ess)
)
print(summary_table, row.names = FALSE)
```

The posterior mode is the MAP, reported alongside the value of your log-likelihood there.
Ordinary least squares is a useful cross-check on the two regression parameters, since with a
flat prior the mode of $(a, b)$ given $\sigma$ is the least-squares fit.

```{r}
ols <- coef(lm(runoff ~ precip))
cat(sprintf("MAP:  a = %.6f, b = %.6f, sigma = %.6f\n",
            fit$map[1], fit$map[2], fit$map[3]))
cat(sprintf("OLS:  a = %.6f, b = %.6f\n", ols[1], ols[2]))
cat(sprintf("log-likelihood at the MAP: %.6f\n", fit$map_fitness))
```

### The fitted line with its credible band

```{r}
grid <- seq(min(precip), max(precip), length.out = 100)
draws <- do.call(rbind, fit$chains)
band <- sapply(grid, function(xi) quantile(draws[, 1] + draws[, 2] * xi, c(0.025, 0.5, 0.975)))

plot(precip, runoff, pch = 19, col = "#5b7a8c",
     xlab = "Basin precipitation (in)", ylab = "Runoff (in)",
     main = "Posterior median and 95% credible band")
polygon(c(grid, rev(grid)), c(band[1, ], rev(band[3, ])),
        col = "#b06a3b33", border = NA)
lines(grid, band[2, ], col = "#b06a3b", lwd = 2)
points(precip, runoff, pch = 19, col = "#5b7a8c")
```

## Where the cross-language guarantee stops

corehydro promises that a seeded run gives the same numbers in R and in Python. On the built-in
path that promise is unconditional: `mcmc_sample()`, `fit_bayesian()`, `dist_random()` and the
rest do every arithmetic operation inside the shared C++ core, so the two languages are running
the same compiled code and there is nothing left to disagree about.

`mcmc_posterior()` is weaker, and the difference deserves a section rather than a footnote. The
draws still come from the core's seeded Mersenne Twister, so the proposals are identical. The
log-density is not: it is your own R arithmetic, evaluated by R, and R and Python do not
guarantee identical rounding for the same formula. MCMC amplifies that without mercy. A single
differing bit in one log-density can flip one accept-or-reject decision, and every state after
it is then drawn from a different point, so the two chains diverge outright instead of drifting
apart slowly. There is no tolerance that repairs it.

A seeded run reproduces across the two languages if and only if your function returns
bit-identical values. Three practical rules follow.

1. `+`, `-`, `*` and `/` are IEEE-deterministic. They reproduce.
2. `log`, `exp`, `sqrt` of a non-exact value, `gamma` and friends come from each platform's own
   math library. They are not guaranteed to agree to the last bit, in either direction, and
   nothing in corehydro can make them.
3. R's `sum()` and `mean()` accumulate in extended precision. Python's do not. An explicit loop
   is the portable spelling in both languages, which is why the log-likelihood above is written
   as one.

The regression above obeys rules 1 and 3 but not rule 2: it calls `log(sigma)`. Measured on this
machine, the R and Python versions of that page happen to return identical posteriors to the
last bit, and the reproduction check below asserts those numbers as a within-R regression rather
than as a cross-language claim, because the agreement is an observation about two math libraries
and not something the package guarantees. The Gibbs model in the next section uses arithmetic and
comparisons alone, which is why it is the one this page pins against the shared oracle.

## Gibbs, the sampler that was out of reach

Gibbs replaces the Metropolis accept-or-reject step with a draw from the full conditional of the
model, which it accepts unconditionally. That draw is model-specific, so there is nothing
sensible to default it to, and until `mcmc_posterior()` existed neither package could run the
ported Gibbs sampler at all.

The model is small enough that its full conditional is exact. Five observations, each Uniform on
$(\mu - 1, \mu + 1)$. Dropping the normalizer, the log-likelihood is $0$ when every
$|x_i - \mu| \le 1$ and $-\infty$ otherwise, so under a flat prior the conditional of $\mu$ is
Uniform on $(\max x - 1, \min x + 1)$, here Uniform(4.2, 5.8). One uniform draw, rescaled, is
the Gibbs step.

```{r}
x <- c(4.9, 5.1, 5.0, 5.2, 4.8)

uniform_width <- function(p) {
  mu <- p[1]
  for (xi in x) {
    d <- xi - mu
    if (d > 1 || d < -1) return(-Inf)
  }
  0
}

full_conditional <- function(parameters, rng) {
  lo <- max(x) - 1
  hi <- min(x) + 1
  lo + rng_uniform(rng, 1) * (hi - lo)
}
```

`rng` is a handle on the generator this chain is already running on, not a generator of your
own. Drawing with `runif()` instead is not an error and nothing will warn you, but it silently
breaks two guarantees: the run stops being reproducible from its seed, and it stops agreeing
with the same run in Python. `rng_uniform(rng, n)` returns `n` values on `[0, 1)`;
`rng_integers(rng, n, min, max)` returns whole numbers with `min` included and `max` excluded.
The handle is borrowed for the duration of the one call it was given to, so storing it and
drawing from it later raises an error rather than reading freed memory.

```{r}
gibbs <- mcmc_posterior(uniform_width, distribution("Uniform", c(0, 10)),
                        sampler = "Gibbs", proposal = full_conditional,
                        iterations = 300, warmup = 100, thinning = 1,
                        output_length = 100, seed = 12345, initialize = "Randomize")

chain <- gibbs$chains[[1]]
cat(sprintf("draws: %d, parameters: %d, chains: %d\n",
            nrow(chain), ncol(chain), length(gibbs$chains)))
cat(sprintf("acceptance rate: %g\n", gibbs$acceptance_rates[1]))
cat(sprintf("MAP: %.16g   log-likelihood there: %g\n", gibbs$map[1], gibbs$map_fitness))
cat(sprintf("first, middle and last states: %.16g, %.16g, %.16g\n",
            chain[1, 1], chain[151, 1], chain[300, 1]))
```

Two of those numbers look wrong at first and are not. The acceptance rate is exactly 0 because
Gibbs draws from the exact conditional and the upstream sampler counts no Metropolis accepts at
all. The log-likelihood at the MAP is exactly 0 because this kernel returns 0 everywhere inside
its support, which also means the MAP is whichever state the chain visited first at that height,
making it a pin on the order of the draws rather than on their values.

```{r}
hist(chain[, 1], breaks = 20, col = "#5b7a8c", border = "white",
     xlab = "mu", main = "Gibbs draws, exact Uniform(4.2, 5.8) conditional")
abline(v = c(4.2, 5.8), col = "#b06a3b", lwd = 2, lty = 2)
```

## Key takeaways

1. `mcmc_posterior()` takes any log-likelihood you can write as an R function, over priors you
   choose, and reaches all eight ported samplers.
2. Flat priors need no term in your log-likelihood. Anything else does, and adding it is your
   job, following the upstream contract.
3. The cross-language guarantee is conditional on the callback path. It holds exactly when your
   function returns bit-identical values, so arithmetic and explicit loops hold it and calls into
   a platform math library are not guaranteed to.
4. Gibbs needs a proposal, and the proposal must draw from the handle it is given. Reaching for
   `runif()` compiles, runs, and quietly costs you reproducibility.
5. Keep callback runs small. The evaluation count is
   `(iterations + output_length / chains) * thinning * chains`, and `initialize = "MAP"` runs a
   differential evolution fit over your function before the first iteration.

## Reproduction check

The Gibbs block below is pinned against `fixtures/callback/callback_cross_language.json`, whose
values are read from the real USACE-RMC Numerics library by the oracle emitter. That fixture
asserts this exact construct at ZERO tolerance, not at a loosened one: the C++, R, Python and C#
runners reproduce it bit for bit, because both callbacks are built from arithmetic and
comparisons alone. It is checked at 1e-15 relative tolerance here only to match the convention
the other example pages use.

`output_length` is worth a word, since the call above sets it. The ported sampler computes its
posterior summaries over an OUTPUT BLOCK collected after the iteration loop, not over the
recorded chain, so the mean and the median depend on it while the chain does not. The fixture
uses 100; left at the ported Gibbs default of 10,000 the same run reports a posterior mean of
5.003896 rather than 4.984070. Every number below is therefore the fixture's own, and R and
Python return the same double for each.

The regression numbers are a within-R regression check on this page, not a cross-language claim,
for the reason given in the cross-language section above.

```{r}
near <- \(x, literal, tol = 1e-15) abs(x / literal - 1) < tol

stopifnot(
  # Gibbs, against the shared fixture oracle.
  nrow(chain) == 300, ncol(chain) == 1, length(gibbs$chains) == 1,
  near(gibbs$map[1], 5.7310633707791565),
  gibbs$map_fitness == 0,
  gibbs$acceptance_rates[1] == 0,
  near(chain[1, 1], 4.369926358014346),
  near(chain[151, 1], 4.430059107020497),
  near(chain[300, 1], 4.246628397703171),

  # Gibbs summary statistics, over the fixture's own 100-draw output block (see above).
  near(gibbs$posterior_mean[1], 4.984069999799132),
  near(gibbs$posterior_median[1], 4.943827163241803),

  # The regression posterior, within R only.
  near(fit$posterior_mean[1], -7.4538132847773335),
  near(fit$posterior_mean[2], 0.6283589815314603),
  near(fit$posterior_mean[3], 0.27803652002522816),
  near(fit$map_fitness, 12.139529268441548),
  all(fit$rhat < 1.01),

  # The MAP of the two regression parameters agrees with least squares.
  abs(fit$map[1] - ols[1]) < 0.01,
  abs(fit$map[2] - ols[2]) < 0.001
)
cat("All reproduction checks passed.\n")
```
