---
title: "09. Statistics"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Ported from
[`09_statistics.ipynb`](https://github.com/USACE-RMC/Numerics-Python-Examples/blob/main/notebooks/09_statistics.ipynb)
in the USACE-RMC Numerics-Python-Examples repository (0BSD licensed). The upstream
notebook drives the C# `Numerics.dll` through pythonnet; this version uses
`corehydror`, whose compiled core is a validated C++ port of the same library, so the
seeded results below reproduce the upstream outputs exactly. The Python version of
this example uses the same core and prints the same numbers.

## What you'll learn

- Descriptive statistics and correlation with base R.
- Empirical plotting positions, the port's public helper for frequency plots.
- Bootstrap confidence bands on a fitted quantile curve.
- Screening for low outliers with the Multiple Grubbs-Beck test.
- Box-Cox and Yeo-Johnson normalizing transformations.

## Setup

The upstream notebook loads the CoreCLR runtime, resolves `Numerics.dll`, and
converts every array with `convert_to_dotnet_array`. Here the only setup is
loading the package.

```{r}
library(corehydror)
```

## Descriptive statistics

The upstream section drives the Numerics `Statistics` class. `summary_statistics()`
exposes it, but the estimators are the same ones base R already has, so this page
stays in base R and the summary table below reproduces the upstream digit for digit:

- `Statistics.Variance` uses Bessel's correction (divide by $n-1$), like `var()`
  and `sd()`.
- `Statistics.Skewness` is the bias-corrected Fisher skewness
  $G_1 = \frac{\sqrt{n(n-1)}}{n-2} \, m_3 / m_2^{3/2}$, computed explicitly below.
- `Statistics.Kurtosis` is the adjusted Fisher *excess* kurtosis (the upstream
  table labels it "Kurtosis").
- `Statistics.Percentile` matches `quantile()` type 7, R's default.

The streamflow sample itself is seeded through the port, so it is bit-identical to
the upstream data. One naming gotcha carried over from C#: `LogNormal` is the
base-10 family (`LnNormal` is base e).

```{r}
streamflow <- dist_random(distribution("LogNormal", c(6.5, 0.8)), 365, seed = 42)
n <- length(streamflow)

central <- function(k) mean((streamflow - mean(streamflow))^k)
G1 <- sqrt(n * (n - 1)) / (n - 2) * central(3) / central(2)^1.5
g2 <- central(4) / central(2)^2 - 3
G2 <- ((n + 1) * g2 + 6) * (n - 1) / ((n - 2) * (n - 3))

summary_df <- data.frame(
  statistic = c(
    "Mean", "Std Dev", "Variance", "Minimum", "Maximum",
    "Skewness", "Excess kurtosis",
    "5th %ile", "25th %ile", "75th %ile", "95th %ile"
  ),
  value = c(
    mean(streamflow), sd(streamflow), var(streamflow),
    min(streamflow), max(streamflow), G1, G2,
    quantile(streamflow, c(0.05, 0.25, 0.75, 0.95), names = FALSE)
  )
)
print(summary_df, row.names = FALSE)
```

```{r}
hist(log10(streamflow),
  breaks = 30, col = "#6b7f3f", border = "white",
  xlab = "log10 streamflow (cfs)", ylab = "Count",
  main = "Seeded streamflow sample (n = 365)"
)
abline(v = log10(mean(streamflow)), col = "#b06a3b", lty = 2)
abline(v = log10(median(streamflow)), col = "#5b7a8c", lty = 2)
legend("topright",
  legend = c(
    sprintf("Mean: %s cfs", format(round(mean(streamflow)), big.mark = ",")),
    sprintf("Median: %s cfs", format(round(median(streamflow)), big.mark = ","))
  ),
  col = c("#b06a3b", "#5b7a8c"), lty = 2, bty = "n"
)
```

## Plotting positions

The upstream notebook builds an empirical CDF by hand for its Kolmogorov-Smirnov
diagnostic. The port exposes `plotting_positions`, the standard family of
empirical non-exceedance probabilities $(i - \alpha)/(n + 1 - 2\alpha)$ used
throughout flood frequency work. The default is Weibull ($\alpha = 0$); `median`,
`blom`, `cunnane`, `gringorten`, and `hazen` are also available, or pass `alpha =`
directly.

```{r}
pp <- plotting_positions(n)
plot(sort(streamflow), pp,
  type = "s", log = "x", col = "#6b7f3f",
  xlab = "Streamflow (cfs)", ylab = "Non-exceedance probability",
  main = "Empirical CDF (Weibull plotting positions)"
)

cat("Weibull, n = 10:", round(plotting_positions(10), 6), "\n")
cat(
  "Median (alpha = 0.3175), n = 10:",
  round(plotting_positions(10, method = "median"), 6), "\n"
)
```

## Correlation

The upstream section uses the Numerics `Correlation` class for three measures:
Pearson (linear association), Spearman (Pearson on ranks, so monotonic
association), and Kendall's $\tau_b$ (concordance, tie-adjusted). `correlation()`
exposes all three, and its matrix form as well (example 28); base R's `cor()`
computes the same coefficients on the same seeded data, so this page stays in base
R. Kendall's $\tau_b$ equals 1 here too.

Two upstream quirks worth knowing. Reusing seed 321 for both the temperature and
the noise makes the noise an exactly linear function of temperature, so the
"noisy" sales data is perfectly linear and every coefficient is exactly 1. And the
outliers the upstream inserts are added after the data has been converted to .NET,
so they never reach the correlation calls. We reproduce the coefficients as
printed.

```{r}
temperature <- dist_random(distribution("Uniform", c(60, 100)), 100, seed = 321)
noise <- dist_random(distribution("Uniform", c(0, 50)), 100, seed = 321)
sales <- 2.5 * temperature - 50 + noise

pearson_r <- cor(temperature, sales)
spearman_r <- cor(temperature, sales, method = "spearman")
cat(sprintf("Pearson correlation:  %.4f\n", pearson_r))
cat(sprintf("Spearman correlation: %.4f\n", spearman_r))

plot(temperature, sales,
  pch = 19, col = "#5b7a8c",
  xlab = "Temperature (F)", ylab = "Ice cream sales",
  main = "A perfectly linear 'noisy' relationship (shared seed)"
)
```

## Covered elsewhere: hypothesis tests, Grubbs-Beck, Jenks

The upstream notebook's middle sections use three groups of classes that have
pages of their own here:

- Hypothesis tests (one- and two-sample t, Mann-Whitney U) and the
  Kolmogorov-Smirnov statistic: `hypothesis_test()` reaches all thirteen
  `HypothesisTests` statics and `gof_test()` the `GoodnessOfFit` ones. Example 28
  works through the first, example 12 the second.
- Jenks natural breaks: `ml_jenks_breaks()`, one of the eight machine learning
  verbs in example 29.
- The single Grubbs-Beck outlier test is the one thing here that stays internal.
  Only the Multiple Grubbs-Beck generalization below is public, because that is the
  form Bulletin 17C uses.

## Bootstrap resampling

The upstream notebook bootstraps the product moments of a Gamma fit through
`BootstrapAnalysis.ProductMoments`. The port wraps the same Numerics
`BootstrapAnalysis` engine behind one public function, `bootstrap_analysis`, which
exposes the quantile-curve surface instead: fit a distribution to the data,
resample it `replications` times, re-estimate the parameters each time, and return
percentile confidence bands on the quantile curve. The Normal family is currently
the one wired to the bootstrap interface, so this recast bootstraps the
Normal(100, 15) sample the upstream notebook generates for its t-test control
group (seed 456; its mean, 96.881317, is printed upstream and asserted below).

```{r}
sample_data <- dist_random(distribution("Normal", c(100, 15)), 50, seed = 456)
probs <- c(0.01, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 0.99)
boot <- bootstrap_analysis(sample_data, "Normal", probs,
  replications = 5000L, seed = 456L, alpha = 0.05
)

cat(sprintf("Sample mean:              %.6f\n", mean(sample_data)))
cat(sprintf(
  "Fitted Normal parameters: %.6f %.6f\n",
  boot$parameters[1], boot$parameters[2]
))
curve_df <- data.frame(
  non_exceedance = probs,
  quantile_mle_fit = boot$mode_curve,
  lower_95 = boot$lower_ci,
  upper_95 = boot$upper_ci
)
print(curve_df, row.names = FALSE)
```

```{r}
plot(range(probs), range(c(boot$lower_ci, boot$upper_ci)),
  type = "n",
  xlab = "Non-exceedance probability", ylab = "Value",
  main = "Parametric bootstrap of a Normal fit (5000 replications)"
)
polygon(c(probs, rev(probs)), c(boot$lower_ci, rev(boot$upper_ci)),
  col = adjustcolor("#8c8c7a", alpha.f = 0.35), border = NA
)
lines(probs, boot$mode_curve, col = "#6b7f3f", lwd = 2)
points(plotting_positions(length(sample_data)), sort(sample_data),
  pch = 19, cex = 0.6, col = "#5b7a8c"
)
legend("topleft",
  legend = c("95% bootstrap band", "Fitted quantile curve", "Sample (Weibull positions)"),
  fill = c(adjustcolor("#8c8c7a", alpha.f = 0.35), NA, NA),
  border = NA, lty = c(NA, 1, NA), pch = c(NA, NA, 19),
  col = c(NA, "#6b7f3f", "#5b7a8c"), bty = "n"
)
```

## Outlier detection: Multiple Grubbs-Beck

The classic Grubbs-Beck test flags outliers under a log-normality assumption using
a critical value that depends on sample size. The Multiple Grubbs-Beck test (MGBT)
generalizes it to detect several potentially influential low floods (PILFs) at
once, which matters in Bulletin 17C flood frequency analysis because low outliers
can distort the fitted Log-Pearson Type III curve and bias upper-tail quantiles.
`mgbt_test` returns the number of low outliers detected. Note that MGBT is a
low-outlier screen only: the high value of 200 planted in this sample is not
flagged.

```{r}
clean <- dist_random(distribution("Normal", c(100, 10)), 47, seed = 789)
values <- c(clean, 35, 30, 200)
n_low <- mgbt_test(values)

mgbt_df <- data.frame(
  metric = c("Sample size", "Mean", "Std Dev", "Low outliers detected (MGBT)"),
  value = c(length(values), mean(values), sd(values), n_low)
)
print(mgbt_df, row.names = FALSE)
```

```{r}
low_idx <- order(values)[seq_len(n_low)]
plot(seq_along(values), values,
  pch = 19, col = "#5b7a8c",
  xlab = "Observation index", ylab = "Value",
  main = "Data with MGBT-detected low outliers"
)
points(low_idx, values[low_idx], pch = 19, col = "#b06a3b")
abline(h = mean(values), col = "#8c8c7a", lty = 2)
legend("topleft",
  legend = c(
    "Data", sprintf("MGBT low outliers: %d", n_low),
    sprintf("Mean = %.2f", mean(values))
  ),
  col = c("#5b7a8c", "#b06a3b", "#8c8c7a"),
  pch = c(19, 19, NA), lty = c(NA, NA, 2), bty = "n"
)
```

## Data transformations: Box-Cox and Yeo-Johnson

Both transformations reshape a skewed variable toward normality. Box-Cox requires
strictly positive data and finds a power $\lambda$ that stabilizes variance and
reduces skew; Yeo-Johnson extends the idea to zero and negative values. The port
fits $\lambda$ by maximum likelihood over $[-5, 5]$ (`box_cox_lambda`,
`yeo_johnson_lambda`) and provides the forward and inverse transforms.

One honest floating-point note: the fitted exponent comes from a numerical search
(Brent), and its trailing digits depend on platform and compiler, so R and Python
agree to about seven decimals here rather than bit-exactly. The reproduction check
therefore asserts the precision the upstream notebook prints.

```{r}
bc_data <- dist_random(distribution("LogNormal", c(1, 0.75)), 500, seed = 789)
lambda_bc <- box_cox_lambda(bc_data)
bc_t <- box_cox(bc_data, lambda_bc)
bc_err <- max(abs(bc_data - box_cox_inverse(bc_t, lambda_bc)))

yj_data <- dist_random(distribution("Normal", c(0, 2)), 500, seed = 456) +
  dist_random(distribution("Exponential", c(1, 1.5)), 500, seed = 456)
lambda_yj <- yeo_johnson_lambda(yj_data)
yj_t <- yeo_johnson(yj_data, lambda_yj)
yj_err <- max(abs(yj_data - yeo_johnson_inverse(yj_t, lambda_yj)))

# Std uses the population formula (n divisor) to match the upstream np.std table.
sd_pop <- function(x) sqrt(mean((x - mean(x))^2))
transform_df <- data.frame(
  transform = c(
    "Box-Cox (original)", "Box-Cox (transformed)",
    "Yeo-Johnson (original)", "Yeo-Johnson (transformed)"
  ),
  mean = c(mean(bc_data), mean(bc_t), mean(yj_data), mean(yj_t)),
  std = c(sd_pop(bc_data), sd_pop(bc_t), sd_pop(yj_data), sd_pop(yj_t)),
  min = c(min(bc_data), min(bc_t), min(yj_data), min(yj_t)),
  max = c(max(bc_data), max(bc_t), max(yj_data), max(yj_t)),
  lambda = c(lambda_bc, lambda_bc, lambda_yj, lambda_yj),
  round_trip_error = c(bc_err, bc_err, yj_err, yj_err)
)
print(transform_df, row.names = FALSE)
```

```{r}
op <- par(mfrow = c(2, 2))
hist(bc_data,
  breaks = 30, col = "#6b7f3f", border = "white",
  xlab = "Value", main = "Box-Cox: original"
)
hist(bc_t,
  breaks = 30, col = "#b06a3b", border = "white",
  xlab = "Value", main = sprintf("Box-Cox: transformed (lambda = %.3f)", lambda_bc)
)
hist(yj_data,
  breaks = 30, col = "#6b7f3f", border = "white",
  xlab = "Value", main = "Yeo-Johnson: original"
)
hist(yj_t,
  breaks = 30, col = "#b06a3b", border = "white",
  xlab = "Value", main = sprintf("Yeo-Johnson: transformed (lambda = %.3f)", lambda_yj)
)
par(op)
```

## Key takeaways

1. Descriptive statistics and correlation come both ways: `summary_statistics()`
   and `correlation()` in the package, or base R and pandas, which agree with them.
2. Bootstrap for uncertainty when you do not want strong assumptions; the port's
   surface is confidence bands on the quantile curve.
3. Screen for low outliers with MGBT before fitting a flood frequency curve, and
   investigate before removing anything.
4. Normalize skewed positive data with Box-Cox; reach for Yeo-Johnson when zeros
   or negatives appear.

## Reproduction check

Values printed by the upstream notebook (run against the real C# library),
compared with this port. "exact" means the value matches every digit the upstream
notebook prints; the two fitted exponents are asserted at that printed precision
because a numerical search sets their trailing digits (see the transforms
section). "internal" rows have no upstream number, so they assert internal
consistency plus cross-language identity against the Python twin.

| Quantity | Upstream C# | This port | Status |
|----------|------------:|----------:|--------|
| Streamflow mean (`LogNormal(6.5,0.8)`, n=365, seed 42) | 1.397387e+07 | `mean(streamflow)` | exact |
| Streamflow std dev | 4.195469e+07 | `sd(streamflow)` | exact |
| Streamflow skewness $G_1$ | 8.697790 | `G1` | exact |
| Streamflow excess kurtosis | 103.8059 | `G2` | exact |
| Streamflow 95th percentile | 5.803155e+07 | `quantile(., 0.95)` | exact |
| Pearson / Spearman correlation | 1.0000 | `cor()` | exact |
| Control sample mean (`Normal(100,15)`, n=50, seed 456) | 96.881317 | `mean(sample_data)` | exact |
| MGBT low outliers | 2 | `mgbt_test` | exact |
| MGBT sample mean / std dev | 98.159548 / 21.844798 | base R | exact |
| Box-Cox lambda | 0.020475 | `box_cox_lambda` | exact |
| Box-Cox transformed mean | 2.380719 | `mean(bc_t)` | exact |
| Yeo-Johnson lambda | 0.709817 | `yeo_johnson_lambda` | exact |
| Yeo-Johnson transformed mean | 1.645339 | `mean(yj_t)` | exact |
| Transform round trips | (not printed) | max abs error < 1e-9 | internal |
| Bootstrap quantile bands | (no upstream analogue) | ordering + literals | internal |
| Plotting positions | (not in upstream 09) | closed form | internal |

The chunk below fails the render if any value drifts. The final block asserts the
cross-language guarantee: the seeded values above are the exact values the Python
notebook prints.

```{r}
# Upstream: 09_statistics.ipynb, cell 4 output (descriptive summary table).
stopifnot(
  abs(mean(streamflow) / 1.397387e7 - 1) < 1e-6,
  abs(sd(streamflow) / 4.195469e7 - 1) < 1e-6,
  abs(G1 - 8.697790) < 5e-7,
  abs(G2 / 103.8059 - 1) < 1e-6,
  abs(quantile(streamflow, 0.95, names = FALSE) / 5.803155e7 - 1) < 1e-6
)

# Upstream: cell 6 output (all three coefficients print 1.0000).
stopifnot(pearson_r > 1 - 1e-12, spearman_r > 1 - 1e-12)

# Upstream: cell 8 output (t-test control group mean, reused for the bootstrap).
stopifnot(abs(mean(sample_data) - 96.881317) < 5e-7)

# Upstream: cell 17 output (Multiple Grubbs-Beck summary).
stopifnot(
  n_low == 2,
  abs(mean(values) - 98.159548) < 5e-7,
  abs(sd(values) - 21.844798) < 5e-7
)

# Upstream: cell 21 output (transformation summary table).
stopifnot(
  abs(lambda_bc - 0.020475) < 5e-7,
  abs(mean(bc_t) - 2.380719) < 5e-7,
  abs(lambda_yj - 0.709817) < 5e-7,
  abs(mean(yj_t) - 1.645339) < 5e-7
)

# Internal consistency: transforms invert, plotting positions match closed form,
# bootstrap bands bracket the fitted curve, and the fitted median of a Normal is
# its mean parameter.
stopifnot(
  bc_err < 1e-9, yj_err < 1e-9,
  all(abs(plotting_positions(10) - (1:10) / 11) < 1e-15),
  all(boot$lower_ci < boot$mode_curve), all(boot$mode_curve < boot$upper_ci),
  boot$mode_curve[7] == boot$parameters[1] # probs[7] == 0.5
)

# Cross-language identity: the Python notebook asserts these same seeded
# literals. The values are bit-exact across languages; the comparisons carry a
# 1e-15 relative tolerance only because R's decimal parser can land one ulp away
# from the written literal.
stopifnot(
  abs(streamflow[1] / 1754357.4518502082 - 1) < 1e-15,
  abs(boot$parameters[1] / 96.88093513274926 - 1) < 1e-15,
  abs(boot$lower_ci[7] / 93.19634521338513 - 1) < 1e-15,
  abs(boot$upper_ci[7] / 100.52462982500435 - 1) < 1e-15,
  abs(mean(values) / 98.15954794981819 - 1) < 1e-15,
  abs(plotting_positions(10, method = "median")[1] / 0.06584659913169319 - 1) < 1e-15
)
cat("All reproduction checks passed.\n")
```
