---
title: "30. Time series"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

The Numerics `TimeSeries` container, which had no R or Python binding until this release. It is
the thing a flood-frequency analysis actually starts from: a gauge record with gaps in it, out of
which you have to get an annual maximum series, or a set of independent peaks, before any of the
fitting machinery is any use.

This page walks one thirty-year daily record from raw gauge output to two fitted flood-frequency
models, and then generates synthetic records from it.

## What you'll learn

- How to build a `time_series()`, and why the interval it carries is not just a label.
- How to repair a record: filling dates that are missing entirely, and interpolating short runs of
  missing values -- and what the library does when a run is too long.
- How to reduce a record to annual maxima over a water year, and what happens to a year that
  contains a missing value (it depends on the block function, and the reason is worth knowing).
- How to extract independent peaks over a threshold, and what the independence criterion is
  measured in.
- How to decompose a series into trend, seasonal and residual components.
- How to generate synthetic records with two resamplers, and why a seeded run gives identical
  numbers in R and Python.

## Setup

```{r}
library(corehydror)
```

## A thirty-year daily record

The record below is illustrative rather than observed, but it is built to be exactly reproducible:
a twelve-value monthly base flow, modulated by a small integer linear congruential generator
written out in full, so R and Python compute byte-identical inputs without either of them calling
a random number generator or a transcendental function.

```{r}
# A deterministic uniform sequence in [0, 1). The arithmetic stays well inside a double's exact
# integer range, so this is the same sequence in every language.
lcg <- function(n, seed = 7) {
  s <- seed
  out <- numeric(n)
  for (i in seq_len(n)) {
    s <- (75 * s + 74) %% 65537
    out[i] <- s / 65537
  }
  out
}

dates <- seq(as.Date("1990-01-01"), as.Date("2019-12-31"), by = "day")
n <- length(dates)
u <- lcg(n)

# Monthly base flow (m^3^/s), a snowmelt-dominated regime peaking in spring.
base <- c(18, 22, 35, 48, 40, 26, 16, 12, 11, 14, 20, 19)
month <- as.integer(format(dates, "%m"))
flow <- round(base[month] * (0.6 + 0.8 * u), 3)

# Storms: the top 1.5% of the modulating sequence become peaks, each scaled by a second
# deterministic sequence so the peaks have a spread rather than all landing on one multiple.
storm <- u > 0.985
flow[storm] <- round(flow[storm] * (2.5 + 7 * lcg(n, seed = 913)[storm]), 3)

length(flow)
range(flow)
```

Two kinds of damage, both of which a real record has:

```{r}
# A short instrument outage: five consecutive missing days in 2002.
outage <- which(dates >= as.Date("2002-06-10") & dates <= as.Date("2002-06-14"))
flow[outage] <- NA

# Three days that never made it into the file at all.
dropped <- which(dates %in% as.Date(c("2004-03-02", "2004-03-03", "2005-11-19")))
gauge <- time_series(dates[-dropped], flow[-dropped], interval = "one_day")
gauge
```

The `interval` is not decoration. It is what `ts_shift(by = "start")` walks, what
`ts_convert_interval()` converts from, and -- the one that matters below -- what
`ts_peaks_over_threshold()` measures its independence criterion in. A "minimum of 5 steps between
events" means five days here and would mean five months on a monthly series.

## Repairing the record

Missing DATES and missing VALUES are different problems and take different verbs.
`ts_fill_missing_dates()` inserts the ordinates that are absent entirely;
`ts_interpolate_missing()` fills short runs of values that are present but unobserved.

```{r}
filled <- ts_fill_missing_dates(gauge, min(gauge$dates), max(gauge$dates))
c(before = length(gauge), after = length(filled))
```

Now every day is present and the three dropped ones are missing values, alongside the five-day
outage. Interpolating with a limit of 3 fills the dropped days and leaves the outage alone -- the
limit is the point, not an inconvenience: a five-day gap in a flashy record is not something linear
interpolation should be inventing values for.

```{r}
repaired <- ts_interpolate_missing(filled, max_missing = 3)
data.frame(
  gap = c("dropped days (2)", "dropped day (1)", "outage (5)"),
  filled = c(
    sum(!is.na(repaired$values[repaired$dates >= as.POSIXct("2004-03-02", tz = "UTC") &
                                 repaired$dates <= as.POSIXct("2004-03-03", tz = "UTC")])),
    sum(!is.na(repaired$values[repaired$dates == as.POSIXct("2005-11-19", tz = "UTC")])),
    sum(!is.na(repaired$values[repaired$dates >= as.POSIXct("2002-06-10", tz = "UTC") &
                                 repaired$dates <= as.POSIXct("2002-06-14", tz = "UTC")]))
  )
)
```

## What is in the record

```{r}
stats <- ts_statistics(repaired)
round(stats[c("Record Length", "Missing Values", "Minimum", "Maximum", "Mean", "Std Dev",
              "Skewness", "50%", "95%")], 3)
```

Per-month statistics show the regime directly. Note that this verb filters missing values out,
while `ts_monthly_percentiles()` does not -- a difference inherited from the library, and the
reason June's percentiles below are missing while its summary row is not.

```{r}
round(ts_monthly_statistics(repaired)[, c("minimum", "p50", "mean", "maximum")], 2)
```

```{r}
round(ts_monthly_percentiles(repaired, c(0.5, 0.95))[5:7, ], 2)
```

The duration curve is the same information as a flow-exceedance plot:

```{r}
dur <- ts_duration(repaired)
round(dur[c(1, 30, nrow(dur) %/% 2, nrow(dur)), ], 3)
```

## Annual maxima

A flood-frequency analysis wants one value per year, and in most of the world it wants the year to
start in autumn so a winter flood season is not split across two of them. That is
`ts_water_year()`:

```{r}
ams <- ts_water_year(repaired, block = "maximum", start_month = 10)
head(data.frame(date = format(ams$dates, "%Y-%m-%d"), peak = ams$values), 8)
```

Two things to notice. The result is an IRREGULAR series carrying each maximum's own date, not the
end of its year -- so you can see when the annual flood happened, which is a seasonality signal in
itself. And the record spans thirty-one water years, not thirty: it starts in January 1990,
part-way through water year 1990, and ends in December 2019, part-way through water year 2020.
Both partial years contribute their own maximum.

Compare it with a calendar-year reduction:

```{r}
list(
  water_year = round(ts_water_year(repaired)$values, 1),
  calendar_year = round(ts_calendar_year(repaired)$values, 1)
)
```

They differ in both length and content: a flood in October or later belongs to the NEXT water year
but the same calendar year.

One behaviour is worth stating outright, because it decides whether a gappy record silently loses a
year. A block containing a missing value is DROPPED by the sum and average block functions -- the
missing value propagates through the arithmetic -- but KEPT by minimum and maximum, because the
comparison is simply false against a missing value and it is never selected. This is the library's
behaviour, and it is why the maximum series above spans every year despite the outage:

```{r}
c(
  maximum_years = length(ts_water_year(filled, block = "maximum")),
  sum_years = length(ts_water_year(filled, block = "sum"))
)
```

Fitting is one call away:

```{r}
fit <- fit_mle(model_univariate("GeneralizedExtremeValue", ams$values))
round(coef(fit), 4)
round(dist_quantile(distribution("GeneralizedExtremeValue", coef(fit)), c(0.5, 0.9, 0.99)), 2)
```

## Peaks over a threshold

Annual maxima throw away most of a record: eight numbers out of nearly three thousand. A
partial-duration series keeps every independent peak above a threshold instead, which is why the
independence criterion matters -- without it, one flood's rising and falling limbs count as many
events.

```{r}
pot_1 <- ts_peaks_over_threshold(repaired, threshold = 120, min_steps_between_events = 1)
pot_10 <- ts_peaks_over_threshold(repaired, threshold = 120, min_steps_between_events = 10)
c(one_day_apart = length(pot_1), ten_days_apart = length(pot_10))
```

```{r}
head(data.frame(date = format(pot_10$dates, "%Y-%m-%d"), peak = round(pot_10$values, 2)), 8)
```

`min_steps_between_events` is counted in the series' own interval, so on this daily record ten
steps is ten days. The extracted series feeds a point-process model directly:

```{r}
pp <- model_point_process(pot_10$values, threshold = 120, total_years = 30)
round(coef(fit_mle(pp)), 4)
```

Smoothing happens BEFORE the extraction, which is how an n-day event series is built: the peaks of
a 3-day moving average are the largest 3-day volumes, not the largest instantaneous flows.

```{r}
pot_3day <- ts_peaks_over_threshold(repaired, threshold = 120,
                                    min_steps_between_events = 10,
                                    smoothing = "moving_average", period = 3)
c(instantaneous = length(pot_10), three_day_mean = length(pot_3day))
```

Far fewer events clear the same threshold, which is the expected result rather than a surprise: a
three-day mean of a flashy record is a smaller number than its instantaneous peak. Choose the
threshold for the series you are actually extracting from.

## Decomposition

`ts_seasonal_decompose()` splits the record into trend, seasonal and residual parts, keeping only
the harmonics of the seasonal frequency for the seasonal component. On a monthly aggregation of
this record the annual cycle is the whole story:

```{r}
monthly <- ts_block_series(repaired, window = "month", block = "average")
monthly <- time_series(monthly$dates, monthly$values, interval = "one_month")
d <- ts_seasonal_decompose(monthly, period = 12)
defined_rows <- head(which(!is.na(d$trend)), 4)
data.frame(
  date = format(d$date[defined_rows], "%Y-%m"),
  round(d[defined_rows, c("trend", "seasonal", "residual")], 3)
)
```

The trend is a moving average over `period` ordinates, so it -- and the residual with it -- is
undefined for the first eleven months. Where all three components exist they add back to the
original value exactly, which the reproduction check at the bottom asserts.

## Synthetic records

Two resamplers generate records of arbitrary length from the observed one.
`ts_resample_knn()` is the conditional k-nearest-neighbour bootstrap of Lall and Sharma (1996): at
each step it finds the `k` historical days closest to the current flow, picks one, and advances to
whatever historically came NEXT. That last step is what preserves the recession behaviour -- taking
the neighbour's own value instead would leave the trajectory hovering near where it started.

```{r}
observed <- ts_replace_missing(repaired, value = 0)
knn <- ts_resample_knn(observed, time_steps = 365, k = 20, seed = 2024)
block <- ts_resample_block_bootstrap(observed, time_steps = 365, block_size = 30, seed = 2024)

data.frame(
  series = c("observed", "knn", "block bootstrap"),
  mean = round(c(mean(observed$values), mean(knn$values), mean(block$values)), 3),
  sd = round(c(sd(observed$values), sd(knn$values), sd(block$values)), 3),
  lag1 = round(c(
    autocorrelation(observed, max_lag = 1)$value[2],
    autocorrelation(knn, max_lag = 1)$value[2],
    autocorrelation(block, max_lag = 1)$value[2]
  ), 3)
)
```

Both preserve the marginal distribution to within sampling error on a 365-day draw, and both
recover most of the lag-1 dependence. Neither recovers all of it, and it is worth being clear about
why rather than reading the table as a ranking: this record's day-to-day persistence comes almost
entirely from its seasonal cycle rather than from a recession, since the modulating sequence is
independent from day to day. A resampler that reorders time cannot keep a seasonal signal it was
never conditioned on. On a real record with genuine recession behaviour the kNN resample is the one
that holds lag-1 dependence, because it conditions each step on the current state; the block
bootstrap keeps dependence only WITHIN a block and breaks it at every boundary.

Seeded runs are reproducible, and the reproduction is not only within one language: the whole
computation, the Mersenne Twister included, happens in the shared C++ core, so these 365 numbers
are bit-identical to the ones the Python page prints.

```{r}
round(knn$values[1:8], 3)
```

## Reproduction check

Every literal below was produced by this page. The Python page asserts the same values.

```{r}
near <- function(a, b, tol = 1e-15) isTRUE(all.equal(a, b, tolerance = tol))
defined <- !is.na(d$trend)

stopifnot(
  # The record and its repair: three dates absent entirely, five values missing, and
  # interpolation fills the short gaps only.
  length(gauge) == 10954L,
  length(filled) == 10957L,
  sum(is.na(filled$values)) == 8L,
  sum(is.na(repaired$values)) == 5L,

  # Summary statistics.
  near(unname(stats["Record Length"]), 10957),
  near(unname(stats["Missing Values"]), 5),
  near(unname(stats["Mean"]), 26.130572315558798),
  near(unname(stats["Maximum"]), 588.07600000000002),

  # Annual maxima: thirty-one water years against thirty calendar years.
  length(ams) == 31L,
  length(ts_calendar_year(repaired)) == 30L,
  near(ams$values[1], 288.59899999999999),
  near(max(ams$values), 588.07600000000002),

  # A missing value costs the sum series three years but the maximum series none.
  length(ts_water_year(filled, block = "maximum")) == 31L,
  length(ts_water_year(filled, block = "sum")) == 28L,

  # The fitted GEV.
  near(unname(coef(fit)[1]), 295.35281082558026),
  near(unname(coef(fit)[2]), 116.76657650129897),
  near(unname(coef(fit)[3]), 0.25436343374399251),

  # Peaks over threshold: the independence criterion declusters, and smoothing before the
  # extraction changes what clears the threshold at all.
  length(pot_1) == 124L,
  length(pot_10) == 106L,
  length(pot_3day) == 30L,
  near(pot_10$values[1], 288.59899999999999),
  near(unname(coef(fit_mle(pp))[1]), 295.7606336759473),

  # The decomposition is additive wherever the trend is defined.
  length(monthly) == 359L,
  sum(defined) == 348L,
  all(abs(d$trend[defined] + d$seasonal[defined] + d$residual[defined] -
            monthly$values[defined]) < 1e-9),

  # The seeded resamplers. These are the numbers the Python page prints too: the generator and
  # every operation around it live in the shared C++ core.
  length(knn) == 365L,
  length(block) == 365L,
  near(knn$values[1], 34.271999999999998),
  near(knn$values[8], 9.2219999999999995),
  near(block$values[1], 23.332999999999998)
)
cat("All reproduction checks passed.\n")
```
