---
title: "05. Adaptive MCMC samplers"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Ported from
[`05_mcmc_adaptive.ipynb`](https://github.com/USACE-RMC/Numerics-Python-Examples/blob/main/notebooks/05_mcmc_adaptive.ipynb)
in the USACE-RMC Numerics-Python-Examples repository (0BSD licensed). The upstream
notebook drives the C# `Numerics.dll` through pythonnet; this version uses
`corehydror`, whose compiled core is a validated C++ port of the same library. The
seeded data and every posterior summary printed below are bit-identical to the
[Python version](python.ipynb) of this example, which runs the same core stream.
Comparisons against the upstream notebook's printed posteriors are statistical
rather than exact because the priors differ (see the reproduction check at the
end).

## What you'll learn

- Why adaptive MCMC improves mixing without manual proposal tuning.
- How ARWMH learns the proposal covariance during warmup.
- How DEMCz and DEMCzs build proposals from a population of past states.
- How HMC and NUTS use gradient information to explore efficiently.
- How to compare samplers using acceptance rate, R-hat, and effective sample size.

## Setup

The upstream notebook loads the CoreCLR runtime through pythonnet, resolves
`Numerics.dll`, wraps a log-likelihood function in a .NET `LogLikelihood`
delegate, builds a `List[IUnivariateDistribution]` of priors, and disables the
samplers' internal chain parallelism to avoid GIL contention. None of that exists
here. `mcmc_sample()` fits a named distribution family to the data and returns
the chains, acceptance rates, posterior summaries, and diagnostics in one call.

One honest difference to keep in mind throughout: the upstream notebook
hand-picks its priors (`Uniform(50, 150)` for $\mu$ and `Uniform(5, 30)` for
$\sigma$ in the Normal example). `mcmc_sample()` always uses uniform priors spanning
the family's parameter constraints, exactly the C# `GetParameterConstraints` bounds,
which is what keeps it a one-liner. (`mcmc_posterior()` takes priors you choose, over
a log-likelihood you write; [example 14](../14-custom-posterior/r.qmd) is that page.)
With data this informative the posteriors are nearly the same, but bit-comparison
against the upstream outputs is impossible, so the reproduction check treats every
upstream posterior value as a statistical comparison.

```{r}
library(corehydror)
```

## Adaptive Random Walk Metropolis-Hastings (ARWMH)

We start with the most basic adaptive sampler: Adaptive Random Walk
Metropolis-Hastings, which tunes the proposal distribution automatically during
warmup [[1]](#references).

### Mathematical foundation

The proposal at iteration $t$ uses a mixture:

$$
\theta^* \sim \begin{cases} \mathcal{N}\!\left(\theta, \, \frac{0.1^2}{d} \, I_d\right) & \text{with probability } \beta \\ \mathcal{N}\!\left(\theta, \, \frac{2.38^2}{d} \, \hat{\Sigma}_t\right) & \text{with probability } 1-\beta \end{cases}
$$

where $d$ is the number of parameters, $\beta = 0.05$ by default,
$\hat{\Sigma}_t$ is the running covariance of the chain's accepted samples, and
$I_d$ is the identity matrix. The small identity component, also used for the
first $100 \times d$ samples, keeps the chain ergodic even when the adaptive
covariance estimate is poor. Both proposal components are symmetric, so the
acceptance criterion reduces to the plain posterior ratio, exactly as in RWMH.

When to use ARWMH:

- Medium-dimensional problems (2 to 20 parameters).
- Correlated parameters.
- When you do not want to tune proposals by hand. It is a sound default for most
  applications.

We generate 100 seeded draws from a Normal(100, 15), then recover the
parameters. The seed reproduces the C# Mersenne Twister stream bit-for-bit, so
`data` matches the upstream notebook's synthetic data and the Python twin
exactly.

```{r}
true_mu <- 100
true_sigma <- 15
data <- dist_random(distribution("Normal", c(true_mu, true_sigma)), 100, seed = 1234)
cat(sprintf("first draw: %.14f, sample mean: %.3f\n", data[1], mean(data)))

arwmh <- mcmc_sample(
  data, distribution = "Normal", sampler = "ARWMH",
  iterations = 2500, warmup = 800, chains = 4, thinning = 1, seed = 12345
)

summary_df <- data.frame(
  parameter = arwmh$parameters,
  true = c(true_mu, true_sigma),
  posterior_mean = arwmh$posterior_mean,
  posterior_sd = arwmh$posterior_sd,
  rhat = arwmh$rhat,
  ess = arwmh$ess
)
cat("ARWMH results\n")
print(summary_df, row.names = FALSE, digits = 5)
cat("\nacceptance rate by chain:", round(arwmh$acceptance_rates, 3), "\n")
```

The acceptance rates sit near the 0.3 to 0.4 range that is close to optimal for
random-walk samplers in low dimensions, with no manual tuning. Trace plots
confirm the chains mix well around the posterior mean:

```{r}
mu_draws <- unlist(lapply(arwmh$chains, function(c) c[, 1]))
sigma_draws <- unlist(lapply(arwmh$chains, function(c) c[, 2]))

op <- par(mfrow = c(2, 1), mar = c(4, 4, 2, 1))
plot(mu_draws, type = "l", lwd = 0.4, col = "#5b7a8c",
     xlab = "", ylab = "mu", main = "ARWMH trace, Normal fit")
abline(h = mean(mu_draws), col = "#b06a3b", lty = 2, lwd = 1.5)
plot(sigma_draws, type = "l", lwd = 0.4, col = "#5b7a8c",
     xlab = "draw (chains concatenated)", ylab = "sigma")
abline(h = mean(sigma_draws), col = "#b06a3b", lty = 2, lwd = 1.5)
par(op)
```

## Differential Evolution MCMC (DEMCz and DEMCzs)

Differential Evolution MCMC (DEMCz) is a population-based sampler
[[2]](#references). DEMCz with snooker update (DEMCzs) is an enhanced version
with better mixing.

### Mathematical foundation

DEMCz generates proposals from the difference of two past states drawn from a
population matrix $Z$, a memory of past states from all chains:

$$
\theta^*_i = \theta_i + \gamma \, (z_{R_1} - z_{R_2}) + e
$$

where $\gamma = 2.38 / \sqrt{2d}$ is the default jump rate, $z_{R_1}$ and
$z_{R_2}$ are two randomly selected past states, and
$e \sim \mathcal{N}(0, b^2)$ is a small noise perturbation with $b = 10^{-3}$.
With probability 0.1 the jump rate is set to $\gamma = 1$, which jumps the full
difference between two past states and lets the chain bridge separated modes.
The population matrix learns the scale and orientation of the posterior on its
own, so no proposal covariance needs to be specified. The snooker update in
DEMCzs adds proposals along the line through the current state and a random past
state, which further improves mixing in awkward geometries.

When to use DEMCz or DEMCzs:

- High-dimensional problems (20+ parameters).
- Multimodal posteriors or complex posterior geometry.
- When ARWMH struggles with convergence.

The upstream notebook's advice holds here too: **DEMCzs is the fastest and most
robust of these samplers, and the best adaptive method to default to.** It is
also the default sampler used by the `*_analysis` functions elsewhere in this
package.

```{r}
demczs_norm <- mcmc_sample(
  data, distribution = "Normal", sampler = "DEMCzs",
  iterations = 2500, warmup = 800, chains = 4, thinning = 1, seed = 12345
)

summary_df <- data.frame(
  parameter = demczs_norm$parameters,
  true = c(true_mu, true_sigma),
  posterior_mean = demczs_norm$posterior_mean,
  posterior_sd = demczs_norm$posterior_sd,
  rhat = demczs_norm$rhat,
  ess = demczs_norm$ess
)
cat("DEMCzs results\n")
print(summary_df, row.names = FALSE, digits = 5)
cat("\nacceptance rate by chain:", round(demczs_norm$acceptance_rates, 3), "\n")
```

## Comparing MCMC samplers

Different samplers have different strengths. We compare all five adaptive
samplers the port exposes (plain RWMH and SNIS are also available through the
same interface):

- ARWMH: adaptive random walk (auto-tuned covariance).
- DEMCz: differential evolution (population-based proposals).
- DEMCzs: DEMCz with snooker update (better mixing).
- HMC: Hamiltonian Monte Carlo (gradient-based).
- NUTS: No-U-Turn Sampler (adaptive HMC).

### HMC mathematical foundation

Hamiltonian Monte Carlo augments the parameter space with momentum variables
$\phi$ and simulates Hamiltonian dynamics to generate distant, high-quality
proposals [[3]](#references)[[4]](#references):

$$
H(\theta, \phi) = U(\theta) + K(\phi) = -\log \pi(\theta) + \frac{1}{2} \phi^T M^{-1} \phi
$$

where $U(\theta)$ is the potential energy (negative log-posterior), $K(\phi)$ is
the kinetic energy, and the mass matrix $M$ is diagonal in this implementation.
HMC is very efficient on smooth posteriors (low autocorrelation, fast
exploration) but needs gradients, is less robust to discontinuities, and its
trajectory length must be tuned by hand.

### NUTS mathematical foundation

The No-U-Turn Sampler removes HMC's most sensitive tuning parameter, the number
of leapfrog steps, by growing the trajectory until it starts to double back on
itself [[5]](#references). It builds a balanced binary tree of leapfrog states
and stops when the U-turn criterion triggers:

$$
(\theta^+ - \theta^-) \cdot \phi^- < 0 \quad \text{or} \quad (\theta^+ - \theta^-) \cdot \phi^+ < 0
$$

where $\theta^{\pm}$ are the trajectory endpoints and $\phi^{\pm}$ their
momenta. Step size is adapted automatically by dual averaging, which makes NUTS
the recommended gradient-based sampler for most problems.

A compact guide:

| Sampler | Best for | Notes |
|---------|----------|-------|
| ARWMH | 2-20 parameters, correlated parameters | Sound default, no tuning |
| DEMCz | High dimensions, multimodal posteriors | Population supplies the proposal scale |
| DEMCzs | Same as DEMCz | Fastest and most robust; default to this |
| HMC | Smooth, differentiable posteriors | Needs step size and step count tuning |
| NUTS | Smooth posteriors, no tuning appetite | Self-tuning HMC, costlier per draw |

### Gumbel example

We fit a Gumbel distribution to seeded synthetic data with all five samplers.
The Gumbel is a good test case for the gradient-based samplers because its
log-density is smooth and cheap to differentiate numerically, so HMC and NUTS
avoid the step-size thrashing that heavier-tailed families can induce. HMC and
NUTS get smaller iteration budgets because each of their iterations costs many
gradient evaluations.

```{r}
true_xi <- 10.0    # location
true_alpha <- 2.0  # scale
gumbel_data <- dist_random(distribution("Gumbel", c(true_xi, true_alpha)), 50, seed = 12345)
cat(sprintf("first draw: %.15f, sample mean: %.3f\n", gumbel_data[1], mean(gumbel_data)))

settings <- list(
  ARWMH  = list(iterations = 2500, warmup = 800, chains = 4),
  DEMCz  = list(iterations = 2500, warmup = 800, chains = 4),
  DEMCzs = list(iterations = 2500, warmup = 800, chains = 4),
  HMC    = list(iterations = 2000, warmup = 600, chains = 2),
  NUTS   = list(iterations = 1200, warmup = 400, chains = 2)
)

runs <- list()
rows <- list()
for (name in names(settings)) {
  cfg <- settings[[name]]
  t0 <- Sys.time()
  r <- mcmc_sample(
    gumbel_data, distribution = "Gumbel", sampler = name,
    iterations = cfg$iterations, warmup = cfg$warmup, chains = cfg$chains,
    thinning = 1, seed = 12345
  )
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  runs[[name]] <- r
  rows[[name]] <- data.frame(
    sampler = name,
    runtime_s = round(elapsed, 2),
    xi_mean = r$posterior_mean[1],
    alpha_mean = r$posterior_mean[2],
    xi_error = r$posterior_mean[1] - true_xi,
    alpha_error = r$posterior_mean[2] - true_alpha,
    accept = mean(r$acceptance_rates),
    max_rhat = max(r$rhat),
    min_ess = min(r$ess)
  )
}

comparison <- do.call(rbind, rows)
cat("Sampler comparison\n")
print(comparison, row.names = FALSE, digits = 4)
```

All five samplers agree on the posterior to two decimal places, matching the
upstream notebook's picture: the random-walk and population samplers finish in
milliseconds, while the gradient-based samplers pay for numerical
differentiation on every leapfrog step but return chains with very low
autocorrelation (note the ESS relative to the number of draws). The posterior
from DEMCzs:

```{r}
best <- runs[["DEMCzs"]]
xi_draws <- unlist(lapply(best$chains, function(c) c[, 1]))
alpha_draws <- unlist(lapply(best$chains, function(c) c[, 2]))

op <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
hist(xi_draws, breaks = 40, freq = FALSE, col = "#6b7f3f", border = "white",
     xlab = "xi (location)", main = "DEMCzs posterior: xi")
abline(v = true_xi, col = "#b06a3b", lty = 2, lwd = 1.5)
hist(alpha_draws, breaks = 40, freq = FALSE, col = "#6b7f3f", border = "white",
     xlab = "alpha (scale)", main = "DEMCzs posterior: alpha")
abline(v = true_alpha, col = "#b06a3b", lty = 2, lwd = 1.5)
par(op)
```

## What was not ported

The upstream notebook ends with two performance benchmarks against PyMC's NUTS
sampler. Those sections depend on a separate probabilistic-programming stack and
compare pythonnet call overhead rather than the library itself, so they are not
ported here.

## Summary

In this notebook you:

- Used ARWMH to adapt the proposal covariance automatically.
- Applied DEMCz and DEMCzs, whose population of past states supplies the
  proposal scale.
- Compared the adaptive samplers with the gradient-based HMC and NUTS.
- Read acceptance rates, R-hat, and ESS to judge sampler behavior.

### Exercises

1. Run ARWMH and DEMCz on the same problem and compare ESS.
2. Increase the sample size or switch families and note how the samplers
   respond.
3. Compare NUTS and HMC on a smooth posterior and report acceptance rates.

## References {#references}

[1] H. Haario, E. Saksman, and J. Tamminen, "An adaptive Metropolis algorithm,"
*Bernoulli*, vol. 7, no. 2, pp. 223-242, 2001.

[2] C. J. F. ter Braak and J. A. Vrugt, "Differential evolution Markov chain
with snooker updater and fewer chains," *Statistics and Computing*, vol. 18,
no. 4, pp. 435-446, 2008.

[3] R. M. Neal, "MCMC using Hamiltonian dynamics," in *Handbook of Markov Chain
Monte Carlo*. CRC Press, 2011.

[4] M. Betancourt, "A conceptual introduction to Hamiltonian Monte Carlo,"
*arXiv preprint arXiv:1701.02434*, 2017.

[5] M. D. Hoffman and A. Gelman, "The No-U-Turn Sampler: Adaptively setting path
lengths in Hamiltonian Monte Carlo," *Journal of Machine Learning Research*,
vol. 15, no. 47, pp. 1593-1623, 2014.

## Reproduction check

The seeded data draws are exact: the port reproduces the C# Mersenne Twister
stream bit-for-bit, and the Python twin prints the same values. Every posterior
comparison against the upstream notebook is statistical, for two reasons stated
honestly: the upstream hand-picks its priors (`Uniform(50, 150)` and
`Uniform(5, 30)` for the Normal; `Uniform(0, 20)` and `Uniform(0.5, 10)` for the
Gumbel) while this port always uses the family's constraint-based uniform
priors, and the chain settings differ, so the sampler streams cannot be
bit-compared. Two posterior-mean literals are additionally asserted as exact
cross-language checks: the Python twin computes the identical values from the
same core stream.

| Quantity | Upstream C# | This port | Status |
|----------|------------:|----------:|--------|
| ARWMH Normal posterior mean of mu | 100.611 | ~100.657 | statistical |
| ARWMH Normal posterior mean of sigma | 15.053 | ~15.097 | statistical |
| DEMCzs Normal posterior mean of mu | 100.596 | ~100.644 | statistical |
| DEMCzs Normal posterior mean of sigma | 15.037 | ~14.997 | statistical |
| Gumbel xi posterior mean, all 5 samplers | 10.375 to 10.422 | 10.387 to 10.400 | statistical |
| Gumbel alpha posterior mean, all 5 samplers | 2.335 to 2.358 | 2.350 to 2.370 | statistical |
| Seeded Normal / Gumbel data draws | (not printed upstream) | bit-exact vs C# stream and Python twin | exact |
| ARWMH Normal and DEMCzs Gumbel posterior means | (cross-language identity) | identical in the Python twin | exact |

The chunk below fails the render if any check drifts.

```{r}
# Upstream: 05_mcmc_adaptive.ipynb, cell 7 (ARWMH Normal), cell 9 (DEMCzs Normal),
# cell 12 (Gumbel sampler-comparison table).

# exact: seeded data and cross-language posterior means, identical in the Python
# twin. The values themselves are bit-exact (the fixture suite enforces that); the
# comparisons carry a 1e-15 relative tolerance only because R's decimal parser can
# land one ulp away from the written literal.
stopifnot(
  abs(data[1] / 86.91533945543637 - 1) < 1e-15,
  abs(gumbel_data[1] / 15.235041366837393 - 1) < 1e-15,
  abs(arwmh$posterior_mean[1] / 100.6571453618418 - 1) < 1e-15,
  abs(arwmh$posterior_mean[2] / 15.096506821290019 - 1) < 1e-15,
  abs(runs[["DEMCzs"]]$posterior_mean[1] / 10.396547639985409 - 1) < 1e-15,
  abs(runs[["DEMCzs"]]$posterior_mean[2] / 2.350782938160503 - 1) < 1e-15
)

# statistical: priors and chain settings differ from upstream (see above), so we
# assert agreement within 5% of the upstream printed values plus CI coverage of truth.
stopifnot(
  abs(arwmh$posterior_mean[1] / 100.611 - 1) < 0.05,
  abs(arwmh$posterior_mean[2] / 15.053 - 1) < 0.05,
  abs(demczs_norm$posterior_mean[1] / 100.596 - 1) < 0.05,
  abs(demczs_norm$posterior_mean[2] / 15.037 - 1) < 0.05
)
for (r in list(arwmh, demczs_norm)) {
  stopifnot(
    r$posterior_lower_ci[1] <= true_mu, true_mu <= r$posterior_upper_ci[1],
    r$posterior_lower_ci[2] <= true_sigma, true_sigma <= r$posterior_upper_ci[2]
  )
}

upstream_gumbel <- list(
  ARWMH = c(10.422, 2.342), DEMCz = c(10.395, 2.349), DEMCzs = c(10.375, 2.335),
  HMC = c(10.379, 2.358), NUTS = c(10.418, 2.356)
)
for (name in names(upstream_gumbel)) {
  r <- runs[[name]]
  up <- upstream_gumbel[[name]]
  stopifnot(
    abs(r$posterior_mean[1] / up[1] - 1) < 0.05,
    abs(r$posterior_mean[2] / up[2] - 1) < 0.05,
    r$posterior_lower_ci[1] <= true_xi, true_xi <= r$posterior_upper_ci[1],
    r$posterior_lower_ci[2] <= true_alpha, true_alpha <= r$posterior_upper_ci[2],
    # internally checkable convergence: seeded, so these are fixed values
    max(r$rhat) < 1.05,
    min(r$ess) > 500
  )
}
cat("All reproduction checks passed.\n")
```
