---
title: "27. Composite distributions"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Most of the distribution surface takes a family name and a parameter vector. Five families do
not, because they are built out of other distributions or out of the data itself: a truncated
distribution, a mixture, a competing-risks system, an empirical curve, and a kernel density.
`corehydror` calls them composite, and each has its own constructor: `dist_truncated()`,
`dist_mixture()`, `dist_competing_risks()`, `dist_empirical()`, and `dist_kde()`. What comes back
is an ordinary `corehydro_dist`: the density, distribution, quantile, random-draw, moment,
parameter, and log-likelihood verbs all take one.

This example uses three of them on one record: a mixture for a basin with two flood-generating
mechanisms, a kernel density as a nonparametric reference, and a truncated distribution for a
physical lower bound.

## What you'll learn

- Fit a two-component mixture with `model_mixture()` and `fit_mle()`, then rebuild the fit as a
  standalone `dist_mixture()` distribution.
- See where a second component changes the frequency curve, and where it does not.
- Build a kernel density, reproduce its default bandwidth, and understand what
  `bounded_by_data` does and does not restrict.
- Renormalize a distribution onto a physical range with `dist_truncated()`.
- Know the two things composites cannot do.

## Setup

```{r}
library(corehydror)
```

## A mixed-population record

Sixty annual peak flows from a basin with two flood seasons: spring snowmelt produces a cluster
of moderate peaks, and summer convective storms produce a smaller number of much larger ones. The
record is written out here so the page is self-contained and both languages read the same
numbers.

```{r}
peaks <- c(
  1240, 8720, 2110, 1240, 11180, 1570, 11620, 4470, 6470, 1960,
  1740, 3530, 7440, 9080, 1500, 5760, 7920, 2480, 1790, 2650,
  1740, 1590, 2080, 2220, 1800, 2120, 11430, 8710, 19870, 1570,
  1540, 5290, 1480, 7750, 1900, 5630, 1910, 20340, 1280, 9270,
  5720, 1390, 14490, 2180, 1590, 1530, 6540, 2470, 2490, 7780,
  13010, 2140, 1220, 11790, 2820, 2000, 2390, 1440, 8740, 2470
)
n <- length(peaks)
cat(sprintf("n = %d, range %s to %s cfs\n", n,
            format(min(peaks), big.mark = ","), format(max(peaks), big.mark = ",")))
```

## One distribution, or two?

`model_mixture()` builds a weighted mixture model, and `fit_mle()` fits it the same way it fits
any other model. The single-family fit is the comparison.

```{r}
single_fit <- fit_mle(model_univariate("LogNormal", peaks))
mix_fit <- fit_mle(model_mixture(c("LogNormal", "LogNormal"), peaks))

round(coef(mix_fit), 4)
```

The parameter vector is the mixture weights first, then each component's own parameters in order,
which is exactly how the core lays a mixture's parameters out.

```{r}
p <- unname(coef(mix_fit))
components <- list(distribution("LogNormal", p[3:4]), distribution("LogNormal", p[5:6]))

for (i in 1:2) {
  cat(sprintf("component %d: weight %.4f, median %s cfs\n", i, p[i],
              format(round(dist_quantile(components[[i]], 0.5)), big.mark = ",")))
}
```

Two seasons, recovered from the record alone: a wide component centered near 8,557 cfs carrying
weight 0.4129, and a narrow one centered near 1,836 cfs carrying the rest.

```{r}
gof <- data.frame(
  parameters = c(length(coef(single_fit)), length(coef(mix_fit))),
  log_likelihood = c(logLik(single_fit), logLik(mix_fit)),
  aic = c(single_fit$aic, mix_fit$aic),
  bic = c(single_fit$bic, mix_fit$bic),
  row.names = c("LogNormal", "Two-component mixture")
)
round(gof, 3)
```

The mixture costs four extra parameters, and both criteria still prefer it: its AIC and its BIC
are lower than the single fit's. A single lognormal cannot be both narrow enough for the snowmelt
cluster and wide enough for the storm peaks.

## Rebuilding the fit as a distribution

`fit_mle()` returns a fit, not a distribution. `dist_mixture()` turns the fitted numbers back
into a `corehydro_dist` the `dist_*()` verbs accept, and the log-likelihood of that composite
over the same record reproduces the fit's own value exactly.

```{r}
mixture <- dist_mixture(components = components, weights = p[1:2])
single <- distribution("LogNormal", unname(coef(single_fit)))

mixture
cat(sprintf("log-likelihood: %.10f from dist_mixture(), %.10f from the fit\n",
            dist_log_likelihood(mixture, peaks), logLik(mix_fit)))
```

## Where the second component changes the answer

```{r}
aeps <- c(0.5, 0.1, 0.01, 0.002)
quantiles <- data.frame(
  single = dist_quantile(single, 1 - aeps),
  mixture = dist_quantile(mixture, 1 - aeps),
  row.names = paste0(100 * aeps, "% AEP")
)
quantiles$ratio <- quantiles$mixture / quantiles$single
round(quantiles, 3)
```

The mixture is not uniformly heavier. It raises the 10% quantile from 9,975 to 11,461 cfs,
because the storm component puts real probability where a single lognormal had a thin shoulder.
It then *lowers* the 1% and 0.2% quantiles, because each component has its own scale and neither
is as diffuse as the single fit forced to straddle both clusters. Splitting a bimodal record into
two components is not the same as making its tail heavier.

```{r}
round(rbind(single = dist_moments(single), mixture = dist_moments(mixture)), 3)
```

The moments say the same thing. The two distributions have nearly the same mean, but the
mixture's standard deviation, skewness, and kurtosis are all smaller and its median is far lower.
Read the mixture's reported mode with care: the core finds it by maximizing the density with a
Brent search over the interval between the 0.1% and 99.9% quantiles, which on a two-humped
density converges to whichever hump the search lands in.

## A kernel density as a nonparametric reference

`dist_kde()` builds a density by summing a kernel centered at each observation. With no bandwidth
supplied it uses the rule `h = s * (4 / (3n))^(1/5)`, where `s` is the sample standard deviation.
Passing that value explicitly reproduces the default exactly.

```{r}
h <- sd(peaks) * (4 / (3 * n))^(1 / 5)
kde <- dist_kde(peaks)
cat(sprintf("bandwidth %.6f; explicit bandwidth reproduces the default: %s\n",
            h, identical(dist_quantile(kde, 0.9), dist_quantile(dist_kde(peaks, bandwidth = h), 0.9))))
```

```{r}
#| fig-alt: "Histogram of the annual peak record with the single lognormal density, the two-component mixture density, and the kernel density overlaid."
grid <- seq(500, 24000, length.out = 400)
hist(peaks,
  breaks = 20, freq = FALSE, border = "white", col = "#d9d2c5",
  xlab = "Annual peak flow (cfs)", main = "Three densities over the same record"
)
lines(grid, dist_pdf(single, grid), lwd = 2, col = "#4d6b8a")
lines(grid, dist_pdf(mixture, grid), lwd = 2, col = "#8c5a3b")
lines(grid, dist_pdf(kde, grid), lwd = 2, lty = 2, col = "#6b7f3f")
legend("topright",
  legend = c("LogNormal", "Two-component mixture", "Kernel density"),
  col = c("#4d6b8a", "#8c5a3b", "#6b7f3f"), lwd = 2, lty = c(1, 1, 2), bty = "n"
)
```

`bounded_by_data`, on by default, sets the distribution's reported minimum and maximum to the
smallest and largest observation instead of extending three bandwidths past each. That gates the
CDF and the quantile function, which cannot then return a value outside the observed range. It
does not gate the density: the kernel sum is evaluated wherever you ask.

```{r}
kde_open <- dist_kde(peaks, bounded_by_data = FALSE)
cat(sprintf("largest observation           %s\n", format(max(peaks), big.mark = ",")))
cat(sprintf("0.2%% AEP quantile, bounded     %.1f\n", dist_quantile(kde, 0.998)))
cat(sprintf("0.2%% AEP quantile, unbounded   %.1f\n", dist_quantile(kde_open, 0.998)))
cat(sprintf("density at 30,000 cfs identical either way: %s\n",
            dist_pdf(kde, 30000) == dist_pdf(kde_open, 30000)))
```

That is the honest limit of a kernel density for design work. With `bounded_by_data` left on it
describes the record it was built from and will not return a quantile past it, which is exactly
why the parametric fits above still have a job. Turned off, the support runs three bandwidths
beyond the largest observation, and the 0.2% AEP quantile lands at 23,424.8 cfs against a record
maximum of 20,340.

```{r}
round(data.frame(
  single = dist_quantile(single, 1 - aeps),
  mixture = dist_quantile(mixture, 1 - aeps),
  kde = dist_quantile(kde, 1 - aeps),
  row.names = paste0(100 * aeps, "% AEP")
), 1)
```

## A physical lower bound

Suppose a downstream diversion guarantees that at least 1,000 cfs passes the gauge in any year,
so an annual maximum below that is impossible. The fitted lognormal does not know this.

```{r}
cat(sprintf("P(annual peak < 1,000 cfs) under the fitted LogNormal: %.4f\n",
            dist_cdf(single, 1000)))
```

`dist_truncated()` restricts a distribution to `[min, max]` and renormalizes: the density becomes
the base density divided by `F(max) - F(min)`, zero outside the bounds, and the CDF becomes
`(F(x) - F(min)) / (F(max) - F(min))`.

```{r}
truncated <- dist_truncated(single, min = 1000, max = 1e6)

cat(sprintf("CDF at the lower bound: %.1f\n", dist_cdf(truncated, 1000)))
cat(sprintf("density ratio at 5,000 cfs: %.10f\n",
            dist_pdf(truncated, 5000) / dist_pdf(single, 5000)))
cat(sprintf("1 / (F(max) - F(min)):       %.10f\n",
            1 / (dist_cdf(single, 1e6) - dist_cdf(single, 1000))))
```

Removing the impossible lower tail pushes every quantile up, and the proportional shift shrinks
as the quantile rises: 1.0706 at the median against 1.0178 at the 0.2% AEP quantile, the part of
the curve a design study reads.

```{r}
shifted <- data.frame(
  base = dist_quantile(single, 1 - aeps),
  truncated = dist_quantile(truncated, 1 - aeps),
  row.names = paste0(100 * aeps, "% AEP")
)
shifted$ratio <- shifted$truncated / shifted$base
round(shifted, 4)
```

## What composites cannot do

Two things. First, none of the five implements linear moments upstream, so `dist_lmoments()`
declines rather than approximating.

```{r}
cat(tryCatch(dist_lmoments(mixture), error = conditionMessage), "\n")
```

Second, a composite has no flat parameter vector, so `distribution()` will not build one and
there is no `dist_fit()` path to a composite family. Mixtures are fitted through
`model_mixture()` and an estimator, as above; competing-risks systems through
`model_competing_risks()`.

```{r}
cat(tryCatch(distribution("Mixture", c(0.5, 0.5)), error = conditionMessage), "\n")
cat(tryCatch(distribution("KernelDensity", 1), error = conditionMessage), "\n")
distribution_names("structured")
```

## Reproduction check

```{r}
# Every fit on this page is deterministic (maximum likelihood by a deterministic optimizer; the
# kernel density and the truncation are closed-form), so these literals are what this port
# computes and what the Python notebook asserts -- which is what proves the cross-language
# identity. The comparisons carry a 1e-15 relative tolerance only because R's decimal parser can
# land one ulp away from the written literal; the fixture suite is what enforces bit-exactness.
near <- \(x, literal) abs(x / literal - 1) < 1e-15
stopifnot(
  # The two maximum-likelihood fits.
  near(coef(single_fit)[1], 3.5398021872265386),
  near(coef(single_fit)[2], 0.35823869011128351),
  near(logLik(single_fit), -562.61957474501503),
  near(coef(mix_fit)[1], 0.41289769144404032),
  near(coef(mix_fit)[2], 0.58710230855595968),
  near(coef(mix_fit)[3], 3.9323412160584628),
  near(coef(mix_fit)[4], 0.18144915787795951),
  near(coef(mix_fit)[5], 3.2639016318942939),
  near(coef(mix_fit)[6], 0.10464430806060765),
  near(logLik(mix_fit), -541.62441863534707),
  near(gof["LogNormal", "aic"], 1129.2391494900301),
  near(gof["Two-component mixture", "aic"], 1095.2488372706941),
  near(gof["LogNormal", "bic"], 1133.4278386144742),
  near(gof["Two-component mixture", "bic"], 1107.8149046440267),

  # The composite rebuilt from the fitted values.
  dist_log_likelihood(mixture, peaks) == logLik(mix_fit),

  # Quantiles: the mixture is higher at 10% and lower at 1% and 0.2%.
  near(quantiles$single[2], 9974.7849506787734),
  near(quantiles$mixture[2], 11461.031257850214),
  near(quantiles$single[3], 23615.111603538928),
  near(quantiles$mixture[3], 19517.845994674906),
  near(quantiles$single[4], 37228.117722712399),
  near(quantiles$mixture[4], 25218.020604342844),
  near(dist_moments(mixture)[["skewness"]], 1.5738083017829989),
  near(dist_moments(single)[["skewness"]], 3.9241120906543738),

  # The kernel density.
  near(h, 2143.864821560871),
  identical(dist_quantile(kde, 0.9), dist_quantile(dist_kde(peaks, bandwidth = h), 0.9)),
  near(dist_quantile(kde, 0.9), 12151.217810408163),
  near(dist_quantile(kde, 0.998), 20088.429334855526),
  near(dist_quantile(kde_open, 0.998), 23424.805148963263),
  dist_pdf(kde, 30000) == dist_pdf(kde_open, 30000),

  # The truncated distribution.
  near(dist_cdf(single, 1000), 0.065928055787046747),
  dist_cdf(truncated, 1000) == 0,
  near(dist_quantile(truncated, 0.5), 3710.536428782953),
  near(dist_quantile(truncated, 0.99), 24116.69387849512),
  near(dist_pdf(truncated, 5000) / dist_pdf(single, 5000), 1.0705813467570697)
)

# Internal consistency: the truncation identities the core documents, the bounded kernel density
# refusing to leave the observed range, and the mixture beating the single fit on both criteria.
stopifnot(
  abs(dist_pdf(truncated, 5000) / dist_pdf(single, 5000) -
        1 / (dist_cdf(single, 1e6) - dist_cdf(single, 1000))) < 1e-12,
  dist_pdf(truncated, 500) == 0,
  dist_quantile(kde, 0.998) < max(peaks),
  dist_quantile(kde_open, 0.998) > max(peaks),
  mix_fit$aic < single_fit$aic, mix_fit$bic < single_fit$bic,
  all(shifted$ratio > 1), all(diff(shifted$ratio) < 0),
  abs(sum(p[1:2]) - 1) < 1e-12
)
cat("All reproduction checks passed.\n")
```
