---
title: "02. Distribution fitting"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Ported from
[`02_distribution_fitting.ipynb`](https://github.com/USACE-RMC/Numerics-Python-Examples/blob/main/notebooks/02_distribution_fitting.ipynb)
in the USACE-RMC Numerics-Python-Examples repository (0BSD licensed). The upstream
notebook drives the C# `Numerics.dll` through pythonnet; this version uses
`corehydror`, whose compiled core is a validated C++ port of the same library, so the
numbers below reproduce the upstream outputs exactly. The Python version of this
example uses the same core and prints the same numbers.

## What you'll learn

- Fitting a distribution by Method of Moments (MOM), Maximum Likelihood (MLE), and
  L-moments.
- Comparing fitted distributions with log-likelihood, AIC, BIC, and RMSE.
- Visual checks: empirical vs fitted CDFs and Weibull plotting positions.
- Ranking many candidate families at once with `fit_distributions`.

## Set up

The upstream setup cell loads the CoreCLR runtime, resolves `Numerics.dll`, and
imports the .NET types. Here the setup is one library call.

```{r}
library(corehydror)
```

## Helper function

The function below plots a histogram of the data with the PDF of a fitted
distribution, the same helper the upstream notebook defines.

```{r}
plot_empirical_vs_model <- function(data, dist, title = "Empirical vs fitted distribution") {
  x <- seq(min(data), max(data), length.out = 500)
  hist(data, breaks = 20, freq = FALSE, col = "#6b7f3f", border = "white",
       main = title, xlab = "Value", ylab = "Density")
  lines(x, dist_pdf(dist, x), col = "#b06a3b", lwd = 2)
  legend("topright", c("Empirical", "Fitted PDF"), bty = "n",
         pch = c(15, NA), lty = c(NA, 1), lwd = c(NA, 2),
         col = c("#6b7f3f", "#b06a3b"))
}
```

## Example data: flood peaks

We simulate 100 annual peak flows from a LogNormal distribution. As in the C#
library, `LogNormal` is the base-10 family: its parameters are the mean and
standard deviation of $\log_{10}(x)$, so `LogNormal(8, 0.5)` produces values around
$10^8$. (The base-e family is `LnNormal`.) The draw is seeded, and the port keeps
the C# Mersenne Twister bit-exact, so these are the same 100 numbers the upstream
notebook generates. The upstream prints the population standard deviation
(NumPy's `np.std`), so we compute that rather than R's sample `sd()`.

```{r}
lognormal <- distribution("LogNormal", c(8, 0.5))
data <- dist_random(lognormal, 100, seed = 123)
std_pop <- sqrt(mean((data - mean(data))^2)) # np.std: population standard deviation

cat("Sample size:", length(data), "\n")
cat(sprintf("Sample mean: %.2f\n", mean(data)))
cat(sprintf("Sample std:  %.2f\n", std_pop))

hist(data, breaks = 20, freq = FALSE, col = "#6b7f3f", border = "white",
     main = "Simulated flood peak data", xlab = "Annual peak flow", ylab = "Density")
```

## Method of Moments (MOM)

The Method of Moments is the oldest and simplest approach to parameter estimation.
The idea: equate sample moments to the corresponding theoretical moments of the
distribution and solve for the parameters. Given a sample $x_1, \ldots, x_n$, the
first two sample moments are the mean and standard deviation:

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i, \qquad
s = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n}(x_i - \bar{x})^2}
$$

A two-parameter family needs only these two; three-parameter families also use the
sample skewness. The Normal$(\mu, \sigma)$ gives $\hat{\mu} = \bar{x}$ and
$\hat{\sigma} = s$ directly, while the base-10 LogNormal solves its log-space
$\mu$ and $\sigma$ from the real-space mean and standard deviation.

Strengths: simple, closed-form solutions, always produces estimates. Weaknesses:
less statistically efficient than MLE, and sensitive to outliers because
conventional moments give disproportionate weight to extreme values.

The upstream code mutates a distribution in place with
`Estimate(data, ParameterEstimationMethod.MethodOfMoments)`; here `dist_fit`
returns the fitted object. The printed values match the upstream `Mean` and
`StandardDeviation` properties, which are real-space moments (the parameters
themselves live in $\log_{10}$ space).

```{r}
mom <- dist_fit("LogNormal", data, method = "mom")
mom_m <- dist_moments(mom)

cat(sprintf("MOM fit: mean=%.2f, sd=%.2f\n", mom_m[["mean"]], mom_m[["sd"]]))
cat(sprintf("  parameters (log10 space): mu=%.4f, sigma=%.4f\n",
            dist_params(mom)[1], dist_params(mom)[2]))
plot_empirical_vs_model(data, mom, title = "MOM fit: LogNormal")
```

## Maximum Likelihood Estimation (MLE)

MLE finds the parameter values that make the observed data most probable under the
assumed model. Given independent observations $x_1, \ldots, x_n$ with PDF
$f(x \mid \boldsymbol{\theta})$, the likelihood is

$$
L(\boldsymbol{\theta} \mid \mathbf{x}) = \prod_{i=1}^{n} f(x_i \mid \boldsymbol{\theta})
$$

and, because products are numerically unstable, optimization is performed on the
log-likelihood:

$$
\ell(\boldsymbol{\theta}) = \sum_{i=1}^{n} \log f(x_i \mid \boldsymbol{\theta}), \qquad
\hat{\boldsymbol{\theta}}_{\text{MLE}} = \underset{\boldsymbol{\theta}}{\text{argmax}} \; \ell(\boldsymbol{\theta})
$$

For some families (Normal, Exponential) the MLE has a closed form; for most
families used in hydrology it is found numerically, with initial values derived
from L-moment estimates. MLE is asymptotically efficient and unbiased, and its
log-likelihood feeds directly into AIC and BIC below. It can be biased for small
samples and requires the full probability model.

The upstream prints `Mean` (real space) and `Sigma` (the $\log_{10}$-space scale
parameter); we print the same two quantities.

```{r}
mle <- dist_fit("LogNormal", data, method = "mle")
mle_m <- dist_moments(mle)

cat(sprintf("MLE fit: mean=%.2f, sigma(log10)=%.2f\n",
            mle_m[["mean"]], dist_params(mle)[2]))
```

## L-moments

L-moments are linear combinations of order statistics that provide robust
alternatives to conventional moments [1]. They are especially valuable for small
samples ($n < 50$), data with outliers, and extreme value analysis. They are
defined through probability-weighted moments (PWMs),

$$
\beta_r = E\left[X \cdot F(X)^r\right], \quad r = 0, 1, 2, \ldots
$$

with the first four L-moments

$$
\lambda_1 = \beta_0, \quad
\lambda_2 = 2\beta_1 - \beta_0, \quad
\lambda_3 = 6\beta_2 - 6\beta_1 + \beta_0, \quad
\lambda_4 = 20\beta_3 - 30\beta_2 + 12\beta_1 - \beta_0
$$

and the dimensionless ratios $\tau = \lambda_2 / \lambda_1$ (L-CV),
$\tau_3 = \lambda_3 / \lambda_2$ (L-skewness), and
$\tau_4 = \lambda_4 / \lambda_2$ (L-kurtosis). Unlike conventional skewness and
kurtosis, the ratios are bounded, which makes them stable and interpretable.
Because L-moments use only linear combinations of order statistics, a single
extreme observation cannot dominate them, and they stay nearly unbiased even for
samples as small as $n = 10$. For hydrological records of 30 to 60 annual values,
that robustness matters.

```{r}
lm_fit <- dist_fit("LogNormal", data, method = "lmom")
lm_m <- dist_moments(lm_fit)

cat(sprintf("L-moments fit: mean=%.2f, sigma(log10)=%.2f\n",
            lm_m[["mean"]], dist_params(lm_fit)[2]))
```

## Comparing the three fits

The three estimation methods land on nearly the same parameters for this sample,
so the fitted PDFs almost coincide.

```{r}
x <- seq(min(data), max(data), length.out = 500)

hist(data, breaks = 20, freq = FALSE, col = adjustcolor("#6b7f3f", 0.5),
     border = "white", main = "Comparison of fitted distributions",
     xlab = "Value", ylab = "Density")
lines(x, dist_pdf(mom, x), lty = 2, lwd = 2, col = "#b06a3b")
lines(x, dist_pdf(mle, x), lty = 2, lwd = 2, col = "#5b7a8c")
lines(x, dist_pdf(lm_fit, x), lty = 2, lwd = 2, col = "#8c8c7a")
legend("topright", c("Data", "MOM", "MLE", "L-moments"), bty = "n",
       pch = c(15, NA, NA, NA), lty = c(NA, 2, 2, 2), lwd = c(NA, 2, 2, 2),
       col = c(adjustcolor("#6b7f3f", 0.5), "#b06a3b", "#5b7a8c", "#8c8c7a"))
```

## Goodness-of-fit diagnostics

We compare the models with the log-likelihood and the information criteria AIC [2]
and BIC [3]. A larger (less negative) log-likelihood and smaller AIC/BIC reflect a
better fit. The upstream notebook calls the C# `GoodnessOfFit` static class; those
metrics are two-line formulas over the log-likelihood, so this port computes them
inline. With $k$ parameters and $n$ observations:

$$
\mathrm{AIC} = 2k - 2\log L, \qquad
\mathrm{BIC} = k \ln(n) - 2\log L
$$

RMSE (smaller is better) compares the sorted observations to the fitted quantiles
at the Weibull plotting positions. One quirk of the C# implementation, reproduced
here so the numbers match the upstream table: it drops the $k$ largest observations
from the sum and divides by $n - k$.

```{r}
k <- 2 # LogNormal has two parameters
n <- length(data)
sorted_data <- sort(data)
pp <- plotting_positions(n) # Weibull: i / (n + 1)

rmse <- function(dist) {
  modeled <- dist_quantile(dist, pp)
  m <- n - k # C# GoodnessOfFit.RMSE drops the k largest values and divides by n - k
  sqrt(sum((modeled[1:m] - sorted_data[1:m])^2) / m)
}

fits <- list("LogNormal (MOM)" = mom, "LogNormal (MLE)" = mle,
             "LogNormal (L-moments)" = lm_fit)
lls <- sapply(fits, \(d) dist_log_likelihood(d, data))
results <- data.frame(
  Distribution = names(fits),
  LogLikelihood = lls,
  AIC = 2 * k - 2 * lls,
  BIC = k * log(n) - 2 * lls,
  RMSE = sapply(fits, rmse),
  row.names = NULL
)
print(results)
```

## Empirical vs fitted CDFs

Comparing the fitted CDFs to the empirical CDF gives a visual check of fit.
The empirical CDF assigns each sorted observation $x_i$ the fraction of values
less than or equal to it, $i/n$: if $\mathrm{ECDF}(x) = 0.80$, about 80% of the
observations are $\leq x$.

```{r}
empirical_cdf <- seq_len(n) / n

plot(sorted_data, empirical_cdf, type = "s", col = "#6b7f3f", lwd = 2,
     main = "Empirical vs fitted CDFs", xlab = "Value", ylab = "CDF")
lines(sorted_data, dist_cdf(mom, sorted_data), lty = 2, col = "#b06a3b")
lines(sorted_data, dist_cdf(mle, sorted_data), lty = 2, col = "#5b7a8c")
lines(sorted_data, dist_cdf(lm_fit, sorted_data), lty = 2, col = "#8c8c7a")
legend("bottomright", bty = "n",
       c("Empirical CDF", "LogNormal (MOM)", "LogNormal (MLE)", "LogNormal (L-moments)"),
       lty = c(1, 2, 2, 2), lwd = c(2, 1, 1, 1),
       col = c("#6b7f3f", "#b06a3b", "#5b7a8c", "#8c8c7a"))
```

## Plotting positions

Plotting positions assign each ordered observation an empirical nonexceedance
probability, the basis for probability paper, hydrologic frequency plots, and
return-period calculations. The Weibull plotting position maps the $i$-th smallest
value in a sample of size $n$ to

$$
p_i = \frac{i}{n + 1}
$$

The upstream calls `PlottingPositions.Weibull(n)`; here `plotting_positions(n)`
computes the same values (Weibull is the default; other conventions are available
through `method =` or `alpha =`).

```{r}
weibull_pp <- plotting_positions(n)

plot(sorted_data, weibull_pp, pch = 16, col = adjustcolor("#6b7f3f", 0.7),
     main = "Weibull plotting positions",
     xlab = "Sorted data", ylab = "Weibull plotting position")
```

## Ranking many families at once

Beyond comparing estimation methods for one family, `fit_distributions` fits 15
candidate families by MLE and reports AIC, BIC, and RMSE for each. On this sample
the LogNormal fits rank near the top, as they should for LogNormal data; two
flexible three-parameter families edge it out by a point of AIC, a normal outcome
for a single 100-value sample. Rows with `converged` FALSE are candidates whose
MLE failed on this sample: here KappaFour, whose four-parameter fit did not
converge on these data. The run is deterministic, seed and all, so the
table is the same on every call and in both languages. (This analysis-level RMSE
follows the C# convention of using the DataFrame's Hirsch-Stedinger plotting
positions, so it is not comparable to the Weibull-based RMSE computed above.)
[Example 16](../16-fifteen-candidate-fitting/r.qmd) is about this verb on its own,
including what the fifteenth candidate is and why a ranking on a short record
should not be trusted.

```{r}
ranking <- as.data.frame(fit_distributions(data))
print(ranking[order(ranking$aic), ], row.names = FALSE)
```

## Summary

You have:

- fit a distribution by MOM, MLE, and L-moments with `dist_fit`,
- compared the fits with log-likelihood, AIC, BIC, and RMSE computed inline,
- checked the fits visually against the empirical CDF and Weibull plotting
  positions,
- ranked 15 candidate families at once with `fit_distributions`.

## Exercise

1. Generate synthetic data from a Weibull distribution
   (`dist_random(distribution("Weibull", c(scale, shape)), n, seed = ...)`).
2. Fit Weibull parameters using MLE and L-moments.
3. Compare fit quality with AIC/BIC.

## References

[1] J. R. M. Hosking, "L-moments: Analysis and estimation of distributions using
linear combinations of order statistics," *Journal of the Royal Statistical
Society: Series B*, vol. 52, no. 1, pp. 105-124, 1990.

[2] H. Akaike, "A new look at the statistical model identification," *IEEE
Transactions on Automatic Control*, vol. 19, no. 6, pp. 716-723, 1974.

[3] G. Schwarz, "Estimating the dimension of a model," *Annals of Statistics*,
vol. 6, no. 2, pp. 461-464, 1978.

## Reproduction check

Values printed by the upstream notebook (run against the real C# library),
compared with this port. Everything here is deterministic seeded math, so the
match is exact; the upstream prints at 2 to 6 decimals, so the assertions compare
the formatted values, and the first seeded draw plus the MLE sigma parameter are
additionally asserted at full precision (the same full-precision values the Python
notebook asserts, the cross-language guarantee).

| Quantity | Upstream C# | This port | Status |
|----------|------------:|----------:|--------|
| Sample mean / std (seed 123) | 158155878.08 / 213541263.95 | `mean(data)` / `std_pop` | exact |
| MOM fit mean / sd | 150193934.68 / 162945821.98 | `dist_moments(mom)` | exact |
| MLE fit mean / sigma | 149606124.89 / 0.38 | `mle_m[["mean"]]` / `dist_params(mle)[2]` | exact |
| L-moments fit mean / sigma | 147689495.62 / 0.37 | `lm_m[["mean"]]` / `dist_params(lm_fit)[2]` | exact |
| Log-likelihood (MOM/MLE/L-mom) | -1972.685554 / -1972.683043 / -1972.712513 | `dist_log_likelihood(d, data)` | exact |
| AIC (MOM/MLE/L-mom) | 3949.371109 / 3949.366086 / 3949.425026 | `2k - 2 log L` | exact |
| BIC (MOM/MLE/L-mom) | 3954.581449 / 3954.576426 / 3954.635366 | `k ln(n) - 2 log L` | exact |
| RMSE (MOM/MLE/L-mom) | 5.995955e+07 / 6.041398e+07 / 6.193926e+07 | inline Weibull RMSE | exact |

The chunk below fails the render if any value drifts.

```{r}
# Upstream: 02_distribution_fitting.ipynb, cells 6, 8, 10, 12, and 16 outputs.
# The upstream prints rounded values, so compare at printed precision; the seeded
# quantities themselves are bit-exact (two full-precision anchors at the end).
stopifnot(
  sprintf("%.2f", mean(data)) == "158155878.08",
  sprintf("%.2f", std_pop) == "213541263.95",
  sprintf("%.2f", mom_m[["mean"]]) == "150193934.68",
  sprintf("%.2f", mom_m[["sd"]]) == "162945821.98",
  sprintf("%.2f", mle_m[["mean"]]) == "149606124.89",
  sprintf("%.2f", dist_params(mle)[2]) == "0.38",
  sprintf("%.2f", lm_m[["mean"]]) == "147689495.62",
  sprintf("%.2f", dist_params(lm_fit)[2]) == "0.37"
)

# Upstream: 02_distribution_fitting.ipynb, cell 16 (goodness-of-fit table).
expected <- data.frame(
  ll = c("-1972.685554", "-1972.683043", "-1972.712513"),
  aic = c("3949.371109", "3949.366086", "3949.425026"),
  bic = c("3954.581449", "3954.576426", "3954.635366"),
  rmse = c("5.995955e+07", "6.041398e+07", "6.193926e+07")
)
stopifnot(
  sprintf("%.6f", results$LogLikelihood) == expected$ll,
  sprintf("%.6f", results$AIC) == expected$aic,
  sprintf("%.6f", results$BIC) == expected$bic,
  sprintf("%.6e", results$RMSE) == expected$rmse
)

# Full-precision anchors, identical across R, Python, and the C# library. The values
# themselves are bit-exact (the fixture suite enforces that); the comparison here
# carries a 1e-15 relative tolerance only because R's decimal parser can land one
# ulp away from the written literal.
stopifnot(abs(data[1] / 180774075.803667 - 1) < 1e-15)
stopifnot(abs(dist_params(mle)[2] / 0.3810574327738998 - 1) < 1e-15)
cat("All reproduction checks passed.\n")
```
