---
title: "04. Bayesian inference with MCMC"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Ported from
[`04_mcmc_bayesian_inference.ipynb`](https://github.com/USACE-RMC/Numerics-Python-Examples/blob/main/notebooks/04_mcmc_bayesian_inference.ipynb)
in the USACE-RMC Numerics-Python-Examples repository (0BSD licensed). The upstream
notebook drives the C# `Numerics.dll` through pythonnet; this version uses
`corehydror`, whose compiled core is a validated C++ port of the same library, so the
seeded MCMC results below reproduce the upstream outputs at every digit the
upstream displays (see the reproduction check at the end for the honest fine
print). The Python version of this example uses the same core and prints the same
numbers.

## What you'll learn

- Bayesian estimation of distribution parameters from real flood data with the
  random walk Metropolis-Hastings (RWMH) sampler.
- How `mcmc_sample()` packages the upstream recipe (constraint-based uniform
  priors, MAP initialization, serial chains, seed 12345) into one call.
- Fitting and comparing several candidate distributions.
- Posterior predictive checks.
- Adaptive versus non-adaptive samplers.

## Setup

The upstream setup loads the CoreCLR runtime, resolves `Numerics.dll`, imports the
MCMC classes, and wraps every Python likelihood in a .NET delegate. It also warns
that the C# sampler's parallel chains contend for Python's GIL, so every upstream
example sets `ParallelizeChains = False`. None of that applies here: the ported
chain driver always runs chains serially (exactly the upstream's
`ParallelizeChains = False` code path), so seeded results are reproducible by
construction.

```{r}
library(corehydror)
```

## Example: flood frequency analysis

Problem: we have annual peak flow data and want to estimate distribution parameters
with uncertainty, using Bayesian inference through an MCMC sampler.

Data source: Tippecanoe River near Delphi, Indiana (from "Flood Frequency
Analysis", A.R. Rao and K.H. Hamed, CRC Press, 2000). The record has 48 annual
peaks.

```{r}
flood = c(6290, 2700, 13100, 16900, 14600, 9600, 7740, 8490, 8130, 12000,
          17200, 15000, 12400, 6960, 6500, 5840, 10400, 18800, 21400, 22600,
          14200, 11000, 12800, 15700, 4740, 6950, 11800, 12100, 20600, 14600,
          14600, 8900, 10600, 14200, 14100, 14100, 12500, 7530, 13400, 17600,
          13400, 19200, 16900, 15500, 14500, 21900, 10400, 7460)

summary_df = data.frame(
  Metric = c("Count", "Sample mean (cfs)", "Sample std (cfs)", "Min (cfs)", "Max (cfs)"),
  Value = round(c(length(flood), mean(flood), sd(flood), min(flood), max(flood)), 2)
)
cat("Flood dataset summary\n")
print(summary_df)

par(mfrow = c(1, 2))
hist(flood, breaks = 15, col = "#6b7f3f", border = "white",
     xlab = "Annual peak flow (cfs)", main = "Observed flood data")
plot(seq_along(flood), sort(flood), type = "o", pch = 16, col = "#5b7a8c",
     xlab = "Rank", ylab = "Annual peak flow (cfs)", main = "Ranked flood magnitudes")
par(mfrow = c(1, 1))
```

### Fit a Normal distribution

The upstream builds every piece by hand: it asks the distribution for its parameter
constraints (`GetParameterConstraints`), turns them into uniform priors (for the
Normal: mean ~ Uniform(-1000000, 1000000) and standard deviation ~
Uniform(0, 100000)), wraps a Python log-likelihood in a .NET delegate, and runs
`RWMH` with MAP initialization and the default seed 12345.

`mcmc_sample()` is that exact recipe in one call: the same constraint-based uniform
priors, the same MAP initialization, the same serial chain driver, and the same C#
defaults (four chains, 3500 recorded draws per chain with a thinning interval of
20, the first 1750 treated as warm-up, plus 10000 dedicated posterior draws that
feed the summary statistics).

```{r}
fit_norm = mcmc_sample(flood, distribution = "Normal", sampler = "RWMH", seed = 12345)

posterior_table = function(fit) {
  data.frame(
    Parameter = fit$parameters,
    `Posterior mean` = round(fit$posterior_mean, 2),
    `Lower 90% CI` = round(fit$posterior_lower_ci, 2),
    `Upper 90% CI` = round(fit$posterior_upper_ci, 2),
    check.names = FALSE
  )
}

cat("Normal distribution results\n")
print(posterior_table(fit_norm))
```

These are the numbers in the upstream results table: posterior mean 12677.64 with
90 percent credible interval (11522.23, 13847.04) for the mean, and 4844.70 with
(4073.12, 5735.17) for the standard deviation. Plot the fit at the posterior-mean
parameters against the data, like the upstream does.

```{r}
plot_fit = function(fit, family) {
  d = distribution(family, fit$posterior_mean)
  x = seq(min(flood) * 0.85, max(flood) * 1.10, length.out = 400)
  par(mfrow = c(1, 2))
  hist(flood, breaks = 15, freq = FALSE, col = "#6b7f3f", border = "white",
       xlab = "Annual peak flow (cfs)", main = paste(family, "fit: density"))
  lines(x, sapply(x, \(v) dist_pdf(d, v)), col = "#b06a3b", lwd = 2)
  s = sort(flood)
  ecdf_y = seq_along(s) / length(s)
  plot(s, ecdf_y, type = "s", col = "#8c8c7a", lwd = 2,
       xlab = "Annual peak flow (cfs)", ylab = "CDF",
       main = paste(family, "fit: CDF"))
  lines(s, sapply(s, \(v) dist_cdf(d, v)), col = "#b06a3b", lwd = 2)
  par(mfrow = c(1, 1))
}

plot_fit(fit_norm, "Normal")
```

### Fit a Logistic distribution

The Logistic distribution has heavier tails than the Normal [1]. The same one-liner
fits it; only the family name changes.

```{r}
fit_logi = mcmc_sample(flood, distribution = "Logistic", sampler = "RWMH", seed = 12345)
cat("Logistic distribution results\n")
print(posterior_table(fit_logi))
plot_fit(fit_logi, "Logistic")
```

### Comparing multiple distribution fits

Add a Gumbel fit and compare the three candidates. The upstream compares the fits
visually and tabulates the posterior means; we add a plug-in log-likelihood
(evaluated at the posterior-mean parameters) and AIC to make the ranking explicit.

```{r}
fit_gum = mcmc_sample(flood, distribution = "Gumbel", sampler = "RWMH", seed = 12345)

fits = list(Normal = fit_norm, Logistic = fit_logi, Gumbel = fit_gum)

par(mfrow = c(1, 3))
x = seq(2000, 25000, length.out = 400)
for (name in names(fits)) {
  d = distribution(name, fits[[name]]$posterior_mean)
  hist(flood, breaks = 15, freq = FALSE, col = "#8c8c7a", border = "white",
       xlab = "Peak flow (cfs)", main = name)
  lines(x, sapply(x, \(v) dist_pdf(d, v)), col = "#b06a3b", lwd = 2)
}
par(mfrow = c(1, 1))

comparison = do.call(rbind, lapply(names(fits), \(name) {
  fit = fits[[name]]
  ll = dist_log_likelihood(distribution(name, fit$posterior_mean), flood)
  data.frame(
    Distribution = name,
    `Parameter 1` = sprintf("%s = %.0f", fit$parameters[1], fit$posterior_mean[1]),
    `Parameter 2` = sprintf("%s = %.0f", fit$parameters[2], fit$posterior_mean[2]),
    `Log-likelihood` = round(ll, 2),
    AIC = round(2 * 2 - 2 * ll, 2),
    check.names = FALSE
  )
}))
cat("Model comparison\n")
print(comparison)
```

The posterior means match the upstream comparison table (Normal 12678 / 4845,
Logistic 12633 / 2826, Gumbel 10386 / 4512). The Normal edges out the other two on
log-likelihood for this record, and `fit$map_fitness` (the log-likelihood at the
posterior mode, so higher is better) gives the same ordering.

## Posterior predictive checks

A key part of Bayesian modeling is checking whether the fitted model can reproduce
the observed data. Posterior predictive checks draw parameter sets from the
posterior and generate synthetic replicates from the model at those parameters [2].
The upstream picks its replicate indices with an unseeded NumPy generator, so its
exact numbers are not reproducible; here every draw goes through the seeded core
RNG, and the Python twin produces bit-identical replicates.

```{r}
# The last 1000 recorded draws of chain 1 are the posterior sample we draw from.
chain1 = fit_norm$chains[[1]]
tail_rows = chain1[(nrow(chain1) - 999):nrow(chain1), ]

# One predictive value per posterior draw (for summary statistics).
pred = vapply(1:1000, \(i) {
  dist_random(distribution("Normal", tail_rows[i, ]), 1, seed = 10000 + i - 1)
}, numeric(1))

# 50 replicated datasets of the same size as the record (for the ECDF overlay).
reps = lapply(1:50, \(i) {
  dist_random(distribution("Normal", tail_rows[(i - 1) * 20 + 1, ]),
              length(flood), seed = 20000 + i - 1)
})

plot(NULL, xlim = range(c(flood, unlist(reps))), ylim = c(0, 1),
     xlab = "Annual peak flow (cfs)", ylab = "CDF",
     main = "Posterior predictive check: ECDF overlay (50 replicates)")
for (r in reps) {
  s = sort(r)
  lines(s, seq_along(s) / length(s), type = "s",
        col = adjustcolor("#5b7a8c", alpha.f = 0.25))
}
s = sort(flood)
lines(s, seq_along(s) / length(s), type = "s", col = "black", lwd = 2)
legend("bottomright", legend = "Observed", col = "black", lwd = 2, bty = "n")

ppc_summary = data.frame(
  Statistic = c("Mean", "Std dev", "5th percentile", "95th percentile"),
  Observed = round(c(mean(flood), sd(flood),
                     quantile(flood, 0.05), quantile(flood, 0.95)), 2),
  `Posterior predictive` = round(c(mean(pred), sd(pred),
                                   quantile(pred, 0.05), quantile(pred, 0.95)), 2),
  check.names = FALSE, row.names = NULL
)
cat("Posterior predictive summary\n")
print(ppc_summary)
```

`corehydror` also ships a turnkey version. `posterior_predictive_check()` fits the
model by MCMC (DEMCz by default), draws replicate datasets, and returns Bayesian
p-values for five test statistics. Values near 0.5 mean the observed statistic sits
in the middle of the predictive distribution; values near 0 or 1 flag misfit.

```{r}
ppc = posterior_predictive_check(flood, "Normal", seed = 12345)
pvals = data.frame(
  Statistic = c("Mean", "Std dev", "Skewness", "Minimum", "Maximum"),
  `Bayesian p-value` = c(ppc$mean_p_value, ppc$sd_p_value, ppc$skewness_p_value,
                         ppc$min_p_value, ppc$max_p_value),
  check.names = FALSE
)
print(pvals)
cat("Misfit flagged:", if (ppc$has_misfit) "yes" else "no", "\n")
```

## Linear regression with uncertainty

The upstream notebook also builds a Bayesian linear regression from a hand-written
likelihood. `mcmc_sample()` fits named distribution families and does not take a
custom likelihood, so that section is not expressible here. Two other pages cover
the same ground: [example 20](../20-bayesian-regression/r.qmd) does the regression
itself, and [example 14](../14-custom-posterior/r.qmd) writes a log-likelihood by
hand and samples it with `mcmc_posterior()`.

## Non-adaptive vs adaptive samplers

Adaptive MCMC learns the proposal covariance from the chain's own history instead
of relying on manual tuning [3]. The upstream demonstrates the idea with DEMCzs;
here we use ARWMH, the adaptive variant of the same random walk sampler, which
makes the comparison like for like. (DEMCz, DEMCzs, HMC, NUTS, and SNIS are also
available through `sampler =`.)

```{r}
fit_arwmh = mcmc_sample(flood, distribution = "Normal", sampler = "ARWMH", seed = 12345)

compare = data.frame(
  Statistic = c("Posterior mean (location)", "Posterior sd (location)",
                "Posterior mean (scale)", "Posterior sd (scale)",
                "Mean acceptance rate"),
  RWMH = round(c(fit_norm$posterior_mean[1], fit_norm$posterior_sd[1],
                 fit_norm$posterior_mean[2], fit_norm$posterior_sd[2],
                 mean(fit_norm$acceptance_rates)), 4),
  ARWMH = round(c(fit_arwmh$posterior_mean[1], fit_arwmh$posterior_sd[1],
                  fit_arwmh$posterior_mean[2], fit_arwmh$posterior_sd[2],
                  mean(fit_arwmh$acceptance_rates)), 4)
)
cat("Sampler comparison\n")
print(compare)

par(mfrow = c(2, 2))
plot(fit_norm$chains[[1]][, 1], type = "l", lwd = 0.4, col = "#6b7f3f",
     xlab = "Recorded draw", ylab = "Mean (cfs)", main = "RWMH: trace of the mean (chain 1)")
abline(v = 1750, col = "#8c8c7a", lty = 2)
plot(fit_arwmh$chains[[1]][, 1], type = "l", lwd = 0.4, col = "#b06a3b",
     xlab = "Recorded draw", ylab = "Mean (cfs)", main = "ARWMH: trace of the mean (chain 1)")
abline(v = 1750, col = "#8c8c7a", lty = 2)
post_rwmh = unlist(lapply(fit_norm$chains, \(ch) ch[1751:nrow(ch), 1]))
hist(post_rwmh, breaks = 40, col = "#6b7f3f", border = "white",
     xlab = "Mean (cfs)", main = "RWMH: posterior of the mean")
post_arwmh = unlist(lapply(fit_arwmh$chains, \(ch) ch[1751:nrow(ch), 1]))
hist(post_arwmh, breaks = 40, col = "#b06a3b", border = "white",
     xlab = "Mean (cfs)", main = "ARWMH: posterior of the mean")
par(mfrow = c(1, 1))
```

The dashed line marks the end of warm-up. Both samplers agree on the posterior.
ARWMH settles at a lower acceptance rate (about 0.37 versus 0.44) because the
adapted proposal takes larger, better-scaled steps; that is the intended behavior,
not a defect.

## Summary

In this example you:

- Applied Bayesian inference to real flood frequency data with `mcmc_sample()`.
- Fit Normal, Logistic, and Gumbel distributions by MCMC and compared the
  posterior parameter estimates.
- Ran posterior predictive checks, both by hand from the chains and with
  `posterior_predictive_check()`.
- Compared a non-adaptive sampler (RWMH) with its adaptive variant (ARWMH).

## References

[1] N. L. Johnson, S. Kotz, and N. Balakrishnan, *Continuous Univariate
Distributions*, vols. 1-2, 2nd ed. Wiley, 1994.

[2] A. Gelman, J. B. Carlin, H. S. Stern, D. B. Dunson, A. Vehtari, and D. B.
Rubin, *Bayesian Data Analysis*, 3rd ed. CRC Press, 2013.

[3] C. J. F. ter Braak and J. A. Vrugt, "Differential evolution Markov chain with
snooker updater and fewer chains," *Statistics and Computing*, vol. 18, no. 4,
pp. 435-446, 2008.

## Reproduction check

This port uses the upstream's exact settings: the same constraint-based uniform
priors, RWMH with MAP initialization, serial chains, and the default seed 12345.
The C# and C++ samplers replay the identical accept-reject stream; all four RWMH
and ARWMH acceptance rates match a direct run of the C# library bit for bit. The
draws themselves are not bit-identical: the MAP starting point comes from a
differential evolution optimization whose optimum differs from the C# result by
about 1e-11 relative (accumulated last-digit floating point drift), and that offset
carries through every subsequent draw at about 1e-9 relative. Every digit the
upstream notebook displays therefore reproduces, and the assertions below pin the
values at that precision. The one place the upstream prints six decimals (its
sampler-comparison table shows 12677.641313 for the posterior mean of the Normal
mean parameter) this port gives 12677.641312, so we state the difference rather
than assert a wrong number.

| Quantity | Upstream C# | This port | Status |
|----------|------------:|----------:|--------|
| Normal posterior mean (mean, sd) | 12677.64 / 4844.70 | same at 2 dp | exact at displayed precision |
| Normal 90% CI (mean) | (11522.23, 13847.04) | same at 2 dp | exact at displayed precision |
| Normal 90% CI (sd) | (4073.12, 5735.17) | same at 2 dp | exact at displayed precision |
| Logistic posterior mean (location, scale) | 12632.66 / 2825.51 | same at 2 dp | exact at displayed precision |
| Logistic 90% CI (location) | (11438.17, 13801.22) | same at 2 dp | exact at displayed precision |
| Logistic 90% CI (scale) | (2308.14, 3440.53) | same at 2 dp | exact at displayed precision |
| Comparison table means (3 families) | 12678/4845, 12633/2826, 10386/4512 | same at 0 dp | exact at displayed precision |
| RWMH / ARWMH acceptance rates | not printed upstream | match the C# library bit for bit | exact |
| First recorded Normal draw (chain 1) | not printed upstream | 13287.2303500459 | exact (cross-language) |
| First posterior predictive draw | recast (unseeded upstream) | 14255.412844410343 | exact (cross-language) |
| Built-in PPC p-values (mean, sd) | recast (unseeded upstream) | 0.497 / 0.495 | exact (cross-language) |

The chunk below fails the render if any value drifts. The cross-language rows
assert the exact values the Python notebook asserts.

```{r}
# Upstream: 04_mcmc_bayesian_inference.ipynb, cell 7 output (Normal results table, 2 dp)
stopifnot(
  round(fit_norm$posterior_mean[1], 2) == 12677.64,
  round(fit_norm$posterior_lower_ci[1], 2) == 11522.23,
  round(fit_norm$posterior_upper_ci[1], 2) == 13847.04,
  round(fit_norm$posterior_mean[2], 2) == 4844.70,
  round(fit_norm$posterior_lower_ci[2], 2) == 4073.12,
  round(fit_norm$posterior_upper_ci[2], 2) == 5735.17
)

# Upstream: cell 9 output (Logistic results table, 2 dp)
stopifnot(
  round(fit_logi$posterior_mean[1], 2) == 12632.66,
  round(fit_logi$posterior_lower_ci[1], 2) == 11438.17,
  round(fit_logi$posterior_upper_ci[1], 2) == 13801.22,
  round(fit_logi$posterior_mean[2], 2) == 2825.51,
  round(fit_logi$posterior_lower_ci[2], 2) == 2308.14,
  round(fit_logi$posterior_upper_ci[2], 2) == 3440.53
)

# Upstream: cell 11 output (model comparison table, 0 dp)
stopifnot(
  round(fit_norm$posterior_mean[1]) == 12678,
  round(fit_norm$posterior_mean[2]) == 4845,
  round(fit_logi$posterior_mean[1]) == 12633,
  round(fit_logi$posterior_mean[2]) == 2826,
  round(fit_gum$posterior_mean[1]) == 10386,
  round(fit_gum$posterior_mean[2]) == 4512
)

# Seeded cross-language identity: the Python notebook asserts the same values.
# The values are bit-exact across R, Python, and C#'s accept-reject stream; the
# comparisons carry a 1e-15 relative tolerance only because R's decimal parser
# can land one ulp away from the written literal.
stopifnot(
  abs(fit_norm$chains[[1]][1, 1] / 13287.2303500459 - 1) < 1e-15,
  abs(fit_norm$chains[[1]][1, 2] / 4176.928509438068 - 1) < 1e-15,
  abs(fit_norm$map_fitness / -473.55840961774754 - 1) < 1e-15,
  abs(fit_norm$acceptance_rates[1] / 0.4352416666666667 - 1) < 1e-15,
  abs(fit_arwmh$acceptance_rates[1] / 0.35859166666666664 - 1) < 1e-15,
  abs(pred[1] / 14255.412844410343 - 1) < 1e-15,
  ppc$mean_p_value == 0.497,
  ppc$sd_p_value == 0.495,
  ppc$has_misfit == 0
)
cat("All reproduction checks passed.\n")
```
