---
title: "18. Numerical methods"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

The rest of this site's examples fit a distribution, a copula, or a model to data. This one steps
back a layer, to the general-purpose numerical toolkit those fits are built on (P2 "math extras"):
root finding, integration in one, two, and many dimensions plus an ODE solver, interpolation, a
linear-system solve, and the two non-tabular univariate functions. None of it is specific to
hydrology, so every example below borrows a hydrologic frame where one is natural rather than
inventing an abstract one.

## What you'll learn

- The four methods `root_find()` offers, and why Newton's quadratic convergence lands closer to
  the true root than Brent's default bracket search on the same problem.
- `ode_solve()` against a problem with a closed-form answer.
- The ten `quadrature()` methods compared on one integrand, and why the fixed-rule methods need
  more steps than their default to compete with the adaptive ones.
- A seeded `quadrature_nd()` Vegas run configured for a rare event with `target_probability`.
- `interpolate()`'s cubic spline against a straight line on a stage-discharge curve.
- Solving a linear system with `qr_solve()`, and evaluating a power law and its inverse with
  `univariate_function()`.
- Where the cross-language guarantee holds for the three stochastic integrators, and where it
  measurably does not.

## Setup

```{r}
library(corehydror)
```

## Finding a root: Brent vs Newton

`root_find()` solves $f(x) = 0$. The default, Brent's method, and its two relatives (bisection,
secant) all need a bracket -- an interval over which $f$ changes sign. Newton-Raphson needs
something different: an analytic derivative and a starting guess, with the bracket optional. Both
find the same root of $f(x) = x^2 - 2$, $\sqrt{2}$, but not to the same precision:

```{r}
f <- function(x) x^2 - 2
brent <- root_find(f, lower = 0, upper = 2)
newton <- root_find(f, method = "newton", df = function(x) 2 * x, first_guess = 1)
c(brent = brent, newton = newton, true = sqrt(2))
```

Newton's quadratic convergence lands on the true value to every digit R prints; Brent's search
stops once its own default tolerance on the root (`1e-8`) is satisfied, which is a looser promise
and leaves it a few units in the last place short -- tightenable with `tolerance`, not a defect in
either method.

The same root reappears as the intersection of a circle and a line, this time with two equations
and `root_find_system()`'s multivariate Newton-Raphson, which needs a Jacobian instead of a scalar
derivative:

```{r}
circle_and_line <- function(v) c(v[1]^2 + v[2]^2 - 4, v[1] - v[2])
circle_and_line_jacobian <- function(v) matrix(c(2 * v[1], 2 * v[2], 1, -1), nrow = 2, byrow = TRUE)
root_find_system(circle_and_line, circle_and_line_jacobian, first_guess = c(1, 1))
```

$x^2 + y^2 = 4$ meets $x = y$ at $(\sqrt{2}, \sqrt{2})$, the same number as above.

## An ordinary differential equation

`ode_solve()` steps a first-order ODE $dy/dt = f(t, y)$ forward from an initial value. A decaying
tracer concentration, $dC/dt = -kC$, has a closed form, $C(t) = C_0 e^{-kt}$, which makes it a
clean check on the numerical solver rather than a demonstration of something otherwise unknowable:

```{r}
k <- 0.5
decay <- ode_solve(function(t, c) -k * c, initial_value = 10, start_time = 0,
                   end_time = 5, time_steps = 51, method = "rk4")
t <- seq(0, 5, length.out = 51)
analytic <- 10 * exp(-k * t)
cat(sprintf("RK4 C(5) = %.6f, analytic = %.6f, difference = %.2e\n",
            decay[51], analytic[51], decay[51] - analytic[51]))
```

```{r}
op <- par(mar = c(4.5, 4.5, 3, 1))
plot(t, analytic, type = "l", col = "#5b7a8c", lwd = 4,
     xlab = "time", ylab = "concentration", main = "RK4 against the closed form")
points(t[c(1, 11, 21, 31, 41, 51)], decay[c(1, 11, 21, 31, 41, 51)], pch = 19, col = "#b06a3b")
legend("topright", c("analytic", "ode_solve() (rk4)"), lty = c(1, NA), pch = c(NA, 19),
       lwd = c(4, NA), col = c("#5b7a8c", "#b06a3b"), bty = "n")
par(op)
```

Fifty steps of fourth-order Runge-Kutta agree with the closed form to about a part in $10^7$ --
`method = "rkf"`/`"cash_karp"` trade the fixed grid above for adaptive step control instead, at a
tolerance you set directly.

## Integrating one function ten ways

`quadrature()` wraps ten ported rules for $\int_a^b f(x)\,dx$: five adaptive (`"gauss_kronrod"`,
the default; `"simpsons"`; `"trapezoidal"`; `"adaptive_simpsons"`; `"gauss_lobatto"`) that subdivide
until two nested estimates agree, and five fixed rules with no such refinement
(`"gauss_legendre"`, `"gauss_legendre20"`, `"simpsons_fixed"`, `"trapezoidal_fixed"`,
`"midpoint"`). All ten integrate the standard normal density over $[-3, 3]$, against the closed-form
answer $2\Phi(3) - 1 = 0.9973002039367398$:

```{r}
g <- function(x) exp(-x^2 / 2) / sqrt(2 * pi)
true_value <- 2 * pnorm(3) - 1

adaptive <- c("gauss_kronrod", "simpsons", "trapezoidal", "adaptive_simpsons", "gauss_lobatto")
fixed <- c("gauss_legendre", "gauss_legendre20", "simpsons_fixed", "trapezoidal_fixed", "midpoint")

results <- data.frame(method = c(adaptive, fixed), value = NA_real_, error = NA_real_)
for (i in seq_along(adaptive)) {
  q <- quadrature(g, -3, 3, method = adaptive[i])
  results$value[i] <- as.numeric(q)
}
for (i in seq_along(fixed)) {
  q <- quadrature(g, -3, 3, method = fixed[i])  # default steps = 2
  results$value[length(adaptive) + i] <- as.numeric(q)
}
results$error <- results$value - true_value
print(results, digits = 10, row.names = FALSE)
```

The five adaptive methods agree with the true value to 8-10 digits without being told how finely to
subdivide. The five fixed rules do not, at their default of two steps -- `"midpoint"` in particular
is off by two parts in a thousand. That is not a weaker method, only an unconfigured one: `steps`
sets how many, and there is no adaptive refinement to fall back on if you leave it low.

```{r}
for (m in c("simpsons_fixed", "trapezoidal_fixed", "midpoint")) {
  q <- quadrature(g, -3, 3, method = m, steps = 1000)
  cat(sprintf("%-16s steps=1000  value=%.10f  error=%.2e\n", m, as.numeric(q),
              as.numeric(q) - true_value))
}
```

At 1000 steps all three land within $10^{-4}$ or better, `"simpsons_fixed"` matching the adaptive
methods' precision. `quadrature_2d()` is the two-dimensional sibling, an adaptive Simpson's rule
over a rectangle -- $\int_0^1\int_0^1 xy\,dx\,dy = 1/4$ exactly:

```{r}
quadrature_2d(function(x, y) x * y, min_x = 0, max_x = 1, min_y = 0, max_y = 1)
```

## A rare event with Vegas

`quadrature_nd()` reaches beyond two dimensions with three ported stochastic integrators: plain
Monte Carlo (the default), Miser (recursive stratified sampling), and Vegas (adaptive importance
sampling). Vegas is the one built for a rare event: `target_probability` calls the ported
`configure_for_rare_events()` helper, which reshapes the sampling grid toward wherever the
integrand is large rather than spending most of its budget on a region that contributes almost
nothing.

The event here is the joint tail of two independent standard Normal draws, $P(Z_1 + Z_2 > 4)$, with
an analytic answer close enough to check against: $Z_1 + Z_2 \sim N(0, 2)$, so the probability is
$1 - \Phi(4/\sqrt{2})$. `method = "vegas"` takes `f(x, weight)` rather than `f(x)`, and mapping the
unit square through the standard Normal quantile turns the integral over $[0,1]^2$ directly into a
probability, with no extra scaling:

```{r}
norm <- distribution("Normal", c(0, 1))
rare_event <- function(x, weight) {
  z1 <- dist_quantile(norm, x[1])
  z2 <- dist_quantile(norm, x[2])
  if (z1 + z2 > 4) 1 else 0
}
vegas_p <- quadrature_nd(rare_event, min = c(0, 0), max = c(1, 1), method = "vegas",
                         seed = 12345, target_probability = 0.01)
true_p <- 1 - dist_cdf(norm, 4 / sqrt(2))
cat(sprintf("vegas estimate = %.7f, analytic = %.7f, function evaluations = %d\n",
            as.numeric(vegas_p), true_p, attr(vegas_p, "function_evaluations")))
```

Half a million evaluations of an indicator that is 1 on roughly one point in four hundred still
lands within 1% of the analytic probability -- rare-event sampling working as intended, not brute
force. `attr(vegas_p, "chi_squared")` is Vegas's own internal-consistency diagnostic, not a
goodness-of-fit test against the true value; a value near 1 across its independent evaluations says
the run's importance grid had converged.

## What reproduces across languages, and what does not

Most pages on this site can promise a seeded run gives bit-identical numbers in R and Python,
because every operation happens inside the shared C++ core and both packages call the same compiled
code. `quadrature_nd()` needs a narrower promise, measured rather than assumed, and it differs by
method.

`MonteCarloIntegration`'s own arithmetic -- a running count of hits divided by the sample size -- has
no near-cancelling subtraction for a fused-multiply-add to leave a mark on, so its `Result`
reproduces bit-for-bit in this package, in `corehydropy`, in the C++ fixture runner under both FMA
settings, and against the real C# library; `fixtures/callback/callback_cross_language.json` pins it
at zero tolerance for exactly that reason. Miser and Vegas are not so lucky: both compute a
`standard_error` shaped like `avg2 - avg*avg`, and Vegas's own `chi_squared` compounds a second such
subtraction, and clang/gcc's default fused-multiply-add contraction does not agree with .NET's
never-fused arithmetic on the last bit of either. Measured directly against the real C# library
(`fixtures/callback/math.json`'s `quadrature_miser_gsl` case): Miser's own `Result` misses the C#
value by 1 ULP under this package's shipped build. Measured directly R against Python at the
settings `fixtures/callback/callback_cross_language.json` itself uses for its Vegas sub-block
(`NdW_SumOfNormals3`, `independent_evaluations = 2`, `function_calls = 300`): `Result` differs by 2
ULP language to language, not just in `standard_error`/`chi_squared`. It is a property of the
classes' own floating-point arithmetic, not a bug in either binding -- and it is why that fixture's
`quadrature_nd_short_exact` case asserts `function_evaluations` and `status` on every method,
`Result` additionally on Monte Carlo alone, and nothing else. The same three constructs, run here:

```{r}
nd_pi <- function(x) if (x[1]^2 + x[2]^2 < 1) 1 else 0
monte_carlo <- quadrature_nd(nd_pi, min = c(-1, -1), max = c(1, 1), method = "monte_carlo",
                             seed = 999, min_iterations = 20, max_iterations = 200,
                             relative_tolerance = 0.1)

nd_gsl <- function(x) (1 / pi^3) / (1 - cos(x[1]) * cos(x[2]) * cos(x[3]))
miser <- quadrature_nd(nd_gsl, min = c(0, 0, 0), max = c(pi, pi, pi), method = "miser",
                       seed = 999, max_function_evaluations = 300)

mu20 <- c(10, 30, 17); sigma20 <- c(2, 15, 5)
sum_of_normals <- function(x, weight) {
  acc <- 0
  for (i in 1:3) acc <- acc + mu20[i] + sigma20[i] * dist_quantile(norm, x[i])
  acc
}
vegas <- quadrature_nd(sum_of_normals, min = rep(1e-16, 3), max = rep(0.9999999999999999, 3),
                       method = "vegas", independent_evaluations = 2, function_calls = 300)

data.frame(
  method = c("monte_carlo", "miser", "vegas"),
  function_evaluations = c(attr(monte_carlo, "function_evaluations"),
                           attr(miser, "function_evaluations"),
                           attr(vegas, "function_evaluations")),
  status = c(attr(monte_carlo, "status"), attr(miser, "status"), attr(vegas, "status"))
)
```

## Interpolating a stage-discharge curve

`interpolate()` mirrors three C# interpolaters -- linear, cubic spline, and polynomial. A straight
line between rating-curve knots kinks at every one of them; a cubic spline is smooth through all of
them, which matters wherever the interpolated value itself gets used in something sensitive to its
derivative:

```{r}
stage <- c(0, 2, 4, 6, 8, 10)
discharge <- c(0, 150, 620, 1450, 2600, 4100)
stage_out <- seq(0, 10, by = 0.25)

linear_q <- interpolate(stage, discharge, stage_out, method = "linear")
spline_q <- interpolate(stage, discharge, stage_out, method = "cubic_spline")

op <- par(mar = c(4.5, 4.5, 3, 1))
plot(stage_out, linear_q, type = "l", col = "#5b7a8c", lwd = 2,
     xlab = "stage", ylab = "discharge", main = "Linear vs cubic spline")
lines(stage_out, spline_q, col = "#b06a3b", lwd = 2)
points(stage, discharge, pch = 19, col = "#6b7f3f")
legend("topleft", c("linear", "cubic spline", "knots"), lty = c(1, 1, NA), pch = c(NA, NA, 19),
       lwd = c(2, 2, NA), col = c("#5b7a8c", "#b06a3b", "#6b7f3f"), bty = "n")
par(op)
```

At the knots the two agree exactly; between them the straight line always underestimates a convex
stretch of the curve, which is what "linear" is doing here -- connecting the dots rather than
following the curvature the knots themselves imply.

## A linear system

`qr_solve()` mirrors the C# `QRDecomposition::Solve` overloads: Householder QR decomposition
followed by back-substitution, for a vector or a matrix right-hand side. A square system with a
known integer solution is enough to see it work:

```{r}
a <- matrix(c(3, 2, -1, 2, -2, 4, -1, 0.5, -1), nrow = 3, byrow = TRUE)
b <- c(1, -2, 0)
x <- qr_solve(a, b)
x
a %*% x  # reproduces b
```

`qr_decomposition()` exposes the `q`/`r` factors directly, and `gauss_jordan()` is the alternative
full-pivot route to the same kind of answer (plus, unlike `qr_solve()`, the matrix inverse itself).

## A power function, evaluated and inverted

`univariate_function()` evaluates the two non-tabular Numerics `IUnivariateFunction`
implementations. `"power"`, $Y = \alpha(X - \xi)^\beta$, is a power law of the shape a rating curve
itself often takes; `inverse = TRUE` evaluates its algebraic inverse instead of the forward
function, which recovers the input exactly on a deterministic call:

```{r}
discharge_at_stage <- univariate_function("power", c(5, 2, 0, 3), 6)
stage_at_discharge <- univariate_function("power", c(5, 2, 0, 3), discharge_at_stage, inverse = TRUE)
c(discharge = discharge_at_stage, recovered_stage = stage_at_discharge)
```

$5 \times (6 - 0)^2 = 180$, and the inverse recovers $6$ back exactly. `"linear"`,
$Y = \alpha + \beta X$, is the other type; both accept `confidence_level` to evaluate the
non-deterministic path over $\epsilon \sim N(0, \sigma)$ instead.

## Key takeaways

1. `root_find()`'s four methods split into two families: three (`"brent"`, `"bisection"`,
   `"secant"`) need a bracket, `"newton"` needs a derivative and a starting guess. All four solve
   the same problem to different precisions and at different costs.
2. The ten `quadrature()` methods are not interchangeable defaults -- the five fixed rules need
   `steps` set explicitly to compete with the five adaptive ones, which set their own refinement.
3. `quadrature_nd()`'s three methods are a tradeoff between simplicity (`"monte_carlo"`),
   stratified refinement (`"miser"`), and importance sampling (`"vegas"`), and only
   `"monte_carlo"`'s aggregated result is proven to reproduce across every one of R, Python, C++,
   and C#.
4. `target_probability` is what turns Vegas from a general integrator into a rare-event sampler --
   it reconfigures the grid toward the tail rather than requiring you to importance-sample by hand.
5. `interpolate()`'s three methods diverge between knots, not at them; which one is right depends
   on whether the curve you are modeling actually has the smoothness a spline assumes.
6. `qr_solve()`, `qr_decomposition()`, and `gauss_jordan()` all solve `a %*% x = b`; they differ in
   what else they hand back (the factors, or the inverse) and in whether `a` must be square.

## Reproduction check

Two halves. The first is a regression on this page's own numbers -- deterministic values, checked
at 1e-15 relative tolerance since R's decimal parser can land one ulp off a written literal. The
second reruns the three `quadrature_nd()` constructs `fixtures/callback/callback_cross_language.json`
itself pins against the real C# library at zero tolerance, checking only what that fixture asserts:
`function_evaluations` and `status` on every method, the `Result` value additionally on
`"monte_carlo"` alone -- Miser's and Vegas's own floating-point fields are deliberately not checked
here for the reason given above.

```{r}
near <- \(x, literal, tol = 1e-15) abs(x / literal - 1) < tol

stopifnot(
  near(brent, 1.4142135623731364),
  near(newton, 1.4142135623730951),
  near(as.numeric(root_find_system(circle_and_line, circle_and_line_jacobian, first_guess = c(1, 1))[1]),
       1.4142135623730951),
  near(decay[51], 0.82085009767073757736),
  near(as.numeric(quadrature(g, -3, 3)), 0.99730020393673990409),
  near(as.numeric(vegas_p), 0.0023315896197769283606),
  near(discharge_at_stage, 180),
  near(stage_at_discharge, 6),
  near(x[1], 0.99999999999999888978),
  near(x[2], -1.9999999999999971134),
  near(x[3], -1.9999999999999984457),
  near(spline_q[5], interpolate(stage, discharge, 1, method = "cubic_spline")),

  # The callback_cross_language.json construct: EMITTER-READ from the real C# library.
  near(as.numeric(monte_carlo), 3.4285714285714284),
  attr(monte_carlo, "function_evaluations") == 21L,
  attr(monte_carlo, "status") == "Success",
  attr(miser, "function_evaluations") == 300L,
  attr(miser, "status") == "None",
  attr(vegas, "function_evaluations") == 500L,
  attr(vegas, "status") == "Success"
)
cat("All reproduction checks passed.\n")
```
