---
title: "01. Probability distributions"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Ported from
[`01_distributions.ipynb`](https://github.com/USACE-RMC/Numerics-Python-Examples/blob/main/notebooks/01_distributions.ipynb)
in the USACE-RMC Numerics-Python-Examples repository (0BSD licensed). The upstream
notebook drives the C# `Numerics.dll` through pythonnet; this version uses
`corehydror`, whose compiled core is a validated C++ port of the same library, so the
numbers below reproduce the upstream outputs exactly. The Python version of this
example uses the same core and prints the same numbers.

## What you'll learn

- Creating and parameterizing distributions by family name.
- Computing PDF, CDF, and quantile values.
- Statistical properties (mean, variance, skewness, kurtosis).
- When to use each distribution.
- Random sampling, including Latin hypercube sampling.

The upstream notebook tours ten of the 43 univariate families in the library:
Normal, Log-Normal, GEV, Gamma, Weibull, Uniform, Triangular, PERT, Poisson, and
Binomial. This port keeps every family and condenses the long mathematical
derivations; see the upstream notebook for the full detail.

## Setup

The upstream setup loads the CoreCLR runtime and imports each distribution class
from `Numerics.Distributions`. Here one `library()` call covers everything, and
every family is constructed by name through `distribution()`.

```{r}
library(corehydror)

# Earth-tone palette used throughout
olive <- "#6b7f3f"
clay <- "#b06a3b"
slate <- "#5b7a8c"
gray <- "#8c8c7a"

cat("Setup complete\n")
```

## Basic example: the Normal distribution

The Normal (Gaussian) distribution is the most widely used continuous
distribution. It arises naturally from the Central Limit Theorem: the sum of many
independent random variables tends toward a Normal distribution regardless of the
underlying distributions. It is parameterized by its mean $\mu$ and standard
deviation $\sigma$:

$$
f(x) = \frac{1}{\sigma\sqrt{2\pi}} \exp\!\left(-\frac{(x - \mu)^2}{2\sigma^2}\right), \quad -\infty < x < \infty
$$

Note that the library reports Pearson (non-excess) kurtosis, which is 3.0 for a
Normal distribution. The "excess kurtosis" convention is Pearson minus 3.

**When to use:** symmetric data, many natural phenomena, Central Limit Theorem
applications.

```{r}
# Create Normal distribution -- Normal(mean = 100, sd = 15)
normal <- distribution("Normal", c(100, 15))
m <- dist_moments(normal)

cat("Mean:", m[["mean"]], "\n")
cat("Standard Deviation:", m[["sd"]], "\n")
cat("Variance:", m[["sd"]]^2, "\n")
cat("Skewness:", m[["skewness"]], "\n")
cat("Kurtosis:", m[["kurtosis"]], "\n")
cat("Minimum:", m[["minimum"]], "\n")
cat("Maximum:", m[["maximum"]], "\n")
cat("5th Percentile:", format(dist_quantile(normal, 0.05), digits = 17), "\n")
cat("Median:", dist_quantile(normal, 0.5), "\n")
cat("95th Percentile:", format(dist_quantile(normal, 0.95), digits = 17), "\n")

# Plot the PDF and CDF over the 0.1% to 99.9% quantile range.
# dist_pdf/dist_cdf/dist_quantile are all vectorized.
x <- seq(dist_quantile(normal, 0.001), dist_quantile(normal, 0.999),
         length.out = 500)

plot(x, dist_pdf(normal, x), type = "l", col = olive, lwd = 2,
     xlab = "x", ylab = "PDF", main = "Probability density function")
polygon(c(x, rev(x)), c(dist_pdf(normal, x), rep(0, length(x))),
        col = adjustcolor(olive, alpha.f = 0.25), border = NA)

plot(x, dist_cdf(normal, x), type = "l", col = clay, lwd = 2,
     xlab = "x", ylab = "CDF", main = "Cumulative distribution function")
```

## Helper functions

Two small helpers make the rest of the tour concise: one plots the PDF and CDF
side by side, the other prints the statistics table the upstream notebook shows
for each family.

```{r}
plot_distribution <- function(dist, x_range = NULL, n_points = 500,
                              title = NULL) {
  # Plot PDF and CDF of a distribution.
  if (is.null(x_range)) {
    x_range <- dist_quantile(dist, c(0.001, 0.999))
  }
  x <- seq(x_range[1], x_range[2], length.out = n_points)
  op <- par(mfrow = c(1, 2))
  plot(x, dist_pdf(dist, x), type = "l", col = olive, lwd = 2,
       xlab = "x", ylab = "PDF", main = title)
  polygon(c(x, rev(x)), c(dist_pdf(dist, x), rep(0, length(x))),
          col = adjustcolor(olive, alpha.f = 0.25), border = NA)
  plot(x, dist_cdf(dist, x), type = "l", col = clay, lwd = 2,
       xlab = "x", ylab = "CDF")
  par(op)
}

print_statistics <- function(dist, name = "Distribution") {
  # Print the statistics table upstream shows for each family.
  m <- dist_moments(dist)
  q <- dist_quantile(dist, c(0.05, 0.5, 0.95))
  df <- data.frame(
    Statistic = c("Mean", "Std Deviation", "Variance", "Skewness",
                  "Kurtosis", "Minimum", "Maximum", "5th Percentile",
                  "Median", "95th Percentile"),
    Value = round(c(m[["mean"]], m[["sd"]], m[["sd"]]^2, m[["skewness"]],
                    m[["kurtosis"]], m[["minimum"]], m[["maximum"]], q), 4)
  )
  cat("\n", name, " statistics:\n", sep = "")
  print(df, row.names = FALSE)
}
```

## Continuous distributions

A parameterization note before the tour: the library uses its own parameter
conventions (for example, GEV uses $\xi$, $\alpha$, $\kappa$ in the Hosking
convention). If you compare to textbooks or other R packages, double-check the
parameter definitions. `dist_params(d)` shows the constructor order for any
family.

### Log-Normal

A random variable $X$ follows a Log-Normal distribution if $\log(X)$ follows a
Normal distribution. It suits strictly positive, right-skewed data arising from
multiplicative processes. Matching the C# library, the `"LogNormal"` family uses
**base-10** logarithms: the parameters are the mean and standard deviation of
$\log_{10}(X)$. The upstream C# class has a mutable `Base` property; that is not
exposed here. Use the `"LnNormal"` family when you want the base-$e$ form.

**When to use:** right-skewed data that is always positive (flows, particle
sizes, incomes).

```{r}
lognormal <- distribution("LogNormal", c(4.0, 0.5))

print_statistics(lognormal, "LogNormal(4.0, 0.5)")
plot_distribution(lognormal, title = "Log-Normal distribution")
```

### Generalized Extreme Value (GEV)

The GEV distribution unifies three classical extreme value distributions into a
single three-parameter family. It is the limiting distribution for block maxima
(annual maximum floods, peak wind speeds) under the Fisher-Tippett-Gnedenko
theorem. The parameterization is location $\xi$, scale $\alpha > 0$, and shape
$\kappa$ in the Hosking convention, where the sign of $\kappa$ follows L-moment
theory:

$$
F(x) = \exp(-e^{-y}), \qquad
y = \begin{cases}
-\dfrac{1}{\kappa}\ln\!\left(1 - \kappa\dfrac{x - \xi}{\alpha}\right) & \kappa \neq 0 \\[6pt]
\dfrac{x - \xi}{\alpha} & \kappa = 0
\end{cases}
$$

- $\kappa < 0$: Frechet type (heavy upper tail, bounded below)
- $\kappa = 0$: Gumbel (unbounded)
- $\kappa > 0$: Weibull type (bounded above)

The mean exists when $|\kappa| < 1$ and the variance when $|\kappa| < 1/2$, which
is why two of the three tables below show NaN moments.

**When to use:** extreme events (floods, droughts, maxima and minima).

```{r}
# Hosking convention: kappa < 0 => Frechet; kappa > 0 => Weibull type
gev_frechet <- distribution("GeneralizedExtremeValue", c(100, 15, -0.5))
gev_gumbel <- distribution("GeneralizedExtremeValue", c(100, 15, 0.0))
gev_weibull <- distribution("GeneralizedExtremeValue", c(100, 15, 0.5))

x <- seq(50, 200, length.out = 500)
plot(x, dist_pdf(gev_frechet, x), type = "l", col = olive, lwd = 2,
     xlab = "x", ylab = "PDF", main = "GEV: effect of the shape parameter")
lines(x, dist_pdf(gev_gumbel, x), col = clay, lwd = 2)
lines(x, dist_pdf(gev_weibull, x), col = slate, lwd = 2)
legend("topright",
       legend = c("kappa = -0.5 (Frechet)", "kappa = 0.0 (Gumbel)",
                  "kappa = 0.5 (Weibull)"),
       col = c(olive, clay, slate), lwd = 2, bty = "n")

print_statistics(gev_frechet, "GEV (Frechet type)")
print_statistics(gev_gumbel, "GEV (Gumbel type)")
print_statistics(gev_weibull, "GEV (Weibull type)")
```

### Gamma

The Gamma distribution is a flexible two-parameter family for positive-valued
data. It generalizes the Exponential distribution and appears in waiting-time
problems, rainfall modeling, and Bayesian statistics. The factory name is
`"GammaDistribution"` (matching the C# class name), with scale $\theta$ first and
shape $\kappa$ second:

$$
f(x) = \frac{x^{\kappa-1}\, e^{-x/\theta}}{\theta^{\kappa}\,\Gamma(\kappa)}, \quad x > 0
$$

**When to use:** positive data, waiting times, rainfall amounts.

```{r}
gamma_dist <- distribution("GammaDistribution", c(2.0, 0.5))
print_statistics(gamma_dist, "Gamma(scale=2, shape=0.5)")
plot_distribution(gamma_dist, x_range = c(0, 15), title = "Gamma distribution")
```

### Weibull

The Weibull distribution is a flexible two-parameter model for nonnegative data.
Its shape parameter controls the failure rate: $\kappa < 1$ gives decreasing
hazard (early failures), $\kappa = 1$ reduces to the Exponential distribution, and
$\kappa > 1$ yields increasing hazard (wear-out):

$$
F(x) = 1 - \exp\left[-\left(\frac{x}{\lambda}\right)^{\kappa}\right], \quad x \ge 0
$$

**When to use:** reliability analysis, wind speeds, failure times.

```{r}
weibull <- distribution("Weibull", c(100, 2.5))
print_statistics(weibull, "Weibull(scale=100, shape=2.5)")
plot_distribution(weibull, x_range = c(0, 150), title = "Weibull distribution")
```

## Bounded distributions

Three families for quantities with known bounds. The **Uniform** distribution
assigns equal density to every value in $[a, b]$: maximum uncertainty given only
the support. The **Triangular** distribution adds a most likely value (the mode)
between the bounds, a common shape for expert elicitation. The **PERT**
distribution (factory name `"Pert"`) takes the same minimum, most likely, and
maximum inputs but maps them to a smooth Beta density on $[a, b]$ with mean
$(a + 4c + b)/6$, giving gentler tails than the Triangular.

**When to use:** bounded quantities; Uniform when only the range is known,
Triangular or PERT when a most likely value is available (project estimates,
expert opinion).

```{r}
uniform <- distribution("Uniform", c(0, 100))
print_statistics(uniform, "Uniform(0, 100)")
plot_distribution(uniform, x_range = c(-10, 110), title = "Uniform(0, 100)")
```

```{r}
triangular <- distribution("Triangular", c(0, 30, 100))
print_statistics(triangular, "Triangular(0, 30, 100)")
plot_distribution(triangular, x_range = c(-10, 110),
                  title = "Triangular(0, 30, 100)")
```

```{r}
pert <- distribution("Pert", c(10, 50, 100))
print_statistics(pert, "PERT(10, 50, 100)")

# Compare PERT vs Triangular with the same min / most likely / max
tri <- distribution("Triangular", c(10, 50, 100))
x <- seq(5, 105, length.out = 500)
plot(x, dist_pdf(pert, x), type = "l", col = olive, lwd = 2,
     xlab = "x", ylab = "PDF", main = "PERT vs Triangular")
lines(x, dist_pdf(tri, x), col = clay, lwd = 2, lty = 2)
legend("topright", legend = c("PERT", "Triangular"),
       col = c(olive, clay), lwd = 2, lty = c(1, 2), bty = "n")
```

## Discrete distributions

The library uses `dist_pdf()` for both continuous and discrete families; for
discrete distributions, read `dist_pdf()` as the probability mass function (PMF).

### Poisson

The Poisson distribution models counts of events over a fixed interval when
events occur independently at a constant average rate $\lambda$:

$$
P(X=k) = \frac{\lambda^{k} e^{-\lambda}}{k!}, \quad k = 0, 1, 2, \ldots
$$

Mean and variance are both $\lambda$.

**When to use:** count data, events per time period, rare events.

```{r}
poisson <- distribution("Poisson", c(3.5))
print_statistics(poisson, "Poisson(3.5)")

k <- 0:14
op <- par(mfrow = c(1, 2))
barplot(dist_pdf(poisson, k), names.arg = k, col = olive, border = "white",
        xlab = "k (count)", ylab = "PMF")
plot(k, dist_cdf(poisson, k), type = "s", col = clay, lwd = 2,
     xlab = "k (count)", ylab = "CDF")
points(k, dist_cdf(poisson, k), col = clay, pch = 16, cex = 0.7)
par(op)

cat(sprintf("P(X = 3): %.4f\n", dist_pdf(poisson, 3)))
cat(sprintf("P(X <= 5): %.4f\n", dist_cdf(poisson, 5)))
```

### Binomial

The Binomial distribution models the number of successes in $n$ independent
Bernoulli trials with success probability $p$:

$$
P(X=k) = \binom{n}{k} p^k (1-p)^{n-k}, \quad k = 0, 1, \ldots, n
$$

The constructor order matches the C# class: probability first, then the number of
trials.

**When to use:** fixed number of trials with binary outcomes.

```{r}
binomial <- distribution("Binomial", c(0.3, 20))
print_statistics(binomial, "Binomial(p=0.3, n=20)")

k <- 0:20
barplot(dist_pdf(binomial, k), names.arg = k, col = olive, border = "white",
        xlab = "k (number of successes)", ylab = "PMF",
        main = "Binomial(p=0.3, n=20)")

bm <- dist_moments(binomial)
cat(sprintf("Expected value: %.2f\n", bm[["mean"]]))
cat(sprintf("Most likely value: %d\n", as.integer(bm[["mean"]])))
```

## Generating and analyzing random samples

Every univariate distribution has a `dist_random()` method (the port of
`GenerateRandomValues`). The upstream notebook draws these samples unseeded; here
they are seeded so the histograms and the table below reproduce exactly, in both
languages. The sample statistics therefore differ from the upstream run's
unseeded table, but match the Python version of this page bit for bit.

```{r}
n_samples <- 10000

normal_samples <- dist_random(distribution("Normal", c(100, 15)),
                              n_samples, seed = 42)
lognormal_samples <- dist_random(distribution("LogNormal", c(4, 0.5)),
                                 n_samples, seed = 42)
gamma_samples <- dist_random(distribution("GammaDistribution", c(2, 0.5)),
                             n_samples, seed = 42)

op <- par(mfrow = c(1, 3))
hist(normal_samples, breaks = 50, freq = FALSE, col = olive, border = "white",
     main = "Normal", xlab = "Value")
hist(lognormal_samples, breaks = 50, freq = FALSE, col = olive,
     border = "white", main = "Log-Normal", xlab = "Value")
hist(gamma_samples, breaks = 50, freq = FALSE, col = olive, border = "white",
     main = "Gamma", xlab = "Value")
par(op)

sample_skew <- function(x) {
  # Adjusted Fisher-Pearson sample skewness (same formula as the Python twin).
  n <- length(x)
  g1 <- mean((x - mean(x))^3) / mean((x - mean(x))^2)^1.5
  g1 * sqrt(n * (n - 1)) / (n - 2)
}

stats_df <- data.frame(
  Distribution = c("Normal", "Log-Normal", "Gamma"),
  Sample.Mean = c(mean(normal_samples), mean(lognormal_samples),
                  mean(gamma_samples)),
  Sample.Std = c(sd(normal_samples), sd(lognormal_samples),
                 sd(gamma_samples)),
  Sample.Skew = c(sample_skew(normal_samples), sample_skew(lognormal_samples),
                  sample_skew(gamma_samples))
)
cat("\nSample statistics (seed=42):\n")
print(stats_df, row.names = FALSE)
```

### Latin hypercube sampling

Latin hypercube sampling (LHS) is a stratified sampling technique. Each
dimension's $[0, 1)$ range is divided into $n$ equal strata and exactly one
sample lands in each stratum:

$$
x_{ij} = \frac{\pi_j(i) + U_{ij}}{n}, \quad i = 0, \ldots, n-1
$$

where $\pi_j$ is an independent random permutation for each dimension and
$U_{ij} \sim \text{Uniform}(0, 1)$. The `median = TRUE` variant replaces $U_{ij}$
with 0.5, placing each point at its stratum center. Projected onto any single
axis, the samples fall exactly one per stratum, which removes the clustering and
gaps of simple random sampling. LHS often reaches the accuracy of plain Monte
Carlo with 5 to 10 times fewer samples.

**When to use:** Monte Carlo with a limited budget, sensitivity analysis,
calibration of expensive models, risk assessment.

```{r}
lhs <- latin_hypercube(200, 2, seed = 12345)

plot(lhs[, 1], lhs[, 2], pch = 16, cex = 0.6, col = slate,
     main = "Latin hypercube samples (Uniform [0,1])",
     xlab = "Dim 1", ylab = "Dim 2")

# The defining property: sorted samples in each dimension fall one per stratum.
n <- nrow(lhs)
stratified <- all(vapply(seq_len(ncol(lhs)), function(j) {
  s <- sort(lhs[, j])
  all(s >= (seq_len(n) - 1) / n & s < seq_len(n) / n)
}, logical(1)))
cat("One sample per stratum in every dimension:", stratified, "\n")
```

## Comparing multiple distributions

More than one distribution may fit a dataset well, so it pays to consider several
candidates. Here we simulate 50 years of annual peak flows from a Log-Normal
distribution (seeded, so the data reproduce exactly) and fit three candidates by
maximum likelihood. Every family fits through `dist_fit()`; GEV also has a dedicated
`gev_fit()` helper whose bespoke path carries quantile standard errors.
Distribution fitting is covered properly in the next example.

```{r}
# Simulate flood frequency data (seeded; upstream uses the same seed)
observed_data <- dist_random(distribution("LogNormal", c(8, 0.6)),
                             50, seed = 123)

lognormal_fit <- dist_fit("LogNormal", observed_data, method = "mle")
weibull_fit <- dist_fit("Weibull", observed_data, method = "mle")
gev_fitted <- dist_fit("GeneralizedExtremeValue", observed_data, method = "mle")
g <- dist_params(gev_fitted)

print(lognormal_fit)
cat("Weibull fit: params =", round(dist_params(weibull_fit), 4), "\n")
cat("GEV fit:\n")
print(g)

hist(observed_data, breaks = 20, freq = FALSE, col = gray, border = "white",
     main = "Comparing distribution fits", xlab = "Annual peak flow",
     ylab = "Density")
x <- seq(min(observed_data), max(observed_data), length.out = 500)
lines(x, dist_pdf(lognormal_fit, x), col = olive, lwd = 2)
lines(x, dist_pdf(gev_fitted, x), col = clay, lwd = 2)
lines(x, dist_pdf(weibull_fit, x), col = slate, lwd = 2)
legend("topright",
       legend = c("Observed data", "Log-Normal", "GEV", "Weibull"),
       col = c(gray, olive, clay, slate), lwd = c(8, 2, 2, 2), bty = "n")
```

## Distribution selection guide

| Distribution | Type | Typical use case | Parameters |
|---|---|---|---|
| Normal | Continuous | Symmetric data, measurements | $\mu$, $\sigma$ |
| Log-Normal | Continuous | Right-skewed positive data | $\mu$, $\sigma$ |
| Uniform | Continuous | Equal probability over range | min, max |
| Triangular | Continuous | Bounded with mode | min, mode, max |
| PERT | Continuous | Project estimates | min, mode, max |
| Gamma | Continuous | Positive data, waiting times | $\theta$, $\kappa$ |
| Weibull | Continuous | Reliability, failure time | $\lambda$, $\kappa$ |
| Exponential | Continuous | Time between events | $\lambda$ |
| Gumbel | Continuous | Extreme maxima | $\xi$, $\alpha$ |
| GEV | Continuous | General extreme values | $\xi$, $\alpha$, $\kappa$ |
| Beta | Continuous | Proportions on [0,1] | $\alpha$, $\beta$ |
| Poisson | Discrete | Count data, rare events | $\lambda$ |
| Binomial | Discrete | Binary outcomes, trials | $n$, $p$ |
| Bernoulli | Discrete | Single trial success/failure | $p$ |

## Summary

You have now explored:

- Creating distributions with different parameters.
- Computing PDF, CDF, and quantile values.
- Generating random samples, including Latin hypercube designs.
- Visualizing distributions.
- Understanding when to use each distribution.

## Exercise

Try creating and plotting these distributions on your own (use `dist_params(d)`
to check the constructor order):

1. `distribution("Beta", c(2, 5))` - useful for probabilities and proportions.
2. `distribution("StudentT", ...)` with 3 degrees of freedom - a heavy-tailed
   alternative to the Normal.
3. `distribution("ChiSquared", c(5))` - goodness-of-fit testing.

## Reproduction check

Values printed by the upstream notebook (run against the real C# library),
compared with this port. Everything here is deterministic, so the match is exact;
most upstream tables print 4 decimal places (or 7 significant figures for the
Log-Normal), so those rows are compared at the precision upstream shows. The
seeded draws, the Latin hypercube design, and the fitted parameters are
additionally asserted bit-identical to the Python version of this page.

| Quantity | Upstream C# | This port | Status |
|----------|------------:|----------:|--------|
| `Normal(100,15).InverseCDF(0.05)` | 75.32719559572791 | `dist_quantile(normal, 0.05)` | exact |
| `Normal(100,15).InverseCDF(0.95)` | 124.67280440427207 | `dist_quantile(normal, 0.95)` | exact |
| `LogNormal(4,0.5).Mean` | 1.940096e+04 | `dist_moments()[["mean"]]` | exact (7 sig figs shown) |
| `LogNormal(4,0.5).Kurtosis` | 346.8718 | `dist_moments()[["kurtosis"]]` | exact (4 dp shown) |
| `LogNormal(4,0.5).InverseCDF(0.95)` | 6.643957e+04 | `dist_quantile(., 0.95)` | exact (7 sig figs shown) |
| GEV(100,15,-0.5) mean / min | 123.1736 / 70.0 | `dist_moments()` | exact (4 dp shown) |
| GEV(100,15,0) mean / sd | 108.6582 / 19.2382 | `dist_moments()` | exact (4 dp shown) |
| GEV(100,15,0.5) mean / max | 103.4132 / 130.0 | `dist_moments()` | exact (4 dp shown) |
| Gamma(2,0.5) mean / kurtosis | 1.0 / 15.0 | `dist_moments()` | exact |
| Gamma(2,0.5).InverseCDF(0.95) | 3.8415 | `dist_quantile(., 0.95)` | exact (4 dp shown) |
| Weibull(100,2.5) mean | 88.7264 | `dist_moments()[["mean"]]` | exact (4 dp shown) |
| Uniform(0,100) sd / kurtosis | 28.8675 / 1.8 | `dist_moments()` | exact |
| Triangular(0,30,100) mean / median | 43.3333 / 40.8392 | `dist_moments()`, `dist_quantile()` | exact (4 dp shown) |
| PERT(10,50,100) mean / median | 51.6667 / 51.2732 | `dist_moments()`, `dist_quantile()` | exact (4 dp shown) |
| Poisson(3.5) P(X=3) / P(X<=5) | 0.2158 / 0.8576 | `dist_pdf(., 3)`, `dist_cdf(., 5)` | exact (4 dp shown) |
| Binomial(0.3,20) mean / 95th pct | 6.0 / 9.0 | `dist_moments()`, `dist_quantile()` | exact |
| First draw, `LogNormal(8,0.6)` seed 123 | (not printed upstream) | bit-identical to Python twin | exact |
| LHS(200, 2, seed 12345) first point | (not printed upstream) | bit-identical to Python twin | exact |

The chunk below fails the render if any value drifts. The final block asserts the
cross-language guarantee: the seeded draws, the Latin hypercube design, and the
fitted Log-Normal parameters are the exact values the Python notebook prints.

```{r}
# Upstream reference values: 01_distributions.ipynb, cells 4, 8, 10, 12, 14,
# 16, 18, 20, 22, 24 outputs. Upstream tables print DataFrame.round(4), so
# 4-decimal literals are compared at that precision; the underlying math is
# deterministic.
shown <- function(x, literal, tol = 5e-5) abs(x - literal) < tol

# Long-decimal literals are compared with a 1e-15 relative tolerance only
# because R's decimal parser can land one ulp away from the written literal.
rel <- function(x, literal, tol = 1e-15) abs(x / literal - 1) < tol

# Normal(100, 15) -- full-precision prints upstream
stopifnot(
  rel(dist_quantile(normal, 0.05), 75.32719559572791),
  dist_quantile(normal, 0.5) == 100.0,
  rel(dist_quantile(normal, 0.95), 124.67280440427207)
)

# LogNormal(4, 0.5) -- shown to 7 significant figures upstream
ln_m <- dist_moments(lognormal)
stopifnot(
  rel(ln_m[["mean"]], 1.940096e4, tol = 1e-6),
  rel(ln_m[["sd"]], 3.225447e4, tol = 1e-6),
  shown(ln_m[["skewness"]], 9.5827),
  shown(ln_m[["kurtosis"]], 346.8718),
  rel(dist_quantile(lognormal, 0.05), 1.505127e3, tol = 1e-6),
  rel(dist_quantile(lognormal, 0.95), 6.643957e4, tol = 1e-6)
)

# GEV shapes
fm <- dist_moments(gev_frechet)
gm <- dist_moments(gev_gumbel)
wm <- dist_moments(gev_weibull)
stopifnot(
  shown(fm[["mean"]], 123.1736), fm[["minimum"]] == 70.0,
  shown(dist_quantile(gev_frechet, 0.95), 202.4619),
  shown(gm[["mean"]], 108.6582), shown(gm[["sd"]], 19.2382),
  shown(gm[["skewness"]], 1.1396), gm[["kurtosis"]] == 5.4,
  shown(dist_quantile(gev_gumbel, 0.5), 105.4977),
  shown(wm[["mean"]], 103.4132), wm[["maximum"]] == 130.0,
  shown(dist_quantile(gev_weibull, 0.95), 123.2056)
)

# Gamma(2, 0.5)
gam_m <- dist_moments(gamma_dist)
stopifnot(
  gam_m[["mean"]] == 1.0, gam_m[["kurtosis"]] == 15.0,
  shown(gam_m[["sd"]], 1.4142), shown(gam_m[["skewness"]], 2.8284),
  shown(dist_quantile(gamma_dist, 0.5), 0.4549),
  shown(dist_quantile(gamma_dist, 0.95), 3.8415)
)

# Weibull(100, 2.5)
wb_m <- dist_moments(weibull)
stopifnot(
  shown(wb_m[["mean"]], 88.7264), shown(wb_m[["sd"]], 37.9667),
  shown(dist_quantile(weibull, 0.95), 155.0962)
)

# Uniform, Triangular, PERT
un_m <- dist_moments(uniform)
tr_m <- dist_moments(triangular)
pt_m <- dist_moments(pert)
stopifnot(
  un_m[["mean"]] == 50.0, un_m[["kurtosis"]] == 1.8,
  shown(un_m[["sd"]], 28.8675),
  shown(tr_m[["mean"]], 43.3333), shown(tr_m[["skewness"]], 0.3561),
  shown(dist_quantile(triangular, 0.5), 40.8392),
  shown(dist_quantile(triangular, 0.95), 81.2917),
  shown(pt_m[["mean"]], 51.6667), shown(pt_m[["sd"]], 16.9617),
  shown(dist_quantile(pert, 0.5), 51.2732),
  shown(dist_quantile(pert, 0.95), 80.2234)
)

# Poisson(3.5) and Binomial(0.3, 20)
po_m <- dist_moments(poisson)
bi_m <- dist_moments(binomial)
stopifnot(
  po_m[["mean"]] == 3.5, shown(po_m[["sd"]], 1.8708),
  shown(dist_pdf(poisson, 3), 0.2158),
  shown(dist_cdf(poisson, 5), 0.8576),
  dist_quantile(poisson, 0.95) == 7.0,
  bi_m[["mean"]] == 6.0, shown(bi_m[["sd"]], 2.0494),
  dist_quantile(binomial, 0.05) == 3.0,
  dist_quantile(binomial, 0.95) == 9.0
)

# Seeded values, bit-identical across R, Python, and the C# library
stopifnot(
  rel(normal_samples[1], 95.20221411105254),
  rel(observed_data[1], 203499515.64666739),
  rel(lhs[1, 1], 0.12249509909539484),
  rel(lhs[1, 2], 0.81019698383868677),
  stratified
)
# Log-Normal MLE is deterministic (closed-form in log space); the tolerance
# only guards platform libm differences.
ln_params <- dist_params(lognormal_fit)
stopifnot(
  rel(ln_params[[1]], 8.0174794125819, tol = 1e-12),
  rel(ln_params[[2]], 0.383659272555292, tol = 1e-12)
)

cat("All reproduction checks passed.\n")
```
