---
title: "13. A custom objective"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

Every other fitting verb in this package -- `dist_fit()`, `fit_mle()`, `model_bulletin17c()` and
friends -- wraps a model the port already knows how to build. Sometimes the model is yours: a
custom rating curve, a regional regression, anything you can write as a likelihood in a handful
of lines. `optim_minimize()`/`optim_maximize()` expose the same six ported Numerics optimizers
those verbs use internally, over a plain R function, so a hand-written likelihood gets the same
optimizers -- including the same seeded differential evolution, whose PARAMETERS reproduce
bit-exact across languages -- as the built-in models. It has no upstream counterpart: the USACE-RMC Numerics-Python-Examples
repository has no general-purpose optimization notebook.

## What you'll learn

- Write a small negative log-likelihood by hand and hand it to `optim_minimize()`.
- Fit it two ways -- global (`"de"`, seeded) and local (`"bfgs"`, from a starting guess) -- and
  see the two agree.
- Confirm the seeded `"de"` run's PARAMETERS are bit-identical to the Python twin's: the
  optimizer's candidate-generating random number stream lives in the shared C++ core, not in R
  or Python.
- `optim_maximize()`'s sign convention: it does not negate the objective's own value.

## Setup

```{r}
library(corehydror)
```

## A synthetic rating curve

A power-law stage-discharge relationship, $Q = a\,(h - h_0)^b$, is the classic hydraulic rating
curve -- discharge as a power of stage above some effective channel-control elevation $h_0$. This
is deliberately NOT the BaRatin model behind `rating_curve_analysis()` (that one fits a
matrix-of-controls addition-mode curve with its own log10-space likelihood); it is a small,
different model written from scratch to show what `optim_minimize()` is for.

Twenty-five stage readings and a synthetic "true" curve with multiplicative log-normal
measurement error -- the noise draw is seeded through the port, so it is bit-identical to the
Python twin's.

```{r}
stage <- seq(1, 10, length.out = 25)
a_true <- 15
b_true <- 1.7
h0_true <- 0.5
discharge_true <- a_true * (stage - h0_true)^b_true

noise <- dist_random(distribution("Normal", c(0, 0.05)), 25, seed = 2024)
discharge <- discharge_true * exp(noise)

cat(sprintf("First three discharges: %s\n",
            paste(sprintf("%.6f", discharge[1:3]), collapse = ", ")))
plot(stage, discharge, pch = 19, col = "#5b7a8c",
     xlab = "Stage (ft)", ylab = "Discharge (cfs)", main = "Synthetic rating curve data")
lines(stage, discharge_true, col = "#6b7f3f", lwd = 2)
legend("topleft", c("Observed (with noise)", "True curve"), pch = c(19, NA), lty = c(NA, 1),
       col = c("#5b7a8c", "#6b7f3f"), bty = "n")
```

## The likelihood

Four parameters: $a$, $b$, $h_0$, and the log-space error standard deviation $\sigma$. Assuming
independent log-normal errors, the negative log-likelihood is

$$
-\ell(a, b, h_0, \sigma) = \sum_{i=1}^{n} \left[
  \tfrac12 \log(2\pi\sigma^2) + \frac{\left(\log Q_i - \log\hat{Q}_i\right)^2}{2\sigma^2}
\right], \qquad \hat{Q}_i = a\,(h_i - h_0)^b
$$

`optim_minimize()` needs `lower`/`upper` bounds for every method, including `"de"`, which takes
no starting guess at all; `"bfgs"` additionally needs one (`initial`).

```{r}
nll <- function(p) {
  a <- p[1]; b <- p[2]; h0 <- p[3]; sigma <- p[4]
  predicted <- a * pmax(stage - h0, 1e-6)^b
  resid <- log(discharge) - log(predicted)
  -sum(dnorm(resid, mean = 0, sd = sigma, log = TRUE))
}

lower <- c(1, 0.5, 0, 0.001)
upper <- c(50, 4, 2, 1)
```

## Fit two ways

Differential evolution (`"de"`) is a global, population-based, stochastic method -- seeded here,
so it reproduces exactly on every run and, more importantly, its PARAMETERS reproduce exactly
in the Python twin (see the reproduction check for the caveat on `value` and on BFGS). BFGS is a
local, gradient-based method that needs a starting guess; `control = list(compute_hessian = FALSE)` skips the (here, unneeded) numerical Hessian
both methods compute by default.

```{r}
fit_de <- optim_minimize(nll, lower = lower, upper = upper, method = "de", seed = 2024,
                         control = list(compute_hessian = FALSE))
fit_bfgs <- optim_minimize(nll, lower = lower, upper = upper, initial = c(10, 1, 0, 0.1),
                           method = "bfgs", control = list(compute_hessian = FALSE))

print(fit_de)
print(fit_bfgs)
```

Global and local agree to about four decimal places, and both land close to the true generating
parameters ($a=15$, $b=1.7$, $h_0=0.5$, $\sigma=0.05$) -- unsurprising with a clean,
well-identified likelihood and only mild noise, but worth checking rather than assuming.

```{r}
comparison <- rbind(
  True = c(a_true, b_true, h0_true, 0.05),
  DE = fit_de$parameters,
  BFGS = fit_bfgs$parameters
)
colnames(comparison) <- c("a", "b", "h0", "sigma")
print(round(comparison, 4))
```

## `optim_maximize()`'s sign convention

Maximizing the log-likelihood directly (the negation of `nll`) with the same seed lands on the
same parameters DE found minimizing `nll` -- and `optim_maximize()`'s reported `value` is the
objective's OWN value at the optimum, not negated, so it comes out as `-fit_de$value` rather than
`fit_de$value`.

```{r}
log_likelihood <- function(p) -nll(p)
fit_de_max <- optim_maximize(log_likelihood, lower = lower, upper = upper, method = "de",
                             seed = 2024, control = list(compute_hessian = FALSE))
cat(sprintf("optim_minimize(nll)$value:        %.6f\n", fit_de$value))
cat(sprintf("optim_maximize(log_likelihood)$value: %.6f\n", fit_de_max$value))
cat(sprintf("Same parameters: %s\n", isTRUE(all.equal(fit_de$parameters, fit_de_max$parameters))))
```

## Plot: fitted curve

```{r}
h_grid <- seq(min(stage), max(stage), length.out = 200)
fitted_curve <- fit_de$parameters[1] * pmax(h_grid - fit_de$parameters[3], 0)^fit_de$parameters[2]

plot(stage, discharge, pch = 19, col = "#5b7a8c",
     xlab = "Stage (ft)", ylab = "Discharge (cfs)", main = "Fitted rating curve (DE)")
lines(h_grid, fitted_curve, col = "#b06a3b", lwd = 2)
lines(stage, discharge_true, col = "#6b7f3f", lwd = 2, lty = 2)
legend("topleft", c("Observed", "Fitted (DE)", "True curve"),
       pch = c(19, NA, NA), lty = c(NA, 1, 2), lwd = c(NA, 2, 2),
       col = c("#5b7a8c", "#b06a3b", "#6b7f3f"), bty = "n")
```

## Key takeaways

1. `optim_minimize()`/`optim_maximize()` take any R function of a numeric parameter vector --
   nothing about the model needs to be one the port already knows how to build.
2. A seeded `"de"` run's PARAMETERS are exactly reproducible, run to run and language to
   language: the PRNG lives in the shared C++ core, not in R's or Python's own random state.
   The reported objective VALUE only agrees to about 15 decimal digits, because it comes from
   re-evaluating each language's own likelihood code.
3. Global (`"de"`) and local (`"bfgs"`) methods are worth cross-checking against each other on
   the same objective, not just trusted individually.
4. `optim_maximize()`'s `value` is the objective's own value at the optimum -- never negated --
   so maximizing `log_likelihood` reports the log-likelihood itself, not its negation.

## Reproduction check

No upstream literals exist for this example (it has no upstream counterpart), so every value
below is an internal-consistency or cross-language check. One honest floating-point note, in the
same spirit as example 09's Box-Cox lambda: `discharge` and DE's `parameters` are bit-identical
across languages, because they never leave the shared, seeded C++ core (the synthetic-noise draw
and DE's own candidate-generation stream). `fit_de$value`, by contrast, comes from re-evaluating
THIS PAGE'S OWN likelihood -- R's `dnorm(..., log = TRUE)` here, a hand-written equivalent in
Python -- at that shared parameter vector, and R's and Python's floating-point libraries do not
guarantee bit-identical rounding for the same formula, so it only agrees to about 15 decimal
digits, not all 17. BFGS goes further: as a local, gradient-based method, it is sensitive enough
to those same sub-ulp differences that its OWN trajectory diverges slightly language to language
(both still land within about 1e-10 of each other and close to the truth), so its numbers below
are asserted within R only, not pinned against a Python literal.

```{r}
near <- \(x, literal, tol = 1e-15) abs(x / literal - 1) < tol

stopifnot(
  # DE and BFGS agree with each other to 1e-3 and recover parameters close to the truth.
  max(abs(fit_de$parameters - fit_bfgs$parameters)) < 1e-3,
  abs(fit_de$parameters[1] - a_true) < 0.2,
  abs(fit_de$parameters[2] - b_true) < 0.05,
  abs(fit_de$parameters[3] - h0_true) < 0.05,

  # optim_maximize()'s value is the exact negation of optim_minimize()'s on the negated objective,
  # and both land on the same parameters.
  isTRUE(all.equal(fit_de$parameters, fit_de_max$parameters)),
  near(fit_de_max$value, -fit_de$value),

  # BFGS reproduces within R (this file's own regression, not a cross-language claim -- see above).
  near(fit_bfgs$parameters[1], 15.049867097337126),
  near(fit_bfgs$parameters[2], 1.6946263072106593),
  near(fit_bfgs$parameters[3], 0.49510447626655224),
  near(fit_bfgs$parameters[4], 0.054387129725377112),
  near(fit_bfgs$value, -37.317232116709945),

  # Cross-language identity: the Python notebook asserts these same literals for `discharge` and
  # DE's `parameters` (bit-exact, see above); DE's `value` is asserted only to 1e-13 relative,
  # comfortably above the ~1e-16 drift actually measured but well below "coincidentally close".
  near(discharge[1], 4.66842623508654),
  near(discharge[2], 12.377812912604149),
  near(discharge[3], 22.499344656381929),
  near(fit_de$parameters[1], 15.050239124703497),
  near(fit_de$parameters[2], 1.6946139965788718),
  near(fit_de$parameters[3], 0.49511501008458236),
  near(fit_de$parameters[4], 0.054386727106046459),
  near(fit_de$value, -37.317232039101832, tol = 1e-13)
)
cat("All reproduction checks passed.\n")
```
