---
title: "12. Model evaluation"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

[Example 21](../21-flood-frequency/r.qmd) fits LogNormal, GEV, and Weibull to a 48-year peak-flow
record and ranks them by log-likelihood and AIC, setting the upstream script's
Kolmogorov-Smirnov ranking aside. This example picks it back up, revisiting the same record with
the whole goodness-of-fit toolbox -- `goodness_of_fit()`, `gof_test()`, `gof_rmse()`,
`aic()`/`aic_weights()` -- and confirms the KS statistic agrees with the AIC ranking. It has no upstream counterpart: the USACE-RMC Numerics-Python-Examples
repository has no dedicated model-comparison notebook.

## What you'll learn

- Fit several candidate distributions to one flood series and score each with a goodness-of-fit
  test statistic (`gof_test`), not just log-likelihood.
- Rank candidates with `aic()`/`aic_weights()`, and read an Akaike weight as "the probability
  this model is the best of the set, given the set."
- Score a fitted quantile curve against the data with `goodness_of_fit()`'s general continuous
  metrics (NSE, KGE, R-squared, ...), the same metrics used to evaluate any modeled-versus-observed
  series, not just a distribution fit.
- `classification_metrics()` for a binary comparison, illustrated (a little artificially) on an
  above/below-median split.

## Setup

```{r}
library(corehydror)
```

## The data

The same 48 years of annual peak flows (cfs) as example 21.

```{r}
peak_flows <- c(
  6290, 2700, 13100, 16900, 14600, 9600, 7740, 8490, 8130, 12000,
  17200, 15000, 12400, 6960, 6500, 5840, 10400, 18800, 21400, 22600,
  14200, 11000, 12800, 15700, 4740, 6950, 11800, 12100, 20600, 14600,
  14600, 8900, 10600, 14200, 14100, 14100, 12500, 7530, 13400, 17600,
  13400, 19200, 16900, 15500, 14500, 21900, 10400, 7460
)
n <- length(peak_flows)
cat(sprintf("n = %d peak flows, %s to %s cfs\n",
            n, format(min(peak_flows), big.mark = ","), format(max(peak_flows), big.mark = ",")))
```

## Fit four candidates

LogNormal and Weibull by maximum likelihood, GEV by L-moments (as in example 21), plus Gumbel by
maximum likelihood as a fourth candidate.

```{r}
ln  <- dist_fit("LogNormal", peak_flows, method = "mle")
gev <- dist_fit("GeneralizedExtremeValue", peak_flows, method = "lmom")
wb  <- dist_fit("Weibull", peak_flows, method = "mle")
gum <- dist_fit("Gumbel", peak_flows, method = "mle")

models <- list(LogNormal = ln, GEV = gev, Weibull = wb, Gumbel = gum)
```

## Rank with goodness-of-fit test statistics

`gof_test()` mirrors the C# `GoodnessOfFit` class's Kolmogorov-Smirnov D and Anderson-Darling
A-squared statistics for a fitted distribution against its data -- smaller is a better fit for
both. `gof_rmse()` compares the fitted quantiles at Weibull plotting positions to the sorted
data. `aic()` needs only the parameter count and the log-likelihood `dist_log_likelihood()`
already provides; `aic_weights()` turns a vector of AIC values into weights that sum to one, each
reading as "the (relative) probability this is the best model of the set."

`aic()`/`gof_test()`/`gof_rmse()` take one model at a time, so `mapply`/`sapply` build the table
one candidate at a time -- there is no vectorized "rank these four" verb, deliberately: the
toolbox mirrors the C# statics one call at a time, and ranking is a few lines of R on top.

```{r}
k <- sapply(models, function(d) length(dist_params(d)))
log_lik <- sapply(models, function(d) dist_log_likelihood(d, peak_flows))
AICs <- mapply(aic, k, log_lik)
ks <- sapply(models, function(d) gof_test(peak_flows, d, test = "ks"))
ad <- sapply(models, function(d) gof_test(peak_flows, d, test = "ad"))
rmse_v <- sapply(models, function(d) gof_rmse(peak_flows, d))
weights <- aic_weights(AICs)

ranking <- data.frame(
  Model = names(models), k = k, LogLik = log_lik, AIC = AICs,
  KS = ks, AD = ad, RMSE = rmse_v, AkaikeWeight = weights,
  row.names = NULL
)
ranking <- ranking[order(ranking$AIC), ]
print(ranking, digits = 6)
```

Weibull wins on AIC, KS, and RMSE, carrying 71% of the Akaike weight -- confirming, with an
actual goodness-of-fit test statistic this time, exactly what example 21 found by log-likelihood
alone: "Weibull edges out GEV, and LogNormal trails." Anderson-Darling is the one holdout: it
weights tail deviations more heavily than KS, and by that measure GEV's A-squared (0.2506) edges
Weibull's (0.2518) -- the two are close enough, and the metrics disagree closely enough, that this
is a real illustration of why a ranking should look at more than one statistic rather than an
example built to make every metric agree.

## Score the winning quantile curve

`goodness_of_fit()` is the general-purpose continuous-metrics verb: any modeled-versus-observed
pair, not just a fitted distribution. Comparing the Weibull quantile curve at the data's own
plotting positions to the sorted observations treats the frequency curve itself as "the model"
and scores it with NSE, KGE, R-squared and friends -- the standard hydrologic model-evaluation
metrics, repurposed here for a frequency curve.

```{r}
pp <- plotting_positions(n)
sorted_flows <- sort(peak_flows)
modeled <- dist_quantile(wb, pp)

fit_scores <- goodness_of_fit(sorted_flows, modeled)
print(round(fit_scores, 4))
```

NSE above 0.98 and KGE above 0.93 both say the same thing the AIC ranking already said: the
Weibull quantile curve tracks the sorted record closely apart from the usual upper-tail spread.

## A binary comparison: `classification_metrics()`

`classification_metrics()` compares two already-binary label vectors -- it has no threshold
argument, in R or in C#, so the labels are built by hand first. This is a contrived use for a
frequency curve (nothing about flood-frequency analysis is naturally binary), but it is the only
public verb that exposes the C# `GoodnessOfFit` classification statics, so it is worth seeing
once: label each year "above the median flow" or not, in both the observed record and the fitted
quantile curve at the same rank, and score the agreement.

```{r}
med <- median(peak_flows)
obs_label <- as.numeric(sorted_flows > med)
mod_label <- as.numeric(modeled > med)
cls <- classification_metrics(obs_label, mod_label)
print(round(cls, 4))
```

## Plot: fitted curves against plotting positions

```{r}
colors <- c(LogNormal = "#6b7f3f", GEV = "#b06a3b", Weibull = "#5b7a8c", Gumbel = "#8c8c7a")
x_grid <- seq(min(peak_flows) * 0.8, max(peak_flows) * 1.1, length.out = 500)

plot(sorted_flows, pp,
  pch = 19, col = "black", log = "x",
  xlab = "Peak flow (cfs)", ylab = "Non-exceedance probability",
  main = "Fitted frequency curves vs plotting positions (Weibull plotting positions)"
)
for (name in names(models)) {
  lines(x_grid, dist_cdf(models[[name]], x_grid), col = colors[[name]], lwd = 2)
}
legend("bottomright",
  legend = c("Data (plotting positions)", sprintf("%s (AIC %.1f)", ranking$Model, ranking$AIC)),
  pch = c(19, NA, NA, NA, NA), lty = c(NA, 1, 1, 1, 1), lwd = c(NA, 2, 2, 2, 2),
  col = c("black", colors[ranking$Model]), bty = "n"
)
```

```{r}
barplot(weights[order(AICs)],
  names.arg = names(models)[order(AICs)],
  col = colors[names(models)[order(AICs)]], border = "white",
  ylab = "Akaike weight", main = "Model weights (aic_weights())"
)
```

## Key takeaways

1. `gof_test()` and `gof_rmse()` give a fitted distribution the same KS/AD/RMSE ranking the C#
   `GoodnessOfFit` class always offered, which is what example 21 sets aside.
2. `aic_weights()` turns a table of AIC values into normalized weights -- useful directly, and
   the same weights `fit_distributions()`'s composite-model machinery consumes internally.
3. `goodness_of_fit()` is not distribution-specific: it scores any two equal-length numeric
   series, so it works as well on a frequency curve as on a rainfall-runoff hydrograph.
4. `classification_metrics()` fills out the C# `GoodnessOfFit` surface for binary comparisons,
   even though flood-frequency analysis rarely needs one.

## Reproduction check

No upstream literals exist for this example (it has no upstream counterpart), so every value
below is an internal-consistency and cross-language check: the ranking is self-consistent (AIC
order matches the KS order here, and matches the AD order everywhere except the top two, as
discussed above), and the literals are the exact values the Python notebook prints, asserted at
1e-15 relative tolerance (R's decimal parser can land one ulp away from the written literal; the
seeded/deterministic values themselves are bit-exact across languages).

```{r}
near <- \(x, literal) abs(x / literal - 1) < 1e-15

stopifnot(
  # Ranking is self-consistent: AIC order (Weibull, GEV, Gumbel, LogNormal) matches the KS order
  # over the same four models; AD agrees except it swaps the top two (see the prose above).
  identical(order(AICs), order(ks)),
  identical(order(AICs)[3:4], order(ad)[3:4]),
  setequal(order(AICs)[1:2], order(ad)[1:2]),
  ranking$Model[1] == "Weibull",
  sum(weights) - 1 < 1e-12,

  # Cross-language identity: the Python notebook asserts these same literals.
  near(AICs[["LogNormal"]], 958.30496639602507),
  near(AICs[["GEV"]], 952.44533103738274),
  near(AICs[["Weibull"]], 950.09168256535804),
  near(AICs[["Gumbel"]], 954.91458850922629),
  near(ks[["Weibull"]], 0.06638744856290868),
  near(ad[["Weibull"]], 0.2518324119771691),
  near(rmse_v[["Weibull"]], 547.45273890460226),
  near(weights[3], 0.7070115798971689), # Weibull's own weight, position 3 in `models`
  near(fit_scores[["nse"]], 0.985895395879688),
  near(fit_scores[["kge"]], 0.93549912169381888),
  near(fit_scores[["r_squared"]], 0.98937885175430651),
  near(fit_scores[["pearson"]], 0.99467524939263796),
  near(cls[["accuracy"]], 95.833333333333329)
)
cat("All reproduction checks passed.\n")
```
