---
title: "26. Copulas and joint frequency"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

A reservoir cares about two things at once. The peak inflow sizes the spillway; the runoff volume
sizes the storage. Fitting each record separately says how often each is exceeded and says
nothing about how often they arrive together. A copula answers that second question: it holds the
two marginal distributions fixed and models only the dependence between them.

`corehydror` exposes the seven bivariate copulas of the Numerics library through `copula()` and
`copula_fit()`, and the multivariate distributions through `mvdist_normal()` and its siblings.
This example fits a copula to a paired peak-and-volume record, prices a coincident design event,
and checks the answer against the model-based `bivariate_analysis()` path.

## What you'll learn

- Fit a copula's dependence parameter three ways: maximum pseudo-likelihood, Kendall's tau
  inversion, and inference from margins.
- Read tail dependence, and watch two copulas that fit the body of a sample almost equally well
  disagree about its joint tail.
- Compute the and-joint exceedance probability of a coincident peak-and-volume event, and set it
  against the answer independence would give.
- Recover the same probability from a bivariate normal, and cross-check the whole thing against
  `bivariate_analysis()`.

## Setup

```{r}
library(corehydror)
```

## Peak and volume pairs

Forty-eight water years. The peaks are the annual peak-flow record used in
[example 21](../21-flood-frequency/r.qmd) and [example 23](../23-censored-flood-frequency/r.qmd);
the volumes are a companion record written for this example, so the page is self-contained and
both languages read exactly the same numbers.

```{r}
peak <- c(
  6290, 2700, 13100, 16900, 14600, 9600, 7740, 8490, 8130, 12000,
  17200, 15000, 12400, 6960, 6500, 5840, 10400, 18800, 21400, 22600,
  14200, 11000, 12800, 15700, 4740, 6950, 11800, 12100, 20600, 14600,
  14600, 8900, 10600, 14200, 14100, 14100, 12500, 7530, 13400, 17600,
  13400, 19200, 16900, 15500, 14500, 21900, 10400, 7460
)
volume <- c(
  7.21, 3.21, 11.88, 10.70, 9.07, 5.31, 7.04, 5.33, 6.18, 9.54,
  9.68, 7.66, 8.99, 5.95, 5.37, 6.45, 7.00, 13.71, 12.75, 18.27,
  11.54, 8.19, 11.15, 10.35, 3.00, 5.67, 6.71, 8.47, 12.02, 12.23,
  10.71, 9.58, 9.82, 8.90, 8.80, 8.55, 8.48, 3.50, 6.60, 16.79,
  7.39, 12.95, 11.09, 7.31, 11.50, 16.79, 8.05, 8.56
)
cat(sprintf(
  "%d water years; peak %s to %s cfs, volume %.2f to %.2f thousand acre-feet\n",
  length(peak), format(min(peak), big.mark = ","), format(max(peak), big.mark = ","),
  min(volume), max(volume)
))
```

The design event this page prices is a 20,000 cfs peak arriving with a 12 thousand acre-foot
volume.

```{r}
#| fig-alt: "Scatter of annual peak flow against runoff volume on logarithmic axes, with the two design thresholds drawn as dashed lines."
q_peak <- 20000
q_vol <- 12

plot(peak, volume,
  log = "xy", pch = 16, col = "#6b7f3f",
  xlab = "Annual peak flow (cfs)", ylab = "Runoff volume (thousand acre-feet)",
  main = "Paired peak and volume, 48 water years"
)
abline(v = q_peak, h = q_vol, lty = 2, col = "#8c5a3b")
```

## Dependence without marginals

`copula_fit(family, x, y, method = "mpl")` maximizes the pseudo-likelihood. That likelihood is
defined on the plotting positions `rank / (n + 1)` rather than on the data scale, so it never
touches the marginal distributions. Pass the raw paired observations: the ranking happens inside
the shared C++ core, which is why R and Python fit the same number from the same input.

```{r}
families <- c("Clayton", "Frank", "Gumbel", "Joe", "Normal")
mpl <- lapply(families, \(f) copula_fit(f, peak, volume, method = "mpl"))
names(mpl) <- families

ranking <- data.frame(
  n_parameters = sapply(mpl, \(cop) length(copula_params(cop))),
  theta = sapply(mpl, \(cop) cop$theta),
  pseudo_loglik = sapply(mpl, \(cop) copula_log_likelihood(cop, peak, volume, method = "pseudo"))
)
round(ranking[order(-ranking$pseudo_loglik), ], 4)
```

All five carry one dependence parameter and all five are evaluated on the same pseudo
observations, so their pseudo log-likelihoods rank them directly. The ranking barely separates
the top two: the Normal copula leads, the Gumbel follows, and both stand well clear of the other
three. Both are carried forward below, because they turn out to disagree about exactly the part
of the sample the design question asks about.

`method = "tau"` inverts Kendall's tau into the dependence parameter with no likelihood
optimization at all. Upstream implements that inversion for Clayton, Gumbel, and AliMikhailHaq
only, closed form for the first two and a root solve for the third, so asking any other family
for it is an error rather than a silent fallback.

```{r}
round(c(
  Clayton = copula_fit("Clayton", peak, volume, method = "tau")$theta,
  Gumbel = copula_fit("Gumbel", peak, volume, method = "tau")$theta
), 4)

cat(tryCatch(copula_fit("Frank", peak, volume, method = "tau"),
             error = conditionMessage), "\n")
```

## Inference from margins

Inference from margins fits each marginal by maximum likelihood first, then estimates the copula
with those marginals held fixed. Naming a family in `margin_x` or `margin_y` is what asks for
that first step; under `"ifm"` an already-parameterized `distribution()` object is attached as
given and skips the fit, while `method = "mle"` re-estimates both marginals jointly with the
dependence parameter. A named marginal is rejected under `"mpl"` and `"tau"`, since neither
method looks at marginals and the name would come back unfitted.

```{r}
gumbel <- copula_fit("Gumbel", peak, volume, method = "ifm",
                     margin_x = "LogNormal", margin_y = "LogNormal")
normal <- copula_fit("Normal", peak, volume, method = "ifm",
                     margin_x = "LogNormal", margin_y = "LogNormal")

margins <- rbind(peak = dist_params(gumbel$margin_x), volume = dist_params(gumbel$margin_y))
round(margins, 4)
```

`LogNormal` in this library is base 10: the two parameters are the mean and standard deviation of
the base-10 logarithm of the variable, the same log space a Bulletin 17C fit reports.

```{r}
cmp <- data.frame(
  theta_mpl = c(Gumbel = mpl$Gumbel$theta, Normal = mpl$Normal$theta),
  theta_ifm = c(gumbel$theta, normal$theta),
  ifm_loglik = c(copula_log_likelihood(gumbel, peak, volume, method = "ifm"),
                 copula_log_likelihood(normal, peak, volume, method = "ifm"))
)
round(cmp, 4)
```

The two estimates of the same parameter differ because they are fitted against different things.
The pseudo-likelihood sees only ranks, so no marginal can influence it. Inference from margins
sees the fitted marginal CDFs, and any misfit there passes straight into the copula. Both are
legitimate; neither is a refinement of the other.

## Tail dependence

The tail dependence coefficients are the limiting probability that one variable is extreme given
that the other is.

```{r}
round(t(sapply(list(Gumbel = gumbel, Normal = normal), copula_tail_dependence)), 4)
```

Neither has lower tail dependence. In the upper tail they could not be further apart. The Gumbel
copula's coefficient is `2 - 2^(1/theta)`, positive for every `theta` above its lower bound of 1,
while the Normal copula returns zero for every correlation. Two copulas that scored within a
fraction of a log-likelihood unit of each other on the body of the sample make opposite
structural claims about its corner.

## Pricing a coincident event

`copula_exceedance()` returns the and-joint exceedance probability `P(U > u, V > v)`, which the
core computes as `1 - u - v + C(u, v)`, or the union probability `1 - C(u, v)` with
`type = "or"`. Both take non-exceedance probabilities, so run the design thresholds through the
fitted marginal CDFs first.

```{r}
u <- dist_cdf(gumbel$margin_x, q_peak)
v <- dist_cdf(gumbel$margin_y, q_vol)

joint <- data.frame(
  p_and = c(
    `Gumbel copula` = copula_exceedance(gumbel, u, v, "and"),
    `Normal copula` = copula_exceedance(normal, u, v, "and"),
    Independence = (1 - u) * (1 - v)
  )
)
joint$return_period <- 1 / joint$p_and
joint$vs_independence <- joint$p_and / joint["Independence", "p_and"]
round(joint, 4)

cat(sprintf(
  "peak alone: p = %.4f (1 in %.1f);  volume alone: p = %.4f (1 in %.1f)\n",
  1 - u, 1 / (1 - u), 1 - v, 1 / (1 - v)
))
cat(sprintf("either one exceeded, Gumbel copula: p = %.4f\n",
            copula_exceedance(gumbel, u, v, "or")))
```

Treating the peak and the volume as independent multiplies their two exceedance probabilities and
returns 0.0194. The fitted copulas put the same event at 0.0943 under the Gumbel and 0.0853 under
the Normal, and the `vs_independence` column prices the mistake: independence understates how
often the two arrive together by a factor of 4.85 or 4.39, depending on the copula. Independence
is not the conservative assumption here.

The tail dependence coefficients above are limits, so their effect shows up when both variables
are pushed out together. Walk the two thresholds along the diagonal, holding each marginal at the
same annual exceedance probability, and the two copulas separate.

```{r}
aep <- c(0.2, 0.1, 0.02, 0.01, 0.002)
corner <- data.frame(
  marginal_aep = aep,
  gumbel = sapply(1 - aep, \(a) copula_exceedance(gumbel, a, a, "and")),
  normal = sapply(1 - aep, \(a) copula_exceedance(normal, a, a, "and")),
  independence = aep^2
)
corner$gumbel_over_normal <- corner$gumbel / corner$normal
round(corner, 6)
```

The Gumbel-to-Normal ratio grows from 1.0915 at a 0.2 marginal exceedance probability to 1.9930
at 0.002: upper tail dependence hardly matters in the body of the distribution and nearly doubles
the joint probability in its corner. Independence falls away far faster than either.

```{r}
#| fig-alt: "And-joint exceedance probability against common marginal annual exceedance probability on logarithmic axes: the Gumbel copula solid, the Normal copula dashed, and independence dotted."
grid_aep <- 10^seq(log10(0.5), log10(0.001), length.out = 60)
g_curve <- sapply(1 - grid_aep, \(a) copula_exceedance(gumbel, a, a, "and"))
n_curve <- sapply(1 - grid_aep, \(a) copula_exceedance(normal, a, a, "and"))

plot(grid_aep, g_curve,
  type = "l", log = "xy", lwd = 2, col = "#8c5a3b",
  ylim = range(g_curve, n_curve, grid_aep^2),
  xlab = "Marginal annual exceedance probability (each variable)",
  ylab = "P(both exceeded)",
  main = "Joint exceedance along the diagonal"
)
lines(grid_aep, n_curve, lwd = 2, lty = 2, col = "#6b7f3f")
lines(grid_aep, grid_aep^2, lwd = 2, lty = 3, col = "#4d6b8a")
legend("bottomright",
  legend = c("Gumbel copula", "Normal copula", "Independence"),
  col = c("#8c5a3b", "#6b7f3f", "#4d6b8a"), lwd = 2, lty = c(1, 2, 3), bty = "n"
)
```

## The Normal copula is the bivariate lognormal

A Normal copula with lognormal marginals is a bivariate normal in log space, so the same joint
probability is available from `mvdist_normal()` as a rectangle probability. `mvdist_interval()`
returns `P(lower <= X <= upper)` through the ported Genz algorithm, and the two routes have to
agree.

```{r}
px <- unname(dist_params(normal$margin_x))
py <- unname(dist_params(normal$margin_y))
rho <- normal$theta
covar <- matrix(c(px[2]^2, rho * px[2] * py[2],
                  rho * px[2] * py[2], py[2]^2), nrow = 2)

mvn <- mvdist_normal(mean = c(px[1], py[1]), covariance = covar, seed = 20250812)
rect <- mvdist_interval(mvn, c(log10(q_peak), log10(q_vol)), c(1e6, 1e6))
p_norm_copula <- copula_exceedance(normal, u, v, "and")

cat(sprintf("Normal copula   %.15f\nMVN rectangle   %.15f\nrelative difference %.1e\n",
            p_norm_copula, rect, rect / p_norm_copula - 1))
```

`mvdist_normal()` takes a `seed` because the Genz integrator draws from its own Mersenne
Twister. Leave it out and the instance is clock-seeded, so any value that depends on those draws
stops being reproducible run to run, let alone across languages. `mvdist_cdf()` avoids the
integrator entirely at dimension one and two, where the ported code uses closed forms.

The same object answers the conditional question directly.

```{r}
cond <- mvdist_conditional(mvn, given = 1, values = log10(q_peak))
cond_mean <- mvdist_mean(cond)
cond_sd <- sqrt(mvdist_covariance(cond)[1, 1])
p_cond <- 1 - dist_cdf(distribution("Normal", c(cond_mean, cond_sd)), log10(q_vol))

cat(sprintf(
  "log10 volume given a 20,000 cfs peak: mean %.4f, sd %.4f\nP(volume > 12 kaf | peak = 20,000 cfs) = %.4f\n",
  cond_mean, cond_sd, p_cond
))
```

`mvdist_marginal()`, `mvdist_conditional()`, and `mvdist_interval()` are `MultivariateNormal`
methods only; `MultivariateStudentT` has no upstream counterpart for any of the three.

## Cross-check against the model path

`bivariate_analysis()` reaches the same copula from the other direction. It wraps the two fixed
marginals and the copula in a model, estimates the dependence parameter with a Bayesian MCMC, and
returns a credible band alongside the point curve.

```{r}
grid_x <- c(15000, 20000, 25000)
grid_y <- c(9, 12, 15)

ba <- bivariate_analysis(
  "LogNormal", peak, px, "LogNormal", volume, py,
  xy_x = grid_x, xy_y = grid_y, copula = "Normal",
  estimation_method = "InferenceFromMargins", sampler = "DEMCz",
  iterations = 1000, output_length = 4000, seed = 20250812,
  number_of_chains = 4, thinning_interval = 1
)

map_copula <- copula("Normal", ba$parameters[1])
gu <- dist_cdf(normal$margin_x, grid_x)
gv <- dist_cdf(normal$margin_y, grid_y)
at_map <- mapply(\(a, b) copula_exceedance(map_copula, a, b, "and"), gu, gv)

cross <- data.frame(
  peak = grid_x, volume = grid_y,
  analysis_curve = ba$mode_curve,
  analysis_lower = ba$lower_ci, analysis_upper = ba$upper_ci,
  copula_at_map = at_map
)
round(cross, 5)

cat(sprintf("theta: %.6f from copula_fit(ifm), %.6f reported by the analysis\n",
            normal$theta, ba$parameters[1]))
```

The two paths agree on the dependence parameter to four decimal places, and the copula verb's
probability sits inside the analysis's own credible band at all three ordinates. They are not
identical, and the reason is worth knowing: `bivariate_analysis()` reports the posterior maximum
in `parameters`, but builds its curve at whatever point estimator the underlying Bayesian
analysis carries, which defaults to the posterior mean. Evaluating `copula_exceedance()` at the
reported maximum therefore answers a slightly different question than the curve does.

## Reproduction check

```{r}
# Every fit on this page is deterministic (the copula estimators are optimizers over a fixed
# sample; the Bayesian cross-check is seeded), so these literals are what this port computes and
# what the Python notebook asserts -- which is what proves the cross-language identity. The
# comparisons carry a 1e-15 relative tolerance only because R's decimal parser can land one ulp
# away from the written literal; the fixture suite is what enforces bit-exactness.
near <- \(x, literal) abs(x / literal - 1) < 1e-15
stopifnot(
  # Maximum pseudo-likelihood across five families.
  near(mpl$Clayton$theta, 2.3063313619874619),
  near(mpl$Frank$theta, 8.6716321936696286),
  near(mpl$Gumbel$theta, 2.8147142423362927),
  near(mpl$Joe$theta, 3.5522244465492729),
  near(mpl$Normal$theta, 0.8652104567754928),
  near(ranking["Normal", "pseudo_loglik"], 30.370210262642061),
  near(ranking["Gumbel", "pseudo_loglik"], 30.043841011715617),

  # Kendall's tau inversion.
  near(copula_fit("Clayton", peak, volume, method = "tau")$theta, 3.4473182880894253),
  near(copula_fit("Gumbel", peak, volume, method = "tau")$theta, 2.7236591440447127),

  # Inference from margins: the marginals and the two copulas.
  near(margins["peak", 1], 4.0674283057469083),
  near(margins["peak", 2], 0.18679432530177564),
  near(margins["volume", 1], 0.92776174913312093),
  near(margins["volume", 2], 0.16835824505465125),
  near(gumbel$theta, 2.7122936851673694),
  near(normal$theta, 0.84268973641367872),
  near(cmp["Gumbel", "ifm_loglik"], 29.900498224775983),
  near(cmp["Normal", "ifm_loglik"], 29.718349666163725),

  # Tail dependence.
  copula_tail_dependence(gumbel)[["lower"]] == 0,
  near(copula_tail_dependence(gumbel)[["upper"]], 0.70881865711981251),
  copula_tail_dependence(normal)[["lower"]] == 0,
  copula_tail_dependence(normal)[["upper"]] == 0,

  # The coincident design event.
  near(u, 0.89445655684934455),
  near(v, 0.81577716213041274),
  near(joint["Gumbel copula", "p_and"], 0.09432008642998968),
  near(joint["Normal copula", "p_and"], 0.085332736299363332),
  near(joint["Independence", "p_and"], 0.019443512615741197),
  near(copula_exceedance(gumbel, u, v, "or"), 0.19544619459025303),

  # The joint corner: both marginals held at the same exceedance probability.
  near(corner$gumbel[2], 0.072808127874079531),
  near(corner$normal[2], 0.061085897213300933),
  near(corner$gumbel[5], 0.001418389605702286),
  near(corner$normal[5], 0.00071170053465230421),
  near(corner$gumbel_over_normal[1], 1.0915335165887639),
  near(corner$gumbel_over_normal[5], 1.9929584658738935),

  # The bivariate-normal route to the same probability.
  near(rect, 0.08533273629936243),
  near(cond_mean, 1.1051865785007626),
  near(cond_sd, 0.090643992854258601),
  near(p_cond, 0.61290374236081369),

  # The seeded bivariate_analysis() cross-check.
  near(ba$parameters[1], 0.84268212806480935),
  near(ba$mode_curve[2], 0.083618465581119183),
  near(ba$lower_ci[2], 0.075962830775020973),
  near(ba$upper_ci[2], 0.090246696612235722),
  near(at_map[2], 0.085331748836733334)
)

# Internal consistency: the and-joint identity the core computes, the ordering of the three
# joint-probability answers, and the copula verb landing inside the analysis's credible band.
stopifnot(
  abs(copula_exceedance(normal, u, v, "and") - (1 - u - v + copula_cdf(normal, u, v))) < 1e-15,
  abs(rect / p_norm_copula - 1) < 1e-12,
  joint["Gumbel copula", "p_and"] > joint["Normal copula", "p_and"],
  joint["Normal copula", "p_and"] > joint["Independence", "p_and"],
  all(diff(corner$gumbel_over_normal) > 0),
  all(corner$gumbel > corner$normal), all(corner$normal > corner$independence),
  all(cross$copula_at_map > cross$analysis_lower),
  all(cross$copula_at_map < cross$analysis_upper),
  all(diff(ba$mode_curve) < 0)
)
cat("All reproduction checks passed.\n")
```
