---
title: "25. Estimation methods"
---

::: {.callout-note appearance="minimal" icon=false}
Language: **R** (Quarto) - [Python version](python.ipynb)
:::

`bulletin17c_analysis()` and `univariate_analysis()` are convenience wrappers: pick a model,
pick a method, get a frequency curve. Underneath them sits a smaller estimation layer that fits
a model directly and hands back the fit itself: `fit_mle()`, `fit_map()`, `fit_bayesian()`, and
`fit_gmm()`. This example fits the same model three different ways over the same censored record
as [example 23](../23-censored-flood-frequency/r.qmd), and compares what each estimator returns.

## What you'll learn

- Fit one model three ways, by maximum likelihood, maximum a posteriori, and a Bayesian MCMC
  chain, and see how an informative prior moves the maximum a posteriori estimate.
- Compare AIC, BIC, and DIC across estimators, and why a method-of-moments fit like `fit_gmm()`
  reports none of the three.
- Set profile-likelihood confidence intervals against posterior credible intervals.
- Read a Bayesian fit's R-hat and effective sample size.

## Setup

```{r}
library(corehydror)
```

## The censored record

The same gauge record and historical information as example 23: forty-eight years of annual
peaks, two pre-gauge floods known only within a range, and a fifty-year perception threshold
that makes those two floods usable at all. See that example for the full derivation; here the
frame is assembled directly.

```{r}
peak_flows <- c(
  6290, 2700, 13100, 16900, 14600, 9600, 7740, 8490, 8130, 12000,
  17200, 15000, 12400, 6960, 6500, 5840, 10400, 18800, 21400, 22600,
  14200, 11000, 12800, 15700, 4740, 6950, 11800, 12100, 20600, 14600,
  14600, 8900, 10600, 14200, 14100, 14100, 12500, 7530, 13400, 17600,
  13400, 19200, 16900, 15500, 14500, 21900, 10400, 7460
)
n <- length(peak_flows)

historical <- data.frame(
  index = c(n, n + 1),
  lower = c(30000, 26000),
  value = c(34000, 29000),
  upper = c(38000, 32000)
)
perception <- data.frame(
  start_index = n, end_index = n + 49, value = 25000, number_above = 2
)
flood_data <- analysis_data(exact = peak_flows, interval = historical, threshold = perception)
```

## An informative prior on skew

`model_univariate()` defaults to flat priors within each parameter's bounds, and the mode of a
flat-times-likelihood posterior is just the likelihood's own mode. A `fit_map()` fit with no
explicit prior would then land exactly on `fit_mle()`'s estimate, leaving nothing to compare.
Log-Pearson III's skew is exactly where a Bulletin 17C analysis leans on outside
information (a regional skew estimate, in practice), so this model gives it a weakly informative
prior instead of a flat one.

```{r}
skew_prior <- distribution("Normal", c(-0.3, 0.5))
m <- model_univariate(
  "LogPearsonTypeIII", flood_data,
  parameters = model_parameter("skew", prior = skew_prior)
)
```

## Three fits of the same model

```{r}
f_mle <- fit_mle(m)
f_map <- fit_map(m)
f_bayes <- fit_bayesian(
  m,
  sampler = "DEMCzs", iterations = 400, output_length = 2000,
  seed = 20250811, thinning_interval = 1
)

params <- rbind(MLE = coef(f_mle), MAP = coef(f_map), Bayesian = coef(f_bayes))
round(params, 4)
```

Every `corehydro_fit`, regardless of method, carries a `converged` flag and a `status` string.
Here both are structurally `TRUE`/`"Success"`: the estimator throws on a genuine failure to
estimate rather than handing back a failed fit, so these two fields record that the run
completed, not that it necessarily found a good optimum.

The skew estimate moves from -0.4657 at MLE toward the prior's mean of -0.3 once a prior is in
play: -0.4189 at MAP, and -0.4263 for the Bayesian fit's own point estimate (a mode-like point
tracked by the sampler, not the posterior mean reported below in `$summary`). Both parameter
priors and censored data pull in the same direction here, pulling the log-Pearson III skew
away from the strongly negative value example 23 found for the systematic-only record.

## Goodness of fit: AIC, BIC, DIC

```{r}
gof <- data.frame(
  method = c("MLE", "MAP", "Bayesian"),
  aic = c(f_mle$aic, f_map$aic, f_bayes$aic),
  bic = c(f_mle$bic, f_map$bic, f_bayes$bic),
  dic = c(NA, NA, f_bayes$dic)
)
gof
```

AIC and BIC are defined at a single point estimate, so all three fits report them; DIC is built
from the posterior draws themselves and only `fit_bayesian()` has those, hence the two `NA`s.
MAP's AIC is worse than MLE's: the prior trades off some data likelihood at its point estimate in
exchange for consistency with the outside skew information, and AIC has no way to credit that
trade. It only sees the likelihood. DIC, which does account for the posterior's effective
complexity, is close to but not equal to the Bayesian fit's own AIC/BIC.

Not shown in this table: a `fit_gmm()` fit's `aic`/`bic`/log-likelihood are reported as `NA` too,
for a different reason. GMM is method-of-moments; it never had a likelihood surface to compute
them from in the first place, so they are not "missing" the way a `NULL` DIC on an MLE fit is.
They are structurally absent. See the GMM section below.

## Profile intervals versus credible intervals

`confint()` dispatches on the fit's own method: MLE and MAP get profile-likelihood confidence
intervals, and a Bayesian fit gets its posterior credible interval.

```{r}
ci <- rbind(
  `MLE (profile)` = confint(f_mle)["Skew (of log) (γ)", ],
  `MAP (profile)` = confint(f_map)["Skew (of log) (γ)", ],
  `Bayesian (credible)` = confint(f_bayes)["Skew (of log) (γ)", ]
)
round(ci, 4)
```

The Bayesian credible interval turns out to be the widest of the three, about 4% wider than the
MLE profile interval, and MAP's is the narrowest, pulled in by the same prior that shifted its
point estimate. All three straddle zero: even with the historical record and, for MAP, an
informative prior, this fit cannot rule out a skew near zero. That ordering is not a sign the
Bayesian fit is less certain than the others. A profile interval comes from the curvature of the
likelihood or posterior around a single point estimate, while a credible interval is read
directly off the spread of the sampled posterior draws, and the two need not land on the same
width. They also answer different questions to begin with: a probability statement about the
parameter given the data and prior, versus a repeated-sampling statement about the estimator.
Close numeric agreement between them, in either direction, was never guaranteed.

## R-hat and effective sample size

```{r}
round(f_bayes$summary[, c("rhat", "ess")], 3)
```

R-hat compares between-chain and within-chain variance across the four parallel chains
`fit_bayesian()` ran; values near 1 indicate the chains agree on where the posterior mass is.
Effective sample size (ESS) is smaller than the raw `output_length = 2000` retained draws because
consecutive MCMC draws are autocorrelated. An ESS in the low hundreds means each parameter's
posterior is effectively summarized by that many independent draws, not 2000.

## A different estimator entirely: GMM

`fit_gmm()` only fits a `model_bulletin17c()` model, since `IGMMModel` has exactly one
implementation, so it is not a fourth entry in the table above. It is, however, the estimator
`bulletin17c_analysis()` itself builds on.

```{r}
f_gmm <- fit_gmm(model_bulletin17c(flood_data))
gmm_coef <- setNames(coef(f_gmm), c("mean (log10)", "sd (log10)", "skew"))
round(gmm_coef, 4)
cat(sprintf(
  "j-statistic %.4f, p-value %s (always NA: Bulletin 17C is always just-identified)\n",
  f_gmm$j_stat, format(f_gmm$j_stat_pval)
))
```

The J-statistic is GMM's own goodness-of-fit diagnostic, an overidentification test. It is the
method-of-moments analogue of a likelihood-based AIC/BIC/DIC, not a value on the same scale as
any of them, and Bulletin 17C's moment conditions exactly match its parameter count, so there is
no over-identified case here for the p-value to describe.

`quantile_variance()` gives the delta-method variance of a fitted quantile off a GMM fit's
sandwich covariance, useful directly without going through `bulletin17c_analysis()`'s full
uncertainty-quantification machinery:

```{r}
quantile_variance(f_gmm, 0.01)
```

## Diagnostics off the MAP fit

`fit_diagnostics()` computes Cook's distance and leverage at a point-estimate fit's optimum
(MAP or GMM), or PSIS-LOO Pareto-k and prior influence off a Bayesian fit's posterior.

```{r}
diag <- fit_diagnostics(f_map)
round(head(data.frame(cooks_distance = diag$cooks_distance, leverage = diag$leverage), 5), 4)
cat(sprintf(
  "largest Cook's distance: observation %d, a %s cfs peak\n",
  which.max(diag$cooks_distance),
  format(peak_flows[which.max(diag$cooks_distance)], big.mark = ",")
))
```

The single largest Cook's distance in the record belongs to the second systematic year, 2,700
cfs, an unusually small peak that pulls a log-space fit the way example 23's Multiple
Grubbs-Beck low-outlier screen exists to guard against.

## Reproduction check

```{r}
# The pipeline is deterministic (MLE/MAP by a deterministic optimizer; the Bayesian fit is
# seeded), so these literals are what this port computes and what the Python notebook asserts --
# which is what proves the cross-language identity. The comparisons carry a 1e-15 relative
# tolerance only because R's decimal parser can land one ulp away from the written literal; the
# fixture suite is what enforces bit-exactness.
near <- \(x, literal) abs(x / literal - 1) < 1e-15
stopifnot(
  # MLE, MAP, and Bayesian fits of the same model.
  near(f_mle$parameters[1], 4.0843461309395188),
  near(f_mle$parameters[2], 0.19914759083044403),
  near(f_mle$parameters[3], -0.46572405672529588),
  near(f_mle$aic, 990.14007771522097),
  near(f_mle$bic, 995.93555461339395),

  near(f_map$parameters[1], 4.0852361248526048),
  near(f_map$parameters[2], 0.19616835592097398),
  near(f_map$parameters[3], -0.41891886262810352),
  near(f_map$aic, 992.39731155402353),
  near(f_map$bic, 998.19278845219651),

  near(f_bayes$parameters[1], 4.076977143876432),
  near(f_bayes$parameters[2], 0.20630009688354539),
  near(f_bayes$parameters[3], -0.42634285952387641),
  near(f_bayes$aic, 992.67913589176896),
  near(f_bayes$bic, 998.47461278994194),
  near(f_bayes$dic, 989.76032492563218),

  # Profile versus credible intervals on skew.
  near(ci["MLE (profile)", "lower"], -0.82156954677035599),
  near(ci["MLE (profile)", "upper"], 0.083539324732687284),
  near(ci["MAP (profile)", "lower"], -0.77253827944035625),
  near(ci["MAP (profile)", "upper"], 0.053143017857846427),
  near(ci["Bayesian (credible)", "lower"], -0.86792864849208073),
  near(ci["Bayesian (credible)", "upper"], 0.073036977766799502),

  # R-hat and ESS.
  near(f_bayes$summary["Mean (of log) (µ)", "rhat"], 1.0112377055613953),
  near(f_bayes$summary["Std Dev (of log) (σ)", "rhat"], 1.1271063062090425),
  near(f_bayes$summary["Skew (of log) (γ)", "rhat"], 1.0830639682776986),
  near(f_bayes$summary["Mean (of log) (µ)", "ess"], 126.17628141940655),
  near(f_bayes$summary["Std Dev (of log) (σ)", "ess"], 142.53659604889492),
  near(f_bayes$summary["Skew (of log) (γ)", "ess"], 120.76568406444153),

  # GMM: the same censored record, a different estimator.
  near(f_gmm$parameters[1], 4.0862266825545781),
  near(f_gmm$parameters[2], 0.2031671188484615),
  near(f_gmm$parameters[3], -0.68625874896060901),
  near(f_gmm$j_stat, 3.6036312001472797),
  is.na(f_gmm$j_stat_pval),
  near(quantile_variance(f_gmm, 0.01), 0.0043287640050973901),

  # Diagnostics off the MAP fit.
  near(diag$cooks_distance[2], 0.28415694269045783),
  near(diag$leverage[2], 0.40243088779094255)
)

# Internal consistency: the informative prior moves MAP away from MLE and toward the prior mean,
# every confidence/credible interval brackets its own point estimate, and GMM's log-likelihood
# based fields are structurally NA while its J-statistic is not.
stopifnot(
  abs(f_map$parameters[3] - (-0.3)) < abs(f_mle$parameters[3] - (-0.3)),
  ci["MLE (profile)", "lower"] < f_mle$parameters[3] && f_mle$parameters[3] < ci["MLE (profile)", "upper"],
  ci["MAP (profile)", "lower"] < f_map$parameters[3] && f_map$parameters[3] < ci["MAP (profile)", "upper"],
  is.na(f_gmm$aic), is.na(f_gmm$bic), is.na(logLik(f_gmm)),
  !is.na(f_gmm$j_stat)
)
cat("All reproduction checks passed.\n")
```
